---
title: "abs-to-conc"
vignette: >
  %\VignetteIndexEntry{abs-to-conc}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
library(patchwork)
```

# TODO

- Consider creating a function for the multiple-curve QC (+ explain idea why it is relevant to look at it)

- For conversion: explain well difference btw mg N-sp/L and mg N/L, and how it relates to molar masses

- Consider using ggplot to build the residuals plots, evtl. make it a function (or even make a function of the whole comparison plotting process (4 plots given from the data of one curve)

- in epilogue: **add examples of downstream steps: correct dilutions, per sample average and outlier removal, conversion to ppm, etc.**

- **make a function out of the concentration inference step**

- update relevant functions (qc_raw_abs, maybe plot_st...) to have an additional argument: how is "abs" called. Sometimes abs, sometimes abs_corrected, but with MR i see that it can take many more forms, something with a default like `abs_column = "abs"`. Maybe even have the option of using tidyselect's helper functions, which could be revelant for datasets with numerous layers

::: callout-caution
## Work in progress

This vignette is still under development, bugs are to be expected
:::

# Introduction

In this vignette, we cover the steps from blank-corrected absorbance to concentration in nitrogen \[mg N / L\], although this can easily be used for dosing other molecules[^1], as long as the the relationship between absorbance and concentration is linear or polynomial.

[^1]: To adapt to other compounds, you may need to add molar masses into the data 'molar_masses', see later.

## The Whole Game

This pipeline picks up where `blank-correction` left off (`std_corrected` and `sample_corrected`), decides between a linear or polynomial model based on how your standard curves fit, computes the regression, and infers sample concentrations:

```{mermaid}
flowchart TD
  A(std_corrected + sample_corrected) --> B{Choose model}
  B -->|Linear| C1(Compute linear model)
  C1 --> O1{{QC / outlier removal}}
  O1 --> C1
  C1 --> D1(Infer concentration)
  B -->|Polynomial| C2(Compute polynomial model)
  C2 --> O2{{QC / outlier removal}}
  O2 --> C2
  C2 --> D2(Infer concentration)
  D1 --> E(Epilogue: downstream steps):::sideThing
  D2 --> E
  
  classDef sideThing fill:#f8fafc,stroke:#cbd5e1,color:#64748b,stroke-width:1px; 
   linkStyle 9 stroke:#cbd5e1,stroke-width:1px;
   linkStyle 10 stroke:#cbd5e1,stroke-width:1px;
```

As with `blank-correction`, there's no honest one-shot snippet that skips outlier removal, since checking your standard curves is a manual, visual step. Here's the fast path for the linear model, assuming clean curves (no outlier) with one dilution per plate (no averaging needed):

```{r}
#| eval: false
# compute linear model
lm_table_raw <- lm_std_curve(std_corrected |> dplyr::group_by(plate_id, column))

#** HERE COMES USER-DECISION ON OUTLIER REMOVAL / PER-DILUTION AVERAGING - ITERATIVE PROCESS *

# infer concentration
data_mg_N_L <- reg_join_abs(lm_table_raw, sample_corrected, target_sp = "N") |> 
  dplyr::mutate(conc_mgNsp_L = abs_corrected / slope) |> 
  convert_molec(masses = molar_masses)
```

For the polynomial model, the steps follow the same overall pattern, just with a different model formula — see [Polynomial model](#sec-polynomial) for the specifics.

## The Cheat Sheet

| Step | Function(s) | Purpose |
|------------------------|------------------------|------------------------|
| Compute model | `lm_std_curve()` (add `model = "poly"` for the polynomial model) | Fit a per-curve regression (linear or polynomial) through the origin |
| QC & visualize | `suspicious_lm()`, `plot_list_lm()`, `density_lm_param()` | Identify and visualize curves that don't fit well *(see the upcoming plotting vignette)* |
| Remove outlier wells | `remove_wells()` | Drop wells identified as outliers — explained in detail in `handling-outliers` |
| Per-dilution average | `std_dilution_average()` | Average multiple curves per plate into one |
| Join model to sample data | `reg_join_abs()` | Attach the fitted regression (linear or polynomial) to sample absorbance data |
| Infer concentration | manual formula (linear) or quadratic solve (polynomial) | Convert absorbance to concentration using the fitted model |
| Convert units | `convert_molec()` | Convert between mg *molecule* / L and mg *element* / L using `molar_masses` |

## Overview of abs-to-conc

- Decide which model to use (linear or polynomial) based on how well your standard curves fit a straight line

- Compute the model

  - if suspicious curves are flagged: identify and remove outliers, recompute

  - if there's more than one standard blank per plate (e.g. several curves per plate, or several blank wells within a single curve): compute the per-dilution average, recompute the model

- Join the fitted model to sample data and infer concentration

- Convert to the units relevant to your study (e.g. mg N/L, later ppm — see Epilogue)

::: callout-important
## Prerequisites

- data has been imported and tidied, see `vignette("import-tidy", package = "plate2N")`

- data has been blank-corrected, see `vignette("blank-correction", package = "plate2N")`

- Outliers have been removed throughout that process, see previous vignettes and also `vignette("handling-outliers", package = "plate2N")`
:::

# 1 - Get blank-corrected data

Here is what the blank-corrected data looks like (see prerequisites)

```{r}
sample_corrected

std_corrected
```

# 2 - Tips for choosing the model {#sec-choice}

::: callout-tip
## Choose your model wisely

In this pipeline, we show how to implement a linear model and a polynomial model. The section on linear model is detailed in great depth, then the polynomial model is presented more briefly, with insistence only on points that differ from the linear model.

For dosage of mineral N pools, a polynomial model can be necessary, for example, when concentrations (and absorbance values) are high. The best red flag for the inappropriateness of the linear model is when the fit of the curve does not seem to suit the plotting of the experimental points.

For new pipelines, always check out the standard curves graphically to ensure a good fit.

Curvature at high absorbance can come from more than one source — for example, the color-forming reagent becoming limiting at high concentrations, or stray light, an optical artifact of the plate reader itself that affects any sufficiently high absorbance reading regardless of chemistry. Switching to a polynomial model is a reasonable empirical fix either way, but it's worth understanding roughly why your curve bends before doing so, since the two causes have different implications — for example, whether readings above your calibrated range can be trusted, or whether diluting samples is a better fix than adding model complexity.
:::

```{r}
#| include: false

#*** TO DO: Keep only a few plates (maybe those without outliers?) *

#*** This chunk is silenced, it just serves to construct a data set. Possibly, we can add this into the raw data etc, just don't know yet if it is necessary *

# First, import and tidy data for TDN
TDN_abs <- csv_to_tibble(system.file("extdata", "TDN_data.csv", package = "plate2N"))

TDN_map <- csv_to_tibble(system.file("extdata", "TDN_maps.csv", package = "plate2N"))

TDN_meta <- readr::read_csv(
  system.file("extdata", "TDN_metadata.csv", package = "plate2N"),
  show_col_types = FALSE) |>
  dplyr::mutate(
    date = as.Date(date, tryFormats = "%d/%m/%Y"),
    dataset = "TDN", .before = plate_id) |> 
  dplyr::select(dataset, plate_id, std_sp, std_unit, std_conc, date)


joined_vertical <- join_abs_map(list(TDN_abs, TDN_map), dataset = "TDN-")

tidy_TDN <- vertical_to_tidy(joined_vertical, column_def = c("abs", "map"))

# blank-correct std curves
raw_meta <- tidy_TDN |> 
  dplyr::left_join(TDN_meta, by = dplyr::join_by(dataset, plate_id))

curve_concentration <- extract_curve(TDN_meta, pipetting_direction = "top_down")

std_data <- raw_meta |> 
  extract_std_data(std_def = "Std") |> 
  dplyr::select(!std_conc) |> 
  dplyr::left_join(curve_concentration, by = dplyr::join_by(row, dataset, plate_id))

std_blank <- std_data |> 
  extract_std_blank(
    std_def = "Std",
    pipetting_direction = "top_down")

std_blank_avg <- std_blank_average(std_blank$trusted)

std_corrected_TDN <-
  blank_correct_abs(
    # ungroup std data, remove rows with the blanks (here: row A)
    raw_wells_data = std_data |>
      dplyr::ungroup() |>
      dplyr::filter_out(row == "A"),
    per_plate_avg_blank = std_blank_avg,
    map_to_exclude = ""
  ) |> 
  # only keep relevant columns (remove metadata clutter, optional)
  dplyr::select(row:std_conc)

# blank-correct sample data --> needed?
#extr_data <- extract_extractant(raw_meta)

extr_avg <- extractant_average(raw_meta, extr_def = "extr")

samples_corrected_TDN <-
  blank_correct_abs(
    raw_wells_data = raw_meta,
    per_plate_avg_blank = extr_avg,
    map_to_exclude = c("empty","Std","extr"))
```

We will illustrate the choice of model on another data set called `tidy_TDN` (equivalent in structure to the `tidy_plates` we had at the beginning of the vignette `blank-correction`):

```{r}
tidy_TDN
```

To illustrate the polynomial vs the linear model, we extract standard data for a single curve from that dataset. We chose here the 9th curve because it illustrates the purpose nicely. But to base a model decision at the dataset scale, you would have to repeat this on several curves, e.g., by sampling a random number of curves, then plotting them in a loop.

```{r}
# take data for a single curve and format it for the plotting
curve <- (std_corrected_TDN |> 
            # whatever steps are required to isolate a single curve
            # Here: grouping by plate and column (2 curves per plate)
            dplyr::group_by(plate_id, column) |> 
            # dataset with several N-species for each plate_id --> choose only one
            dplyr::filter(std_sp == "NO3") |> 
           # dplyr::rename(abs = abs_corrected) |> 
            # split the data by group:
            # it creates a list where each group is one element
            dplyr::group_split()
          # take the 9th element of the list
          )[[9]]

# check it out
curve
```

Now we compute both models: linear and polynomial, so that we may compare them.

The linear model takes `abs_corrected ~ 0 + std_conc` as a formula, which is the equivalent to say: `y = mx + p` with y = absorbance and x = concentration. m is the slope, and because we blank-corrected the data already, we are constraining the model to fit through the origin, i.e., p = 0, which justifies the `0 + std_conc` in the model call.

The polynomial model takes `abs_corrected ~ 0 + std_conc + I(std_conc^2)` as a formula, which is the equivalent to say: `y = ax^2 + bx + c`. With the same logic as above, y = absorbance, x = concentration and c = 0 (blank-corrected data).

```{r}
# compute both models
lm_linear <- stats::lm(abs_corrected ~ 0 + std_conc, data = curve)
lm_poly <- stats::lm(abs_corrected ~ 0 + std_conc + I(std_conc^2), data = curve)
```

While both models return p-values corresponding to highly significant models (see next chunk), and very high adjusted R-squared, we will see with the plots that, indeed, the polynomial model is a much better fit in this case

```{r}
(sum_linear <- summary(lm_linear))
(sum_poly <- summary(lm_poly))
```

Seeing the summary of both models: both are significant, but the p-value of the coefficient for the second degree term (a in ax\^2) in the polynomial model is \<\<0.05, which indicates that that term significantly contributes to the model.

Now, we build the 2 plots. In the call to the plot below, we rename abs_corrected with abs, as that is the argument that the function `plot_std()` takes.

```{r}
# build the linear model plot
p_linear <- 
  plot_std(
    curve |> dplyr::rename(abs = abs_corrected), 
    through_origin = TRUE, 
    model = "linear") + 
  ggplot2::theme(legend.position = "none") + 
  ggplot2::labs(title = "Linear model")

# build the polynomial model plot
p_poly <- 
  plot_std(
    curve |> dplyr::rename(abs = abs_corrected), 
    through_origin = TRUE, 
    model = "poly") + 
  ggplot2::theme(legend.position = "none") + 
  ggplot2::labs(title = "Polynomial model")

# looking at both plots next to each other (package "patchwork" needed)
p_linear + p_poly
```

Indeed, the polynomial model fits a lot better

Let's look at the Residual plot to confirm this intuition

```{r}
# Extract residual data from the 2 models
res_linear <- stats::residuals(sum_linear)
res_poly <- stats::residuals(sum_poly)

# plot both on the same graph

# plot residuals for linear model
plot(curve$std_conc, res_linear, main = "Residuals Analysis", xlab = "Concentration", ylab = "Residuals", col = "grey30", pch = 16)
# add red line at y = 0
graphics::abline(h = 0, col = "red", lty = 2)
# add residuals for polynomial model
graphics::points(curve$std_conc, res_poly,  col = "magenta", pch = 15)
# add legend
graphics::points(0, y = -0.1, col = "grey30", pch = 16)
graphics::text(x = 6, y = -0.1, labels = "linear\nmodel", col = "grey30", adj = 0)
graphics::points(0, y = -0.15, col = "magenta", pch = 15)
graphics::text(x = 6, y = -0.15, labels = "polynomial\nmodel", col = "magenta", adj = 0)
```

Whereas this quick approach suffices to convince us that a polynomial model is a better fit for this particular curve, one should evaluate several curves before taking a decision that concerns a larger data set.

# 3 - Linear model from Standard Curves {#sec-linear}

`lm_std_curve()` computes a per-curve linear regression between 2 columns of the input data with `lm(abs_corrected ~ 0 + std_conc, data = curve)` which is the linear model that constraints the curve to go through the origin. `lm_std_curve()` returns a table containing one row per standard curve, and a series of information characterizing the performance of the linear model for that curve (see below and also `?lm_std_curve()`.

## 3.1 - Compute linear model, round 1

`lm_std_curve()` defines a curve based on the groups (defined by `dplyr::group_by()`) of the data given as input (`std_corrected` in the example below). Additionally to the columns `abs_corrected` and `std_conc` (numeric), the function also requires the column `unique_curve_id`.

```{r}
# compute the model and store model details
(lm_table_raw <- lm_std_curve(
  std_corrected |> dplyr::group_by(plate_id, column)))

# check out column names, type, and data
str(lm_table_raw)
```

## 3.2 - QC Standard curves - check conditions of linear model

The function `suspicious_lm()` extracts from an `lm_table` as produced above all plates where the linear model is not optimal, i.e., either the p-value of the model is above 0.05, or its residuals are not normally distributed, or there is heteroscedasticity of residuals. `lm_table_suspicious` can serve for the identification of outlier wells.

In this simplified data set, there is only one suspicious curve:

```{r}
# extract all plates where "something" is not perfect 
(lm_table_suspicious <- lm_table_raw |> suspicious_lm())
```

For visual aid (useful for larger data sets), `plot_list_lm()` creates a [list]{.underline} of plots of each curve given as argument. This means that it can be used to store plots any subset of curves, or for the whole dataset. Calling individual plots from this list can help spotting possible outlier wells, which can be removed with similar steps as shown above.

Each element of the list is one plot, so that calling `suspicious_lm_plotlist[[i]]` will return the i-th plot. The plots are named in the list, so that you can also access any plot by calling the `unique_curve_id`, e.g., with `suspicious_lm_plotlist$NO3_1F1_col12`

```{r}
# here, we create a list of plots from only the suspicious curve
suspicious_lm_plotlist <- plot_list_lm(
  lm_data = lm_table_suspicious,
  std_data = std_corrected)

# check out the names of elements of the list (here only 1)
names(suspicious_lm_plotlist)

# check one plot out
suspicious_lm_plotlist[[1]]

# or by name
suspicious_lm_plotlist$NO3_1F1_col12
```

When there are numerous (suspicious) curves, we can take advantage of the package `patchwork` to display multiple plots (example hereunder with the whole lm_table)

```{r}
full_plotlist <- plot_list_lm(
  lm_data = lm_table_raw,
  std_data = std_corrected)


patchwork::wrap_plots(full_plotlist, axis_titles = "collect_y") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
```

## 3.3 - outlier removal

At this point, you may want to remove obvious outlier wells. Follow steps as shown in vignettes from prerequisites to `remove_wells()`.

Let's say that we want to remove well E12 from plate NO3_1F1 in dataset Nmin:

```{r}
to_remove <- tibble::tibble(
  dataset = "Nmin",
  plate_id = "NO3_1F1",
  well_id = "E12"
)

std_corrected_wash1 <- std_corrected |> remove_wells(to_remove)
```

From now on, we no longer use `std_corrected`, but only `std_corrected_wash1`. We recommend always running one more round of quality check on the cleaned datasets before approving regression equations.

## 3.4 - Per-dilution averages (if 2+ curves per plate)

Once the outlier wells have been removed from single curves, in the case where several curves were pipetted per 96-well plate, we still need to perform a per-dilution average of absorbance. Indeed, there likely have been 2 events of pipetting of the same dilution, rather than 2 independent dilutions.

::: callout-warning
## WARNING

The next step computes per plate per row means for the standard curves.

If some wells have been swapped in some plates, this may cause problems. Make sure there was no pipetting issue, or correct raw data or solve it through code
:::

`std_dilution_average()` does that and creates an artificial "column 13".

Notice that the average reduced the number of rows \~ 2-fold, and notice unique curve ids ending with `_col13`.

```{r}
std_dilution_avg <- std_corrected_wash1 |> std_dilution_average()

# Check it out (notice )
std_dilution_avg
```

## 3.5 - Compute linear model + QC - round 2

We can now rerun the linear model on the cleaned and (if required) per-dilution averaged curve, by repeating the same steps as above: computation of linear model, identification of suspicious curves and plotting

```{r}
# Generate linear model data
(lm_std_mean <- lm_std_curve(
  std_dilution_avg |> dplyr::rename(abs_corrected = abs_mean)))

# look for suspicious curves
(lm_suspicious_mean <- lm_std_mean |> suspicious_lm())
```

Good news, there are no more suspicious linear models anymore. Should there be any, one more round of QC as described above can still be helpful in some cases (`plot_list_lm()`). However, one should not expect to always reach normality of residuals with only 7 points. If the curve fits and the p-value of the model if very low, non-normality and heteroscedasticity can be accepted. They are used as tools to "flag" suspicious curves, but should not be used as too strict criteria.

Let's store the last correction into a clean variable name to reduce possible confusion, and let's compute all the plots in a big list, for storage purposes. We could then export this as one output data in a single list with `readr::write_rds()`, or save the plots with `ggsave()`.

```{r}
# save the last wash in a clean variable name
std_data_clean <- std_dilution_avg
lm_table_clean <- lm_std_mean

# run model on the last, clean version
lm_plots_clean <- plot_list_lm(
  lm_table_clean, std_data_clean |> dplyr::rename(abs_corrected = abs_mean))

# store output and clean data in a list for export (optional)
lm_output <- list(
  "std_data_clean" = std_data_clean,
  "lm_table_clean" = lm_table_clean,
  "lm_plots_clean" = lm_plots_clean,
  "sample_corrected" = sample_corrected
)

# just an example of how to save this in a file for downstream steps
#lm_output |> write_rds("output/data/lm_output.rds")
```

## 3.6 - Multiple curve QC

The function `density_lm_param()` plots the density curve of either p-values or adjusted R\^2 of the model (parameter `p_or_r`). See also `?density_lm_param()` for details and examples.

First, let's look at the distribution of p-values of the std curve regressions

```{r}
density_lm_param(
  lm_table_clean, 
  p_or_r = "p", threshold = 0.05, 
  facetting_std_sp = FALSE, color_std_sp = FALSE) +
  # add caption with number of curves
  ggplot2::labs(caption = paste0("n = ", nrow(lm_table_clean)))
```

Then, same with R_squared (or adjusted?)

```{r}
density_lm_param(
  lm_table_clean, 
  p_or_r = "adjR2", threshold = 0.95, 
  facetting_std_sp = FALSE, color_std_sp = TRUE) +
  # add caption with number of curves
  ggplot2::labs(caption = paste0("n = ", nrow(lm_table_clean))) 
```

::: callout-note
## A note on R² thresholds for regression through the origin

Both the linear and polynomial models throughout this pipeline are constrained to pass through zero (`abs_corrected ~ 0 + std_conc`, since blank-corrected absorbance should read zero at zero concentration). R² for this kind of model is calculated differently than for a standard intercept model — relative to zero rather than the mean of the response — which tends to produce higher values than an equivalent intercept-model R² would. The 0.95 threshold used here (and again in Section 4.4 for the polynomial model) should be read as a practical guideline calibrated for this specific class of model, not as directly comparable to R² thresholds you might see quoted elsewhere for ordinary regression.
:::

Now we plot all curves on same plot. Here we only have 5 curves. But it can be useful to visualize the multi-curve plot to spot plates where something possibly went wrong.

```{r}
#colors <- c("#7FC97F", "#BEAED4", "#FDC086")

lm_output$std_data_clean |> 
  ggplot2::ggplot(ggplot2::aes(
    x = as.numeric(std_conc), y = abs_mean, 
    groups = plate_id, colour = dataset, fill = dataset)) +
  ggplot2::theme_minimal() +
  ggplot2::geom_smooth(
    formula = y ~ 0 + x, method = "lm", se = TRUE, 
    alpha = 0.1, linewidth = 0.5) +
  ggplot2::geom_point() 
```

Now, finally, we decide that we are happy with our standard curves, so we can move on to apply the equations on the data

## 3.7 - Infer sample concentration from regression equation

Check that we are now left with only one curve per plate (i.e., we indeed took a per-dilution average)

```{r}
if (
  (lm_output$std_data_clean |> dplyr::group_by(plate_id) |>  dplyr::n_groups()) == 
  (lm_output$std_data_clean |> dplyr::group_by(unique_curve_id) |>  dplyr::n_groups())
) {message("All good: there is exactly one curve per plate")} else {
  warning("Warning: there is at least one plate with several curves")
}
```

The regression equation is Abs = slope \* Concentration

There is a default vector containing relevant molar masses. Make sure to append it with values that are relevant for your study. This is needed to convert concentrations from mg *molecule* per L to mg *element* per L (e.g., mg NO3/L to mg N/L)

```{r}
molar_masses
```

Here, we

- connect regression data to sample absorbance data with `reg_join_abs()`.

- apply the regression equation to go from absorbance to concentration in mg N-sp per L

- convert unit to mg N per L using `molar_masses` and `convert_molec()`.

```{r}
data_mg_N_L <- 
  # add slope + info regression (p-val and R2) to absorbance data
  reg_join_abs(lm_output$lm_table_clean, lm_output$sample_corrected, target_sp = "N") |> 
  # compute concentration from absorbance
  dplyr::mutate(conc_mgNsp_L = abs_corrected / slope) |> 
  convert_molec(masses = molar_masses)

# Check it out
data_mg_N_L
```

We finally have our computed concentration for each well, expressed in mg N per L. This is not the end of the data pipeline, but it is the end of what really belongs in this vignette. Downstream steps are different for each study. However, we propose below, in [Epilogue](#sec-epilogue), a few examples of further data transformation.

# 4 - Polynomial model {#sec-polynomial}

In this section, we cover the sames steps as in the previous 2 sections, but implementing a polynomial, rather than linear, model. This was necessary, for example, when dosing total dissolved nitrogen, i.e., dosing nitrate after a total oxidation of all N-compounds to nitrate. For this experiment, standard curve concentrations have been increased ten-fold. This resulted in highly concentrated solutions generating absorbance values above 3. It appears that a polynomial model is more appropriate in this case, as can be seen in the next chunks (example of a single curve).

::: callout-tip
## Don't need a polynomial model?

If your standard curves fit a straight line well (see [Tips for choosing the model](#sec-choice)), you can skip this section entirely — jump to [Epilogue](#sec-epilogue) once you've finished the linear model in Section 3.
:::

::: callout-note
## Does a polynomial model change blank correction?

No. Blank correction (see the `blank-correction` vignette) was already carried out using simple subtraction, and that remains valid here even though the model in this section is polynomial. The blank is measured at (or near) zero concentration, before either of the mechanisms mentioned in [Tips for choosing the model](#sec-choice) becomes relevant — so blank subtraction and curve shape are answering two separate questions. Nothing about using a polynomial model here requires revisiting the earlier blank-correction step.
:::

::: callout-warning
## Only polynomial model-specific steps are reviewed in detail

In theory, applying a polynomial model would go through the same logic as the linear model:

- 1st computation of the model + plotting suspicious curves

- optional removal of outliers

- if outliers were removed: 2nd computation of the model + plotting suspicious curves again, etc until we are satisfied with the curves

- In case of several curves per plate: computation of the per-dilution (per-plate-row) average of the standard curve

- If average was computed: 3rd computation of the model on per-dilution averages

- Infering sample concentration

In this section, we mainly focus on steps that differ from the linear model. If it appears too cryptic here, go and check the linear model-equivalent section.

For tips on choice of model, see [Tips for choosing the model](#sec-choice)
:::

The dataset for Total Dissolved Nitrogen data (TDN), and its standard curve data look like this (very similar to what we have seen with other data sets before)

```{r}
std_corrected_TDN

samples_corrected_TDN
```

## 4.1 - Compute polynomial model on Standard curves

These steps follow the same steps as in section [Linear model from Standard Curves](#sec-linear), with some adaptations.

First, we perform a polynomial model on each NO3 curve individually. In the background, the function `lm_std_curve()` performs the model `stats::lm(abs_corrected ~ 0 + std_conc + I(std_conc^2), data = curve)` on each individual curve from the data

```{r}
# Compute the polynomial model
lm_TDN_raw <- lm_std_curve(
  grouped_data = std_corrected_TDN |> dplyr::group_by(unique_curve_id), 
  model = "poly") 

# Check it out
lm_TDN_raw
```

This produces a table that has a similar structure to the one from the linear model, though columns are slightly different to reflect the differences in the model. This table has one row per standard curve. Columns poly_a and poly_b correspond to the coefficients `a` and `b`, respectively, in the equation `y = a*x^2 + b*x + c`, where

- c = 0 because we used blank-corrected absorbance and the model was forced to go through the origin

- y = blank-corrected absorbance

- x = concentration in the compound of interest (here: nitrate)

Other columns in `lm_TDN_raw` contain the p-values associated to `a` and `b` (`poly_a_p` and `poly_b_p`, respectively), the R2 and adjusted R2 values, the p-value associated with the model (`lm_p`), and results of a normality test of residuals (columns `normality_lm_residuals` and `shapiro_p`), and results of a test of homoscedasticity (columns `homoscedasticity_lm_residuals` and `breusch_pagan_p`)

## 4.3 - QC Standard curve

Then we take a subset to examine individually. The function `suspicious_lm()` allows the subsetting of those curves where the linear model doesn't seem to perform ideally. i.e., either non-significant model (p-value \> 0.05), residuals not normally distributed, heteroscedasticity, or p-value of one of the coefficients \> 0.05.

In our case, this concerns 7 curves (from 32)

```{r}
# extract all plates where "something" is not perfect 
(lm_suspicious <- lm_TDN_raw |> suspicious_lm(model = "poly"))
```

We can visually evaluate those plates. For visual support, we create with `plot_list_lm()` a list of plots where we store each individual plot of "suspicious" standard curves.

```{r}
suspicious_plots <- plot_list_lm(
  lm_suspicious, 
  std_data = std_corrected_TDN, 
  model = "poly")

# Look at the first plot
suspicious_plots[[1]]
```

Then we plot them together. Notice that the "problem" of each curve is displayed on the plots. In this case, non-normality of the residuals is always the issue. Should p-values (model or coefficients) be above 0.05 or the residuals be heteroscedastic, this would be displayed as well.

```{r}
patchwork::wrap_plots(suspicious_plots, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
```

Whereas the shapiro test that is run in the background by `lm_std_curve()` to test for normality of residuals is considered a trustworthy test in general, normality is not easily achieved with only 8 data points. This non-normality result should therefore be seen as an aid to spot possibly faulty curves, rather than a strict criterion to exclude them. Visual assessment is therefore crucial.

In this case, all 7 "suspicious" curves look quite good despite not hitting the normality threshold. Should a well be very obviously outside of the curve, then use `remove_wells()` to remove outliers as described many times above and in previous vignettes. Then, re-run the model, etc. until you are satisfied with the curves. Per-dilution averages in case of 2 or more curves per plate can be computed as shown above for the linear model.

::: {#cau-parable .callout-caution}
## Downward facing parable or descending curve?

These functions for the polynomial model have been tested in the case where the data form a downward-facing parable. Knowing this AND the fact that we are in an ascending part of the curve, the smallest solution for x (concentration) is the correct one when solving the equation `a*x^2 + b*x - y = 0` for any value of y (absorbance).

For the combination of upward-facing parable AND ascending curve, however, the highest solution for x would be the right one.

Should you have data where the polynomial fit is correct, but the curves do not look like those displayed here, do check that the solutions for x in the next sections are correct. Please contact us should it not be the case
:::

## 4.4 - Multiple curve QC

First, let's look at the distribution of p-values and adjusted R\^2 of the std curve regressions

```{r}
TDN_p <- density_lm_param(
  lm_TDN_raw, 
  p_or_r = "p", threshold = 0.05, 
  facetting_std_sp = TRUE, color_std_sp = FALSE)

TDN_adjR2 <- density_lm_param(
  lm_TDN_raw, "adjR2", 0.95, color_std_sp = FALSE 
)

TDN_p / TDN_adjR2 + patchwork::plot_annotation(title = "Multiple-curve QC of the\npolynomial model for TDN data")
```

Now we plot all curves on a single plot. Here, we check for batch-effects by assigning the date to the color aesthetics, this should be adapted as needed.

```{r}
(annotation <- std_corrected_TDN |> 
  dplyr::select(plate_id, abs_corrected, std_conc, date) |> 
  dplyr::slice_max(std_conc, with_ties = TRUE))

std_corrected_TDN |> 
  ggplot2::ggplot(ggplot2::aes(
    x = as.numeric(std_conc), 
    y = abs_corrected, groups = plate_id,
    color = date, fill = date)) +
  ggplot2::theme_minimal() + 
  ggplot2::theme(legend.position = "right") +
  ggplot2::geom_smooth(
    formula = y ~ 0 + x + I(x^2), method = "lm", se = TRUE, 
    alpha = 0, linetype = 1, linewidth = 0.15) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::xlab("Standard Concentration [mg N-species / L]") +
  ggplot2::ylab("Blank-corrected absorbance") +
  ggplot2::labs(title = "Inter-plate variability of the Standard Curves") +
  ggplot2::xlim(c(0, 300)) +
  ggplot2::annotate(
    geom = "text",
    x = annotation$std_conc*1.01, 
    y = annotation$abs_corrected, 
    label = annotation$plate_id, size = 2,
    hjust = 0
  ) 
```

## 4.5 - Infer sample concentration from regression equation

Here we use the last version of the polynomial model data (starts with "lm"). In this case, no outlier removal or per-dilution average computation was necessary, therefore we are still working with `lm_TDN_raw`.

First, we join the polynomial-model data (on a per-plate basis) to the sample absorbance data, i.e., each row (= sample-containing well) of the sample data receives additional columns containing polynomial model data. That is what the function `reg_join_abs()` does (it recognizes which model is given as input, so it works either way)

```{r}
data <- lm_TDN_raw |> 
  reg_join_abs(
    samples_corrected_TDN, 
    target_sp = "N")

# Check it out
data
```

Then, we can compute the concentration in N-species in mg N / L (i.e., mg NO3-N / L, mg NH4-N / L, etc.). Because the concentration of the standard curve dilution was expressed in mg N-sp / L (i.e., mg NO3 / L or mg NH4 /L, etc.), a conversion using molar masses of both the target species (N as an element) and the origin species (e.g., NO3-) needs to take place as a last step. For that purpose, `plate2N` stores molar masses of N, NO3-, NO2-, NH4+ in a vector called `molar_masses`.

In the script below, we take the smallest value of the solutions to the 2nd degree equation `a*x^2 + b*x - y = 0`. See @cau-parable for more details on when to adapt this choice, which you can do by changing `min()` to `max()`.

**TO DO: make a function out of the concentration inference step**

```{r}
# Check out molar_masses
molar_masses

data_mg_N_L <- data |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    conc_mgNsp_L = min(c(
      # we take the smallest of the solutions to the 2nd degree equation
        (-poly_b + sqrt(poly_b^2 - 4*poly_a*(-abs_corrected))) / (2*poly_a),
        (-poly_b - sqrt(poly_b^2 - 4*poly_a*(-abs_corrected))) / (2*poly_a)
      )),
    .before = std_sp
  ) |> 
  # remove std_unit to avoid confusion in the new unit
  dplyr::select(!std_unit) |> 
  convert_molec(masses = molar_masses)

# Check it out
data_mg_N_L
```

# 5 - Epilogue {#sec-epilogue}

We now have clean and tidy concentration data, which can be exported for further downstream analysis, for example using `data_mg_N_L |> write_rds("path/to/my/output/TDN_data_mg_N_L")`.

It's worth being explicit about what this concentration represents: `conc_mgN_L` is the concentration **in the well as measured** — one value per well, not yet a final, per-sample or publication-ready number. Depending on your experimental design, one or more further steps are typically still needed, and which ones apply is study-specific. For example, in our own soil dataset:

- **Correcting for dilution.** Any sample that was diluted before reading needs its concentration multiplied back by the dilution factor to reflect the original sample, e.g. `data_mg_N_L |> dplyr::mutate(conc_corrected = conc_mgN_L * dilution_factor)`.
- **Per-sample averaging.** So far we still have 4 data points per sample (in our example data sets, they were pipetted in 4 replicates in the 96-well plates) — these typically need to be averaged, with outliers checked for, to get one value per sample.
- **Unit conversion.** In our case, this means converting to ppm, i.e. mg N / kg dry soil. This requires bringing in external data collected in the lab, not part of the plate-reader pipeline itself: the fresh soil weight used in the extraction, the water content of that fresh soil, and the soil-to-extractant ratio used.

A pipeline for spotting and removing sample-outliers and computing per-sample averages is under development and may arrive later. In the meantime, these downstream steps depend on your specific experimental design and aren't something this vignette can fully generalize.
