---
title: "abs-to-conc"
vignette: >
  %\VignetteIndexEntry{abs-to-conc}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
```

::: callout-caution
## Work in progress

This vignette is still under development, bugs are to be expected
:::

# Introduction

In this vignette, we cover the steps from blank-corrected absorbance to concentration in nitrogen \[mg N / L\], although this can easily be used for dosing other molecules, as long as the Beer-Lambert equation is respected (the relationship between absorbance and concentration is linear)[^1].

[^1]: To adapt to other compounds, you may need to add molar masses into the data \`molar_masses\`, see later.

::: callout-important
## Prerequisites 

- data has been imported and tidied, see `vignette("import-tidy", package = "plate2N")`

- data has been blank-corrected, see `vignette("blank-correction", package = "plate2N")`

- Outliers have been removed throughout that process, see previous vignettes and also `vignette("handling-outliers", package = "plate2N")`
:::

# 1 - Get blank-corrected data

Little reminder how how the blank-corrected data looks like (see prerequisites)

```{r}
sample_corrected

std_corrected
```

# 2 - Compute linear model on Standard Curves

## lm_std_curve(), suspicious_lm() & plot_list_lm()

## reg_join_abs() & convert_molec()

## 2.1 - Compute linear model, round 1

## 2.2 - QC Standard curves - check conditions of linear model

## 2.3 - outlier removal

## 2.4 - Per-dilution averages (if 2+ curves per plate)

## 2.5 - Compute linear model + QC - round 2

## 2.6 - Multiple curve QC

# 3 - Infer sample concentration from regression equation

# 4 - Epilogue
