---
title: "abs-to-conc"
vignette: >
  %\VignetteIndexEntry{abs-to-conc}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
```

# TODO

- Integrate polynomial modelling

- write out last sections: 2.6, 3 and 4

- Consider creating a function for the multiple-curve QC (+ explain idea why it is relevant to look at it)

- For conversion: explain well difference btw mg N-sp/L and mg N/L, and how it relates to molar masses

::: callout-caution
## Work in progress

This vignette is still under development, bugs are to be expected
:::

# Introduction

In this vignette, we cover the steps from blank-corrected absorbance to concentration in nitrogen \[mg N / L\], although this can easily be used for dosing other molecules, as long as the Beer-Lambert equation is respected (the relationship between absorbance and concentration is linear)[^1].

[^1]: To adapt to other compounds, you may need to add molar masses into the data \`molar_masses\`, see later.

::: callout-important
## Prerequisites

- data has been imported and tidied, see `vignette("import-tidy", package = "plate2N")`

- data has been blank-corrected, see `vignette("blank-correction", package = "plate2N")`

- Outliers have been removed throughout that process, see previous vignettes and also `vignette("handling-outliers", package = "plate2N")`
:::

# 1 - Get blank-corrected data

Little reminder how how the blank-corrected data looks like (see prerequisites)

```{r}
sample_corrected

std_corrected
```

# 2 - Compute linear model on Standard Curves

`lm_std_curve()` computes a per-curve linear regression between 2 columns of the input data with `lm(abs_corrected ~ 0 + std_conc, data = curve)` which is the linear model that constraints the curve to go through the origin. `lm_std_curve()` returns a table containing one row per standard curve, and a series of information characterizing the performance of the linear model for that curve (see below and also `?lm_std_curve()`.

## 2.1 - Compute linear model, round 1

`lm_std_curve()` defines the curve based on the groups (`dplyr::group_by()`) of the input data (`std_corrected` in the example below). Additionally to the columns `abs_corrected` and `std_conc` (numeric), the function also requires the column `unique_curve_id`.

```{r}
(lm_table_raw <- lm_std_curve(std_corrected |> dplyr::group_by(plate_id, column)))
```

## 2.2 - QC Standard curves - check conditions of linear model

The function `suspicious_lm()` extracts from an `lm_table`\` as produced above all plates where the linear model is not optimal, i.e., either the p-value of the model is above 0.05, or its residuals are not normally distributed, or there is heteroscedasticity of residuals. lm_table_suspicious can serve for identification of outliers

```{r}
# extract all plates where "something" is not perfect 
(lm_table_suspicious <- lm_table_raw |> suspicious_lm())
```

For visual aid (useful for larger data sets, `plot_list_lm()` creates a list of plots of each curve given as argument (here: suspicious lm's). Calling individual plots can help spotting possible outlier wells, which can be removed with similar steps as shown above.

```{r}
suspicious_lm_plotlist <- plot_list_lm(
  lm_data = lm_table_suspicious,
  std_data = std_corrected)

# check one plot out
suspicious_lm_plotlist[[1]]
```

When there are numerous suspicious curves, we can take advantage of the package `patchwork`\` to display multiple plots (example hereunder with the whole lm_table)

```{r}
full_plotlist <- plot_list_lm(
  lm_data = lm_table_raw,
  std_data = std_corrected)


patchwork::wrap_plots(full_plotlist, axis_titles = "collect_y") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
```

## 2.3 - outlier removal

At this point, you may want to remove obvious outlier wells. Follow steps as shown in vignettes from prerequisites to `remove_wells()`.

Let's say that we want to remove well E12 from plate NO3_1F1 in dataset Nmin:

```{r}
to_remove <- tibble::tibble(
  dataset = "Nmin",
  plate_id = "NO3_1F1",
  well_id = "E12"
)

std_corrected_wash1 <- std_corrected |> remove_wells(to_remove)
```

From now on, we no longer use `std_corrected`, but only `std_corrected_wash1`. We recommend always running one more round of quality check on the cleaned datasets before approving regression equations.

## 2.4 - Per-dilution averages (if 2+ curves per plate)

Once the very few monstrously wrong wells have been removed from single curves, in the case where several curves were pipetted per 96-well plate, we still need to perform a per-dilution average of absorbance. Indeed, there have been 2 events of pipetting of the same dilution, rather than 2 successive dilutions.

::: callout-warning
## WARNING

The next step computes per plate per row means for the standard curves.

If some wells have been swapped in some plates, this may cause problems. Make sure there was no pipetting issue, or correct raw data or solve it through code
:::

`std_dilution_average()` does that and creates an artificial "column 13".

```{r}
std_dilution_avg <- std_corrected_wash1 |> std_dilution_average()
```

## 2.5 - Compute linear model + QC - round 2

We can now rerun the linear model on the cleaned and (if required) per-dilution averaged, by repeating the same steps as above: computation of linear model, identification of suspicious curves and plotting

```{r}
(lm_std_mean <- lm_std_curve(std_dilution_avg |> dplyr::rename(abs_corrected = abs_mean)))
(lm_suspicious_mean <- lm_std_mean |> suspicious_lm())
```

Good news, there are no more suspicious linear models anymore. Should there be any, one more round of QC as described above can still be helpful (`plot_list_lm()`).

Let's store the last correction into a clean variable name to reduce possible confusion, and let's compute all the plots in a big list, for storage purposes. We could then export this as one output data in a single list with `readr::write_rds()`\`.

```{r}
std_data_clean <- std_dilution_avg
lm_table_clean <- lm_std_mean
lm_plots_clean <- plot_list_lm(
  lm_table_clean, std_data_clean |> dplyr::rename(abs_corrected = abs_mean))

lm_output <- list(
  "std_data_clean" = std_data_clean,
  "lm_table_clean" = lm_table_clean,
  "lm_plots_clean" = lm_plots_clean,
  "sample_corrected" = sample_corrected
)

# just an example of how to save this in a file for downstream steps
#lm_output |> write_rds("output/data/lm_output.rds")
```

## 2.6 - Multiple curve QC

TODO –\> make a function? Remove this?

First, let's look at the distribution of p-values of the std curve regressions

```{r}
p_threshold <- 0.05

lm_output$lm_table_clean |> 
  tidyr::separate_wider_delim(
    cols = unique_curve_id, delim = "_", 
    names = c("n_sp", "rest"), too_many = "merge") |>
  ggplot2::ggplot(ggplot2::aes(x = -log(lm_p))) + 
  ggplot2::theme_minimal() +
  # geom_histogram() +
  ggplot2::geom_density() +
  ggplot2::geom_vline(ggplot2::aes(xintercept = -log(p_threshold)), linetype = 2, colour = "purple") +
  ggplot2::annotate(
    geom = "label", label = paste0("p-val = ", p_threshold),
    x = -log(p_threshold), y = 0.15, hjust = 0.25, colour = "purple" ) +
  ggplot2::facet_wrap(~n_sp, nrow = 3) +
  ggplot2::xlim(0,max(-log(lm_output$lm_table_clean$lm_p))) +
  ggplot2::xlab("-log(p-value of linear model)") +
  ggplot2::labs(
    title = "Distribution of p-values of the linear model",
    subtitle = "logarithmic scale"
  )
```

Then, same with R_squared (or adjusted?)

```{r}
threshold <- 95

lm_output$lm_table_clean |> 
  tidyr::separate_wider_delim(
    cols = unique_curve_id, delim = "_", 
    names = c("n_sp", "rest"), too_many = "merge") |>
  ggplot2::ggplot(ggplot2::aes(x = adj_r_squared, colour = n_sp, fill = n_sp)) + 
  ggplot2::theme_minimal() +
  #geom_histogram() +
  ggplot2::geom_density(alpha = 0.3) +
  ggplot2::geom_vline(ggplot2::aes(xintercept = 0.95), linetype = 2, colour = "purple") +
  ggplot2::annotate(
    geom = "label", label = paste0("adjusted R2 = ", threshold, "%"),
    x = threshold/100, y = 500, hjust = 0.25, colour = "purple" ) +
  #facet_wrap(~n_sp, nrow = 3, scales = "free_y") +
  ggplot2::xlim(
    0.945,
    max(lm_output$lm_table_clean$adj_r_squared)) +
  ggplot2::xlab("Adjusted R2") +
  ggplot2::labs(
    title = "Distribution of adjusted R2 values of the linear model"
  )
```

Now we plot all curves on same plot

```{r}
colors <- c("#7FC97F", "#BEAED4", "#FDC086")

lm_output$std_data_clean |> 
  dplyr::filter_out(dataset == "TDN") |> 
  ggplot2::ggplot(ggplot2::aes(x = as.numeric(std_conc), y = abs_mean, groups = plate_id, colour = dataset, fill = dataset)) +
  ggplot2::theme_minimal()+
  ggplot2::geom_smooth(
    formula = y~x-1, method = "lm", se = TRUE, 
    alpha = 0.05) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~std_sp, scales = "free") +
  ggplot2::scale_color_discrete(palette = colors[1:2]) +
  ggplot2::scale_fill_discrete(palette = colors[1:2]) 
```

Now, finally, I decide that I am happy with my standard curves, so I can move on to apply the equations on my data

# 3 - Infer sample concentration from regression equation

Check that we are now left with only one curve per plate (i.e., we indeed took a per-dilution average)

```{r}
if (
  (lm_output$std_data_clean |> dplyr::group_by(plate_id) |>  dplyr::n_groups()) == 
  (lm_output$std_data_clean |> dplyr::group_by(unique_curve_id) |>  dplyr::n_groups())
) {message("All good: there is exactly one curve per plate")} else {
  warning("Warning: there is at least one plate with several curves")
}
```

Regression equation is Abs = slope \* Concentration

There is a default vector containing relevant molar masses. Make sure to append it with values that are relevant for your study. This is needed to convert concentrations from mg *molecule* per L to mg *element* per L (e.g., mg NO3/L to mg N/L)

```{r}
molar_masses
```

Here, we

- connect regression data to sample absorbance data with `reg_join_abs()`.

- apply the regression equation to go from absorbance to concentration in mg N-sp per L

- convert unit to mg N per L using `molar_masses` and `convert_molec()`.

```{r}
data_mg_N_L <- 
  # add slope + info regression (p-val and R2) to absorbance data
  reg_join_abs(lm_output$lm_table_clean, lm_output$sample_corrected, target_sp = "N") |> 
  # compute concentration from absorbance
  dplyr::mutate(conc_mgNsp_L = abs_corrected / slope) |> 
  convert_molec(masses = molar_masses)

data_mg_N_L
```

# 4 - Epilogue

We now have clean and tidy concentration data
