---
title: "abs-to-conc"
vignette: >
  %\VignetteIndexEntry{abs-to-conc}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
library(patchwork)
```

# TODO

- Integrate polynomial modelling, take it from personal pipeline

- write out last sections: 2.6, 3 and 4

- Consider creating a function for the multiple-curve QC (+ explain idea why it is relevant to look at it)

- For conversion: explain well difference btw mg N-sp/L and mg N/L, and how it relates to molar masses

::: callout-caution
## Work in progress

This vignette is still under development, bugs are to be expected
:::

# Introduction

In this vignette, we cover the steps from blank-corrected absorbance to concentration in nitrogen \[mg N / L\], although this can easily be used for dosing other molecules, as long as the Beer-Lambert equation is respected (the relationship between absorbance and concentration is linear)[^1].

[^1]: To adapt to other compounds, you may need to add molar masses into the data \`molar_masses\`, see later.

::: callout-important
## Prerequisites

- data has been imported and tidied, see `vignette("import-tidy", package = "plate2N")`

- data has been blank-corrected, see `vignette("blank-correction", package = "plate2N")`

- Outliers have been removed throughout that process, see previous vignettes and also `vignette("handling-outliers", package = "plate2N")`
:::

# 1 - Get blank-corrected data

Little reminder how how the blank-corrected data looks like (see prerequisites)

```{r}
sample_corrected

std_corrected
```

# 2 - Compute linear model on Standard Curves {#sec-linear}

::: callout-tip
## Choose your model wisely

We cover later a script to implement a polynomial model rather than a linear model. For dosage of mineral N pools, such a model can be necessary, for example, when concentrations (and absorbance values) are high. The best red flag for the inappropriateness of the linear model if when the fit of the curve does not seem to suit the plotting of the experimental points.

For new pipelines, always check out the standard curves graphically to ensure a good fit. See later, in section @sec-polynomial
:::

`lm_std_curve()` computes a per-curve linear regression between 2 columns of the input data with `lm(abs_corrected ~ 0 + std_conc, data = curve)` which is the linear model that constraints the curve to go through the origin. `lm_std_curve()` returns a table containing one row per standard curve, and a series of information characterizing the performance of the linear model for that curve (see below and also `?lm_std_curve()`.

## 2.1 - Compute linear model, round 1

`lm_std_curve()` defines the curve based on the groups (`dplyr::group_by()`) of the input data (`std_corrected` in the example below). Additionally to the columns `abs_corrected` and `std_conc` (numeric), the function also requires the column `unique_curve_id`.

```{r}
(lm_table_raw <- lm_std_curve(std_corrected |> dplyr::group_by(plate_id, column)))
```

## 2.2 - QC Standard curves - check conditions of linear model

The function `suspicious_lm()` extracts from an `lm_table`\` as produced above all plates where the linear model is not optimal, i.e., either the p-value of the model is above 0.05, or its residuals are not normally distributed, or there is heteroscedasticity of residuals. lm_table_suspicious can serve for identification of outliers

```{r}
# extract all plates where "something" is not perfect 
(lm_table_suspicious <- lm_table_raw |> suspicious_lm())
```

For visual aid (useful for larger data sets, `plot_list_lm()` creates a list of plots of each curve given as argument (here: suspicious lm's). Calling individual plots can help spotting possible outlier wells, which can be removed with similar steps as shown above.

```{r}
suspicious_lm_plotlist <- plot_list_lm(
  lm_data = lm_table_suspicious,
  std_data = std_corrected)

# check one plot out
suspicious_lm_plotlist[[1]]
```

When there are numerous suspicious curves, we can take advantage of the package `patchwork`\` to display multiple plots (example hereunder with the whole lm_table)

```{r}
full_plotlist <- plot_list_lm(
  lm_data = lm_table_raw,
  std_data = std_corrected)


patchwork::wrap_plots(full_plotlist, axis_titles = "collect_y") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
```

## 2.3 - outlier removal

At this point, you may want to remove obvious outlier wells. Follow steps as shown in vignettes from prerequisites to `remove_wells()`.

Let's say that we want to remove well E12 from plate NO3_1F1 in dataset Nmin:

```{r}
to_remove <- tibble::tibble(
  dataset = "Nmin",
  plate_id = "NO3_1F1",
  well_id = "E12"
)

std_corrected_wash1 <- std_corrected |> remove_wells(to_remove)
```

From now on, we no longer use `std_corrected`, but only `std_corrected_wash1`. We recommend always running one more round of quality check on the cleaned datasets before approving regression equations.

## 2.4 - Per-dilution averages (if 2+ curves per plate)

Once the very few monstrously wrong wells have been removed from single curves, in the case where several curves were pipetted per 96-well plate, we still need to perform a per-dilution average of absorbance. Indeed, there have been 2 events of pipetting of the same dilution, rather than 2 successive dilutions.

::: callout-warning
## WARNING

The next step computes per plate per row means for the standard curves.

If some wells have been swapped in some plates, this may cause problems. Make sure there was no pipetting issue, or correct raw data or solve it through code
:::

`std_dilution_average()` does that and creates an artificial "column 13".

```{r}
std_dilution_avg <- std_corrected_wash1 |> std_dilution_average()
```

## 2.5 - Compute linear model + QC - round 2

We can now rerun the linear model on the cleaned and (if required) per-dilution averaged, by repeating the same steps as above: computation of linear model, identification of suspicious curves and plotting

```{r}
(lm_std_mean <- lm_std_curve(std_dilution_avg |> dplyr::rename(abs_corrected = abs_mean)))
(lm_suspicious_mean <- lm_std_mean |> suspicious_lm())
```

Good news, there are no more suspicious linear models anymore. Should there be any, one more round of QC as described above can still be helpful (`plot_list_lm()`).

Let's store the last correction into a clean variable name to reduce possible confusion, and let's compute all the plots in a big list, for storage purposes. We could then export this as one output data in a single list with `readr::write_rds()`\`.

```{r}
std_data_clean <- std_dilution_avg
lm_table_clean <- lm_std_mean
lm_plots_clean <- plot_list_lm(
  lm_table_clean, std_data_clean |> dplyr::rename(abs_corrected = abs_mean))

lm_output <- list(
  "std_data_clean" = std_data_clean,
  "lm_table_clean" = lm_table_clean,
  "lm_plots_clean" = lm_plots_clean,
  "sample_corrected" = sample_corrected
)

# just an example of how to save this in a file for downstream steps
#lm_output |> write_rds("output/data/lm_output.rds")
```

## 2.6 - Multiple curve QC

TODO –\> make a function? Remove this?

First, let's look at the distribution of p-values of the std curve regressions

```{r}
p_threshold <- 0.05

lm_output$lm_table_clean |> 
  tidyr::separate_wider_delim(
    cols = unique_curve_id, delim = "_", 
    names = c("n_sp", "rest"), too_many = "merge") |>
  ggplot2::ggplot(ggplot2::aes(x = -log(lm_p))) + 
  ggplot2::theme_minimal() +
  # geom_histogram() +
  ggplot2::geom_density() +
  ggplot2::geom_vline(ggplot2::aes(xintercept = -log(p_threshold)), linetype = 2, colour = "purple") +
  ggplot2::annotate(
    geom = "label", label = paste0("p-val = ", p_threshold),
    x = -log(p_threshold), y = 0.15, hjust = 0.25, colour = "purple" ) +
  ggplot2::facet_wrap(~n_sp, nrow = 3) +
  ggplot2::xlim(0,max(-log(lm_output$lm_table_clean$lm_p))) +
  ggplot2::xlab("-log(p-value of linear model)") +
  ggplot2::labs(
    title = "Distribution of p-values of the linear model",
    subtitle = "logarithmic scale"
  )
```

Then, same with R_squared (or adjusted?)

```{r}
threshold <- 95

lm_output$lm_table_clean |> 
  tidyr::separate_wider_delim(
    cols = unique_curve_id, delim = "_", 
    names = c("n_sp", "rest"), too_many = "merge") |>
  ggplot2::ggplot(ggplot2::aes(x = adj_r_squared, colour = n_sp, fill = n_sp)) + 
  ggplot2::theme_minimal() +
  #geom_histogram() +
  ggplot2::geom_density(alpha = 0.3) +
  ggplot2::geom_vline(ggplot2::aes(xintercept = 0.95), linetype = 2, colour = "purple") +
  ggplot2::annotate(
    geom = "label", label = paste0("adjusted R2 = ", threshold, "%"),
    x = threshold/100, y = 500, hjust = 0.25, colour = "purple" ) +
  #facet_wrap(~n_sp, nrow = 3, scales = "free_y") +
  ggplot2::xlim(
    0.945,
    max(lm_output$lm_table_clean$adj_r_squared)) +
  ggplot2::xlab("Adjusted R2") +
  ggplot2::labs(
    title = "Distribution of adjusted R2 values of the linear model"
  )
```

Now we plot all curves on same plot

```{r}
colors <- c("#7FC97F", "#BEAED4", "#FDC086")

lm_output$std_data_clean |> 
  dplyr::filter_out(dataset == "TDN") |> 
  ggplot2::ggplot(ggplot2::aes(x = as.numeric(std_conc), y = abs_mean, groups = plate_id, colour = dataset, fill = dataset)) +
  ggplot2::theme_minimal()+
  ggplot2::geom_smooth(
    formula = y~x-1, method = "lm", se = TRUE, 
    alpha = 0.05) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~std_sp, scales = "free") +
  ggplot2::scale_color_discrete(palette = colors[1:2]) +
  ggplot2::scale_fill_discrete(palette = colors[1:2]) 
```

Now, finally, I decide that I am happy with my standard curves, so I can move on to apply the equations on my data

# 3 - Infer sample concentration from regression equation

Check that we are now left with only one curve per plate (i.e., we indeed took a per-dilution average)

```{r}
if (
  (lm_output$std_data_clean |> dplyr::group_by(plate_id) |>  dplyr::n_groups()) == 
  (lm_output$std_data_clean |> dplyr::group_by(unique_curve_id) |>  dplyr::n_groups())
) {message("All good: there is exactly one curve per plate")} else {
  warning("Warning: there is at least one plate with several curves")
}
```

Regression equation is Abs = slope \* Concentration

There is a default vector containing relevant molar masses. Make sure to append it with values that are relevant for your study. This is needed to convert concentrations from mg *molecule* per L to mg *element* per L (e.g., mg NO3/L to mg N/L)

```{r}
molar_masses
```

Here, we

- connect regression data to sample absorbance data with `reg_join_abs()`.

- apply the regression equation to go from absorbance to concentration in mg N-sp per L

- convert unit to mg N per L using `molar_masses` and `convert_molec()`.

```{r}
data_mg_N_L <- 
  # add slope + info regression (p-val and R2) to absorbance data
  reg_join_abs(lm_output$lm_table_clean, lm_output$sample_corrected, target_sp = "N") |> 
  # compute concentration from absorbance
  dplyr::mutate(conc_mgNsp_L = abs_corrected / slope) |> 
  convert_molec(masses = molar_masses)

data_mg_N_L
```

# 4 - Polynomial model {#sec-polynomial}

In this section, we cover the sames steps as in the previous 2 sections, but implementing a polynomial, rather than linear, model. This was necessary, for example, when dosing total dissolved nitrogen, i.e., dosing nitrate after a total oxidation of all N-compounds to nitrate. For this experiment, standard curve concentrations have been increased ten-fold. This resulted in highly concentrated solutions generating absorbance values above 3. It appears that a polynomial model is more appropriated in this case, as can be seen in the next chunks (example of a single curve)

::: callout-warning
## Only polynomial model - specific steps are reviewed in detail

In theory, applying a polynomial model would go through the same logic as the linear model:

- 1st computation of the model + plotting suspicious curves

- optional removal of outliers

- if outliers were removed: 2nd computation of the model + plotting suspicious curves again, etc untill we are satisfied with the curves

- In case of several curves per plate: computation of the per-dilution (per-plate-row) average of the standard curve

- If average was computed: 3rd computation of the model on per-dilution averages

- Infering sample concentration

In this section, we mainly focus on steps that differ from the linear model
:::

[**TO DO: Keep only a few plates (maybe those without outliers?)**]{.underline}

```{r}
#| include: false

#*** This chunk is silenced, it just serves to construct a data set. Possibly, we can add this into the raw data etc, just don't know yet if it is necessary *

# First, import and tidy data for TDN
TDN_abs <- csv_to_tibble(system.file("extdata", "TDN_data.csv", package = "plate2N"))

TDN_map <- csv_to_tibble(system.file("extdata", "TDN_maps.csv", package = "plate2N"))

TDN_meta <- readr::read_csv(
  system.file("extdata", "TDN_metadata.csv", package = "plate2N"),
  show_col_types = FALSE) |>
  dplyr::mutate(
    date = as.Date(date, tryFormats = "%d/%m/%Y"),
    dataset = "TDN", .before = plate_id) |> 
  dplyr::select(dataset, plate_id, std_sp, std_unit, std_conc, date)


joined_vertical <- join_abs_map(TDN_abs, TDN_map, dataset = "TDN-")

tidy_TDN <- vertical_to_tidy(joined_vertical, abs_def = "abs", map_def = "map")

# blank-correct std curves
raw_meta <- tidy_TDN |> 
  dplyr::left_join(TDN_meta, by = dplyr::join_by(dataset, plate_id))

curve_concentration <- extract_curve(TDN_meta, pipetting_direction = "top_down")

std_data <- raw_meta |> 
  extract_std_data(std_def = "Std") |> 
  dplyr::select(!std_conc) |> 
  dplyr::left_join(curve_concentration, by = dplyr::join_by(row, dataset, plate_id))

std_blank <- std_data |> 
  extract_std_blank(
    std_def = "Std",
    pipetting_direction = "top_down")

std_blank_avg <- std_blank_average(std_blank$trusted)

std_corrected_TDN <-
  blank_correct_abs(
    # ungroup std data, remove rows with the blanks (here: row A)
    raw_wells_data = std_data |>
      dplyr::ungroup() |>
      dplyr::filter_out(row == "A"),
    per_plate_avg_blank = std_blank_avg,
    map_to_exclude = ""
  ) |> 
  # only keep relevant columns (remove metadata clutter, optional)
  dplyr::select(row:std_conc)

# blank-correct sample data --> needed?
#extr_data <- extract_extractant(raw_meta)

extr_avg <- extractant_average(raw_meta, extr_def = "extr")

samples_corrected_TDN <-
  blank_correct_abs(
    raw_wells_data = raw_meta,
    per_plate_avg_blank = extr_avg,
    map_to_exclude = c("empty","Std","extr"))
```

## 4.1 - Choice of the best fitting model

The dataset for Total Dissolved Nitrogen data (TDN), and its standard curve data look like this (very similar to what we have seen with other data sets before)

```{r}
tidy_TDN

std_corrected_TDN

samples_corrected_TDN
```

To illustrate the polynomial vs the linear model, we extract standard data for a single curve

```{r}
# take data for a single curve and format it for the plotting
curve <- (std_corrected_TDN |> 
            dplyr::group_by(plate_id, column) |> 
            dplyr::filter(std_sp == "NO3") |> 
            dplyr::rename(abs = abs_corrected) |> 
            dplyr::group_split()
          )[[9]]

# check it out
curve
```

Now we compute both models: linear and polynomial, so that we may compare them

```{r}
# compute both models
lm_linear <- stats::lm(abs ~ 0 + std_conc, data = curve)
lm_poly <- stats::lm(abs ~ 0 + std_conc + I(std_conc^2), data = curve)
```

While both models return p-values corresponding to highly significant models (see next chunk), and very high adjusted R-squared, we will see with the plots that, indeed, the polynomial model is a much better fit

```{r}
(sum_linear <- summary(lm_linear))
(sum_poly <- summary(lm_poly))
```

Seeing the summary of both models: both are significant, but the p-value of the coefficient for the second degree term (b in bx\^2) in the polynomial model is \<0.05, which indicates that that term significantly contributes to the model. This becomes also very obvious when we look at the plots (the polynomial model fits a lot better)

```{r}
p_linear <- plot_std(curve, through_origin = TRUE, model = "linear") + 
  ggplot2::theme(legend.position = "none") + 
  ggplot2::labs(title = "Linear model")
p_poly <- plot_std(curve, through_origin = TRUE, model = "poly") + 
  ggplot2::theme(legend.position = "none") + 
  ggplot2::labs(title = "Polynomial model")

p_linear + p_poly
```

Let's look at the Residual plot to confirm this intuition

```{r}
# Extract residual data from the 2 models
res_linear <- stats::residuals(sum_linear)
res_poly <- stats::residuals(sum_poly)

# plot both on the same graph

# plot residuals for linear model
plot(curve$std_conc, res_linear, main = "Residuals Analysis", xlab = "Concentration", ylab = "Residuals", col = "grey30", pch = 16)
# add red line at y = 0
graphics::abline(h = 0, col = "red", lty = 2)
# add residuals for polynomial model
graphics::points(curve$std_conc, res_poly,  col = "magenta", pch = 15)
# add legend
graphics::points(0, y = -0.1, col = "grey30", pch = 16)
graphics::text(x = 6, y = -0.1, labels = "linear\nmodel", col = "grey30", adj = 0)
graphics::points(0, y = -0.15, col = "magenta", pch = 15)
graphics::text(x = 6, y = -0.15, labels = "polynomial\nmodel", col = "magenta", adj = 0)
```

Whereas this quick approach suffices to convince us that a polynomial model is a better fit for this particular curve, one should evaluate several curves before taking a decision that concerns a larger data set

## 4.2 - Compute polynomial model on Standard curves

These steps follow the same steps as in section @sec-linear , with some adaptations.

First, we perform a polynomial model on each NO3 curve individually. In the background, the function `lm_std_curve()` performs the model `stats::lm(abs_corrected ~ 0 + std_conc + I(std_conc^2), data = curve)` on each individual curve from the data

```{r}
# Compute the polynomial model
lm_TDN_raw <- lm_std_curve(
  grouped_data = std_corrected_TDN |> dplyr::group_by(unique_curve_id), 
  model = "poly") 

```

This produces a table that has a similar structure to the one from the linear model, though columns are slightly different to reflect the differences in the model. This table has one row per standard curve. Columns poly_a and poly_b correspond to the coefficients `a` and `b`, respectively, in the equation `y = a*x^2 + b*x + c`, where

- c = 0 because we used blank-corrected absorbance and the model was forced to go through the origin

- y = blank-corrected absorbance

- x = concentration in the compound of interest (here: nitrate)

## 4.3 - QC Standard curve

Other columns in `lm_TDN_raw` contain the p-values associated to `a` and `b` (poly_a_p and poly_b_p, respectively), the R2 and adjusted R2 values, the p-value associated with the model, and results of a normality test of residuals (columns `normality_lm_residuals` and `shapiro_p`), and results of a test of homoscedasticity (columns `homoscedasticity_lm_residuals` and `breusch_pagan_p`)

```{r}
# Check it out
lm_TDN_raw
```

Then we take a subset to examine individually. The function `suspicious_lm()` allows the subsetting of those curves where the linear model doesn't seem to perform ideally. i.e., either non-significant model (p-value \> 0.05), residuals not normally distributed, heteroscedasticity, or p-value of one of the coefficients \> 0.05.

In this case, this concerns 7 curves (from 32)

```{r}
# extract all plates where "something" is not perfect 
(lm_suspicious <- lm_TDN_raw |> suspicious_lm(model = "poly"))
```

We can visually evaluate those plates. For visual support, we create with `plot_list_lm()` a list of plots where we store each individual plot of "suspicious" standard curves.

```{r}
suspicious_plots <- plot_list_lm(
  lm_suspicious, 
  std_data = std_corrected_TDN, 
  model = "poly")
```

Then we plot them together. Notice that the "default" of each curve is displayed on the plots. In this case, non-normality of the residuals is always the issue. Should p-values (model or coefficients) be above 0.05 or the residuals be heteroscedastic, this would be displayed as well.

```{r}
patchwork::wrap_plots(suspicious_plots, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
```

Whereas the shapiro test that is ran in the background by `lm_std_curve()` to test for normality of residuals is considered a trustworthy test, normality is not easily achieved with only 8 data points. This non-normality result should therefore be seen as an aid to spot possibly faulty curves, rather than a strict criterion to exclude curves. Visual assessment is therefore crucial.

In this case, all 7 "suspicious" curves look quite good despite not hitting the normality threshold. Should a well be very obviously outside of the curve, then use `remove_wells()` to remove outliers as described many times above, the re-run the model, etc. until you are satisfied with the curves. Per-dilution averages in case of 2 or more curves per plate can be computed as shown above for the linear model.

::: {#cau-parable .callout-caution}
## Downward facing parable?

These functions for the polynomial model have been tested in the case where the data form a downward-facing parable, i.e., when the curve is fitted to go through the origin, the smallest value for x (concentration) is the correct one when solving the equation `a*x^2 + b*x - y = 0` for any value of y (absorbance). For an upward-facing parable, however, the highest solution for x would be the right one (provided the relationship between absorbance and concentration is an ascending curve).

Should you have data where the polynomial fit is correct, but the curves do not look like those displayed here, do check that the solutions for x in the next sections are correct. Please contact us should it not be the case
:::

## 4.4 - Multiple curve QC

First, let's look at the distribution of p-values and adjusted R\^2 of the std curve regressions

```{r}
TDN_p <- density_lm_param(
  lm_TDN_raw, 
  p_or_r = "p", threshold = 0.05, 
  facetting_std_sp = TRUE, color_std_sp = FALSE)

TDN_adjR2 <- density_lm_param(
  lm_TDN_raw, "adjR2", 0.95, color_std_sp = FALSE 
)

TDN_p / TDN_adjR2 + patchwork::plot_annotation(title = "Multiple-curve QC of the polynomial model for TDN data")

```

Now we plot all curves on same plot. Here, we check for batch-effects by assigning the date to the color aesthetics, this should be adapted as needed.

```{r}
annotation <- std_corrected_TDN |> 
  dplyr::select(plate_id, abs_corrected, std_conc, date) |> 
  dplyr::slice_max(std_conc, with_ties = TRUE) 

std_corrected_TDN |> 
  ggplot2::ggplot(ggplot2::aes(
    x = as.numeric(std_conc), 
    y = abs_corrected, groups = plate_id,
    color = date, fill = date)) +
  ggplot2::theme_minimal() + 
  ggplot2::theme(legend.position = "right") +
  ggplot2::geom_smooth(
    formula = y ~ 0 + x + I(x^2), method = "lm", se = TRUE, 
    alpha = 0, linetype = 1, linewidth = 0.15) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::xlab("Standard Concentration [mg N-species / L]") +
  ggplot2::ylab("Blank-corrected absorbance") +
  ggplot2::labs(title = "Inter-plate variability of the Standard Curves") +
  ggplot2::xlim(c(0, 300)) +
  ggplot2::annotate(
    geom = "text",
    x = annotation$std_conc*1.01, 
    y = annotation$abs_corrected, 
    label = annotation$plate_id, size = 2,
    hjust = 0
  ) 
```

## 4.5 - Infer sample concentration from regression equation

Here we use the last version of the linear model data (starts with "lm"). In this case, no outlier removal or per-dilution average computation was necessary, therefore we are still working with `lm_TDN_raw`.

First, we join the liner-model data (on a per-plate basis) to the sample data, i.e., each row (= sample-containing well) of the sample data receives additional columns containing polynomial model data. That is what the function `reg_join_abs()` does (it works on linear or polynomial model)

```{r}
data <- lm_TDN_raw |> 
  reg_join_abs(
    samples_corrected_TDN, 
    target_sp = "N")

# Check it out
data
```

Then, we can compute the concentration in N-species in mg N / L (i.e., mg NO3-N / L, mg NH4-N / L, etc.). Because the concentration of the standard curve dilution was expressed in mg N-sp / L (i.e., mg NO3 / L or mg NH4 /L, etc.), a conversion using molar masses of both the target species (N as an element) and the origin species (e.g., NO3-) needs to take place as a last step. For that purposes, `plate2N` stores molar masses of N, NO3-, NO2-, NH4+ in a vector called `molar_masses`.

In the script below, we take the smallest value of the solutions to the 2nd degree equation `a*x^2 + b*x - y = 0`. See @cau-parable for more details on when to adapt this choice (change `min()` to `max()`).

```{r}
# Check out molar_masses
molar_masses

data_mg_N_L <- data |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    conc_mgNsp_L = min(c(
      # we take the smallest of the solutions to the 2nd degree equation
        (-poly_b + sqrt(poly_b^2 - 4*poly_a*(-abs_corrected))) / (2*poly_a),
        (-poly_b - sqrt(poly_b^2 - 4*poly_a*(-abs_corrected))) / (2*poly_a)
      )),
    .before = std_sp
  ) |> 
  # remove std_unit to avoid confusion in the new unit
  dplyr::select(!std_unit) |> 
  convert_molec(masses = molar_masses)

# Check it out
data_mg_N_L
```

# 5 - Epilogue

We now have clean and tidy concentration data, which can be exported for further downstream analysis, for example using `data_mg_N_L |> write_rds("path/to/my/output/TDN_data_mg_N_L")`.

Notice that so far we still have 4 data points per sample (in our example data sets, they were pipetted in 4 replicates in the 96-well plates). A pipeline for spotting and removing sample-outliers and computing per-sample average is under development and may arrive later.
