---
title: "blank-correction"
vignette: >
  %\VignetteIndexEntry{blank-correction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
```

::: callout-note
## Actively evolving

This pipeline works for its current use cases, but it is still gaining flexibility — some behavior that is hardcoded today may later become a parameter, and new features get added as real use cases come up (e.g. support for multiple extractants per plate was added in response to a user's needs, and touched a fair amount of the underlying functions). So don't be surprised if some details shift between versions. If anything here feels more cryptic than the `import-tidy` vignette, that's fair — this one hasn't had as many passes yet, and it'll keep getting clearer.
:::

# TO DO

- Consider making a function of creating the "to_remove" table for extractant outliers (section 3.3.3, \[the multiple-extractants section\](#sec-multiple-extr))
- In the function qc_raw_qbs(), separate condition for export plot and show plot. If we export it, we may not want to plot it as an output...
- allow other digit and separator delimitors in extract_curve –\> make it a parameter
- the last paragraph of 3.3.3 is confusing –\> change the raw data so that we can remove this bit and re-run the coefficient of variation as we should

# Introduction

This vignette shows the basic functions that allow the correction of raw absorbance data (`abs`) to blank-corrected absorbance data, referred to as `abs_corrected` throughout this package.

This pipeline is adapted to the case where the standard curve solutions were prepared with a different blank than the samples. Because of that, there are 2 parallel pipelines to perform the blank-correction (first on standard curves data, then on samples data), then data from standard curves and from samples are merged again.

In theory, the "samples" pipeline should be adaptable to the case where all wells used the same blank (including those containing standard curve solutions), though this has yet to be tested, and bugs may occur. Feel free to reach out to suggest improvements or signal bugs.

To avoid confusion between "blank of the standard curve" and "blank of the samples", the sample-blank is referred to as "extractant" or `extr` throughout this vignette, to refer to the solution that was used to extract (N-)compounds from soil samples and now serves as blank. The standard blank is referred to as `std_blank`.

|   | Standard curve blank | Sample blank |
|----|----|----|
| Name used | `std_blank` | `extr` (extractant) |
| What it physically is | The zero-concentration point of the standard curve itself | The blank solution matching the sample matrix[^1] |
| What it corrects | Standard curve absorbance readings | Sample absorbance readings |

[^1]: In our experiments, this is the solution used to extract N-compounds from soil samples. More generally, this pipeline works for any dosage following a Beer-Lambert-style relationship between absorbance and concentration — `extr` can be whatever solution serves as the blank for your own sample matrix.

## The Whole Game

This pipeline runs as two parallel branches — one for the standard curve, one for samples — that merge again once both are blank-corrected:

```{mermaid}
flowchart TD
  A[Raw absorbance + metadata]
  A --> B1[Extract std curve data]
  B1 --> C1[Compute blank average]
  C1 --> O1{{Outlier removal}}
  O1 --> C1
  C1 --> D1[Blank-correct std curve]
  D1 --> E1[std_corrected]

  A --> B2[Extract extractant data]
  B2 --> C2[Compute blank average]
  C2 --> O2{{Outlier removal}}
  O2 --> C2
  C2 --> D2[Blank-correct samples]
  D2 --> E2[sample_corrected]

  E1 --> F[abs-to-conc]
  E2 --> F
```

There's no honest one-shot code snippet for the whole thing, because deciding which wells are outliers is a manual, visual step — not something that can be scripted. But if your data happens to be clean (no outliers to remove), here's the fast path:

```{r}
#| eval: false
# join raw data with per-plate metadata
raw_meta <- tidy_plates |> 
  dplyr::left_join(meta, by = dplyr::join_by(dataset, plate_id))

###############################
# --- standard curve branch ---
###############################

# extract serial dilutions of the standard curve
curve_concentration <- extract_curve(meta, pipetting_direction = "top_down")

# extract standard data and connect it to curve concentration
std_data <- raw_meta |> 
  # extract data corresponding to the standard curve only
  extract_std_data(std_def = "Std") |> 
  # remove the 1-liner standard curve concentration
  dplyr::select(!std_conc) |> 
  # create correspondence with curve concentration
  dplyr::left_join(curve_concentration, by = dplyr::join_by(row, dataset, plate_id))

# extract the blank from standard data
std_blank <- std_data |> extract_std_blank(std_def = "Std", pipetting_direction = "top_down")

# compute per-curve blank average from trusted wells
std_blank_avg <- std_blank_average(std_blank$trusted)

#** HERE COMES USER-DECISION ON OUTLIER REMOVAL - ITERATIVE PROCESS *

# blank-correct standard curve data
std_corrected <- blank_correct_abs(
  raw_wells_data = std_data |> dplyr::ungroup() |> dplyr::filter_out(row == "A"),
  per_plate_avg_blank = std_blank_avg,
  map_to_exclude = ""
) |> dplyr::select(row:abs_corrected)

#######################
# --- sample branch ---
#######################

# compute per-plate average for extractant data
extr_avg <- extractant_average(raw_meta, extr_def = "extr")

#** HERE COMES USER-DECISION ON OUTLIER REMOVAL - ITERATIVE PROCESS *

# blank-correct sample data 
sample_corrected <- blank_correct_abs(
  raw_wells_data = raw_meta,
  per_plate_avg_blank = extr_avg,
   map_to_exclude = c("empty", "Std", "extr"))
```

In practice, you'll almost always want the quality-check and outlier-removal steps covered in detail below — this shortcut just shows the shape of the pipeline before we get into the specifics.

## The Cheat Sheet

+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Step                                 | Function(s)                                                                                | Purpose                                                                                                                     |
+======================================+============================================================================================+=============================================================================================================================+
| Join data                            | `dplyr::left_join()`                                                                       | Join two data frames together based on shared column values (e.g., matching per-plate metadata to raw absorbance data)      |
+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Extract curve concentrations         | `extract_curve()`                                                                          | Expand compact per-plate curve concentrations into one row per well                                                         |
+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Extract std curve & blank            | `extract_std_data()`, `extract_std_blank()`                                                | Subset standard curve wells and identify their blank wells                                                                  |
+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Average std blank                    | `std_blank_average()`                                                                      | Compute the per-plate average of the standard blank                                                                         |
+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Extract & average extractant         | `extractant_average()`                                                                     | Compute the per-plate average of the sample blank (extractant)                                                              |
+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| QC & visualize outliers              | `qc_raw_extr()`, `plot_blank_var_distrib()`, `boxplot_outlier_extr()`, `suspicious_extr()` | Identify and visualize suspicious/outlier wells                                                                             |
|                                      |                                                                                            |                                                                                                                             |
|                                      |                                                                                            | *(see the upcoming plotting vignette)*                                                                                      |
+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Remove outlier wells                 | `remove_wells()`                                                                           | Drop wells identified as outliers — explained in detail in `handling-outliers`                                              |
+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Blank-correct (std curve or samples) | `blank_correct_abs()`                                                                      | Subtract the relevant blank average from absorbance — same function for both branches, just called with different arguments |
+--------------------------------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+

## Overview of blank-correction

- Blank correction of standard curve

  - extracting standard curve data and their blank

  - if there is more than one blank well to compare (e.g. several curves per plate, or several blank wells within a single curve):

    - identifying and remove outliers

    - compute the per-plate mean of blanks without outliers

  - correct the rest of the standard curve data by subtracting the mean of its blank

- Blank correction of samples

  - extracting extractant (= sample-blank) data

  - identifying and removing outliers

  - compute the per-plate mean of blanks without outliers

  - correct samples raw data by subtracting the mean of their blank (extractant)

Particularly useful throughout is the automated identification of possible outliers, so that you only need to assess and take outlier-removal decisions for those "suspicious" curves or extractant data.

::: callout-important
## Prerequisites

- data has been imported and tidied, see `vignette("import-tidy", package = "plate2N")`

- (recommended) a preliminary quality check has been run on raw absorbance data, see `vignette("handling-outliers", package = "plate2N")`
:::

# 1 - Getting raw absorbance data

The vignette `import-tidy` shows how to import and tidy absorbance data. The vignette `handling-outliers` shows how to run some preliminary quality checks with the function `qc_raw_abs()` and explains in detail the process of removing outliers once they are identified, which will be needed thoughout this pipeline. To access those vignettes, run the following commands

`vignette("import-tidy", package = "plate2N")`

`vignette("handling-outliers", package = "plate2N")`

Tidy data as imported according to `import-tidy` should look something like this:

```{r}
tidy_plates
```

For quality checking of the standard curve, we will require some metadata for each 96-well plate containing

- at least the concentrations of the dilutions of the standard curve

- anything else that is valid on a per-plate basis (`metadata` has one row per plate) and would be useful in your downstream analysis

`metadata` is an example data set that is part of this package (run `?metadata` to see its documentation).

```{r}
(meta <- metadata |> dplyr::select(dataset, plate_id, std_sp, std_unit, std_conc))
```

We now join (raw) plate data and metadata (`meta`). Note that there has to be a correspondence in the columns `dataset` and `plate_id` between the raw absorbance data and the metadata.

```{r}
# Checking out tidy_plates again and noticing the columns dataset and plate_id
tidy_plates

# joining tidy_plates with meta
(raw_meta <- tidy_plates |> 
  dplyr::left_join(meta, by = dplyr::join_by(dataset, plate_id)))

# Notice the combined structure of both tables
str(raw_meta)
```

# 2 - Blank-correction of standard curves

## 2.1 - Extract curve concentrations

To fit in one row per plate, the concentrations for standard curve are, so far, stored in a compact manner, like this:

```{r}
raw_meta |> dplyr::select(plate_id, std_conc) |> head(n = 3)
```

But in truth, only one of those concentration values corresponds to each value within wells of the standard curve.

Note that concentration values are separated by a `-` and digits are marked by a `.`, which is important in the function call to `extract_curve()`. Also, concentration values MUST be in ascending order in the metadata file. See also `?metadata` and `?extract_curve()`

`extract_curve()` "pulls" the curve concentration data to have only one concentration value per row (under the column `std_conc`), and adds corresponding "row" data in the order defined by the argument `pipetting_direction`.

```{r}
(curve_concentration <- extract_curve(meta, pipetting_direction = "top_down"))
length(unique(raw_meta$plate_id))
```

Notice that `curve_concentration` now has `r nrow(curve_concentration)` rows = nb of plates (`r length(unique(meta$plate_id))`) x nb of elements in each standard curve (8).

::: callout-tip
## Too many rows?

A common mistake is to run `extract_curve()` on `raw_meta` instead of `meta`. But `meta` has one row per plate, whereas `raw_meta` has one row per well, so that calling `extract_curve(raw_meta)` would result in a table that is much too long (\~96 times too long). If you run into issues later, this might be the source.
:::

We now have curve concentrations corresponding to the `dataset`, `plate_id` and `row` of each plate stored inside `curve_concentration`, which we can use for all downstream steps. This goes under the assumption that standard curves are pipetted vertically in a complete column of the 96-well plate. Check `?extract_curve()` for more details. Should other standard curve layouts be required, please reach out.

## 2.2 - Extract Standard data

Here, we extract a subset of the table `raw_meta`, keeping only absorbance reads from the standard curve (defined by the value `Std` in the column `map`), from which we remove the "old" column `std_conc` (with all dilutions of the curve), and replace it by the "new" column `std_conc` as created in `curve_concentration` (only 1 dilution level per row).

```{r}
# Preparing standard curve data
std_data <- raw_meta |> 
  # extract standard curve data
  extract_std_data(std_def = "Std") |> 
  # remove "old" std_conc column
  dplyr::select(!std_conc) |> 
  # adding the "new" std_conc column by joining the data with curve_concentration
  dplyr::left_join(curve_concentration, by = dplyr::join_by(row, dataset, plate_id))

# Check it out (rearranging rows and columns for better readibility)
std_data |> 
  dplyr::arrange(plate_id, column) |> 
  dplyr::relocate(well_id, dataset, plate_id, map, abs, std_conc, std_unit)
```

The correspondence between absorbance and concentration of each standard well is now made, which will allow plotting, outlier removal, and, later, computing the regression equation.

## 2.3 - Extract blank data from standard curve data

`extract_std_blanc()` returns a list with 3 elements concerning standard blanks only (notice that `std_blank$all` has 8x less rows than `std_data` as it only contains the blank of the standard curve, i.e., one well per column)

```{r}
# extract blank data from std_data
std_blank <- std_data |> 
  extract_std_blank(
    std_def = "Std",
    pipetting_direction = "top_down")

# check number of rows
nrow(std_data) ; nrow(std_blank$all)
```

- `std_blank$all` fetches all wells that are expected to contain the standard blank, i.e., all standard data from plate-row A (`pipetting_direction = "top_down"`) or H (`pipetting_direction = "bottom_up"`)

```{r}
  std_blank$all
```

- `std_blank$untrusted` identifies expected blank wells that do not correspond to the lowest absorbance value of their standard curve[^2]. This item may be empty

[^2]: This quality check was added because, in case of top_down pipetting, A1 often will contain the standard blank, but it is also often the first well to be filled during pipetting. With automated pipettes, a small ejection of liquid has to occur before dispensing the first dose of solution. Failure to intentionally eject will result in the first well receiving the "ejection" dose, which is not the correct volume. This is not trivial because with big data sets containing hundreds of plates, it is not unlikely that this human error happens a coupls of times. This can end up in various responses in terms of absorbance, depending on which reagent has been wrongly pipetted.

```{r}
  std_blank$untrusted
```

- `std_blank$trusted` is the complement to `std_blank$untrusted`.

```{r}
  std_blank$trusted
```

## 2.4 - Quality check of std_blank & outlier removal

::: callout-note
## Check out untrusted standard blanks

The computation of per-plate standard curve blank average should be made on trusted blank wells only. It is therefore important to check curves containing "untrusted" blank wells and decide whether to keep them or not
:::

The function `plot_std()` allows the visualization of (a subset of) curves. With small datasets (as here, with only 5 plates), we could run it on all plates. With bigger datasets however, it comes in quite handy to plot only the automated subset of curves that really need a check on their blank data.

Notice the argument `through_origin = FALSE` (default is true), which removes the default constraint on the smooth curve to go through the origin (which does not make sense before we have blank-corrected data).

::: callout-note
## The reference line assumes a linear relationship

`plot_std()` draws a smoothed reference line assuming a standard linear relationship between absorbance and concentration. If your own data is more concentrated and the curve looks noticeably bent rather than straight, don't worry — this is expected for some experiments and is addressed later, when choosing between a linear and polynomial model in the vignette `abs-to-conc`. Nothing about this affects the blank-correction steps covered here.
:::

Notice also that `plot_std()` is based on the ggplot grammar of graphics. This means that you can easily add additional ggplot-based layers like we do in the following chunk.

```{r}
# Select subset of std_data to be plotted because their blanks are "untrusted"
to_plot <- std_data |> 
  dplyr::filter(
    unique_curve_id %in% std_blank$untrusted$unique_curve_id) 

# look at "suspicious" curves
to_plot |> 
  plot_std(through_origin = FALSE) +
  # ggplot layer: facetting per plate (adds the plate_id title and would group curves per plate)
  ggplot2::facet_wrap(~plate_id, scales = "free") +
  # remove legend (because there is only one colour in this case)
  ggplot2::theme(legend.position = "none")
```

If the absorbance in well A is very obviously wrong (like here), then remove those wells for the computation of standard blank average. There are 2 options to do so:

- either by keeping `std_blank$trusted` as is (all "untrusted" wells should indeed be removed)

- or by using `remove_wells()` on `std_blank$all` to generate a better-adapted list of "trusted" standard blank wells (when you decide that some of the "untrusted" wells should be trusted).

Either way, blank averages will then be computed on (really) trusted wells only. Of course, this only works if there were several standard curves on problematic plates, otherwise you will be removing the only `std_blank` of the plate[^3]. Here are 2 examples of how to compute average blanks, along the 2 options mentioned above. Both use the function `std_blank_average()` to compute the blank average.

[^3]: In such cases, you must consider your options. If the inter-plate variability of `std_blank` is sufficiently small, taking an across-dataset or an across-batch mean might do the trick.

```{r}
# Option 1 - We keep std_blank$trusted
std_blank_avg_1 <- std_blank_average(std_blank$trusted)

# Check it out (don't be scared of the "NA" values: you cannot compute a standard deviation on 1 value)
std_blank_avg_1
```

For the second option, we could decide to reject only one of the blank wells from `std_blank$untrusted`: let's say the first one, for plate NO3_1F3.

```{r}
# Option 2 - We create a table of wells to remove
(to_remove <- std_blank$untrusted |> dplyr::filter(plate_id == "NO3_1F3"))

# remove those untrusted wells from std_blank$all ~ new version of std_blank$trusted
std_blank_clean <- std_blank$all |> remove_wells(to_remove)

# compute per-plate blank average from that new trusted table
std_blank_avg_2 <- std_blank_average(std_blank_clean)

# Check it out
std_blank_avg_2
```

Notice that in `std_blank_avg_2`, plate NO3_1F5 now received 2 wells to compute the average from (no NA values under standard deviation and coefficient of variation)

We proposed the 2nd option for the sake of the example, but we will move on with the first option from here on, as the decision to keep the untrusted well for plate NO3_1F3 was obviously misled.

```{r}
std_blank_avg <- std_blank_avg_1
```

::: callout-note
## Outlier check of the whole curve comes later

Throughout this pipeline, we search for and remove outliers prior to critical aggregation steps that will not be trustworthy otherwise (see also vignette `handling-outliers`). With this logic in mind, the outlier removal of other wells[^4] of the standard curve will be done later on, before computing the regression between absorbance and concentration. This is shown in the next vignette, `abs-to-conc`.
:::

[^4]: I.e., wells in the rows "B" to "H" (top-down pipetting) or "A" to "G" (bottom-up pipetting)

## 2.5 - Blank-correction of Standard Curve

Once we are confident in our `std_blank`s, we can use the `std_blank_avg` to blank-correct the raw absorbance values for the whole standard curves. This is done with the function `blank_correct_abs()`, which will also be used to correct sample absorbance data in the next section.

`blank_correct_abs()` is a function that works for standard data and for sample data (@sec-sample-blank), so it's arguments are named in a way that fits both cases. It takes 3 main arguments:

- `raw_wells_data` contains the data of the standard curves. It should be ungrouped within the call to `blank_correct_abs()` and contain only non-blank data (i.e., in the case of "top_down" pipetting, rows B to H)

- `per_plate_avg_blank` takes blank averages (e.g., `std_blank_avg`)

- `map_to_exclude` tells which rows to exclude from `raw_wells_data` based on their value for the column "map". Its default setting fits the blank-correction of sample data (next section), not that of standard data, so we simply replace it by `""`, as we do not wish to exclude any rows here

You can ignore the message `Joining with by = join_by(...)`, which just depends on the additional columns you may have in your data tables

```{r}
# prepare the first argument: ungroup and remove blank data (here: row A)
raw_wells_data <- std_data |>
      dplyr::ungroup() |>
      dplyr::filter_out(row == "A")

# check it out
raw_wells_data

# blank-correct standard curve data
std_corrected <-
  blank_correct_abs(
    raw_wells_data = raw_wells_data,
    per_plate_avg_blank = std_blank_avg,
    map_to_exclude = ""
  ) |> 
  # only keep relevant columns (remove metadata clutter, optional)
  dplyr::select(row:abs_corrected)

# Check it out
std_corrected
```

Notice that the `abs` column has been removed to avoid mistaking it for corrected absorbance. Instead, it has been replaced by `abs_corrected`.

# 3 - Blank-correction of samples {#sec-sample-blank}

## 3.1 - Extract extractant data (sample blank)

In a real world, `raw_meta` may have undergone some cleaning steps (e.g., outlier removal, see `handling-outliers`). In this example dataset, there are always 8 wells attributed to the sample blank (or extractant). Extractant wells are taken from `raw_meta`, where they are found because their mapping (column "map" in `raw_meta`) contains the string "extr", as can be seen here:

```{r}
# look at extractant data
raw_meta |> dplyr::filter(map == "extr")
```

This filtering is also what `extract_extractant()` does in the background, which is the first step taken by the function `extractant_average()` (see below).

```{r}
(extr_data <- extract_extractant(raw_meta))
```

## 3.2 - Compute per-plate average of extractant

The string "extr" is the default of the argument `extr_def` within `extractant_average()` and can be adapted to reflect your mapping. Like `std_blank_average()`, `extractant_average()` computes the per-plate average, standard deviation and coefficient of variation (%) of the blanks.

```{r}
(extr_avg <- extractant_average(raw_meta, extr_def = "extr")) 
```

::: {#tip-multiple_extr .callout-tip}
## More than 1 extractant per plate?

`extractant_average()` also works when there are several extractants per plate, but the argument `extr_def` must be given a vector defining extractants (e.g., `extr_def = c("extr_1", "extr_2")`. Note that it complicates a bit the process down the line because you will need to attribute an extractant to each sample based on something more than just plate identifier. See also `?extractant_average()` for examples and \[the multiple-extractants section\](#sec-multiple-extr)
:::

## 3.3 - Quality check of extractant and outlier removal

### 3.3.1 - Identifying suspicious extractant wells

To show in detail all features of the function calls below, we go through the simple case of when there is only 1 extractant per plate. For cases with 2 or more extractants per plate, see \[the multiple-extractants section\](#sec-multiple-extr)

`plot_blank_var_distrib()` plots a distribution of the coefficient of variation throughout the data set, which can be used as a criterion to identify "suspicious" wells (which becomes relevant in big data sets). Here, we have artificially generated "very wrong" wells for the purpose of the illustration. Note that whereas a threshold of 5% of coefficient of variation is reasonable in simple absorbance-based datasets with 8 replicates (8 wells for the extractant), you may need to adapt your noise-tolerance level for more noise-prone experiments or with fewer replicates.

Notice, once more, that `plot_blank_var_distrib()` relies on the ggplot grammar of graphics, and you can easily add additional ggplot-based layers.

```{r}
plot_blank_var_distrib(extr_avg) +
  # (optional) add threshold line + label
  ggplot2::geom_vline(xintercept = 5, colour = "red", linetype = 2) +
  ggplot2::annotate(
    geom = "label", x = 6, y = 1.5,  hjust = 0.2,
    label = "threshold of 5%", colour = "red")
```

In large data sets, there is bound to be some plate where one or two wells went wrong in the lab, and seeing that there are some plates with much higher variation can be a sign that you need to investigate to remove outliers (like here, with a maximum coefficient of variation of almost 60%). You can take advantage of `dplyr::arrange(desc())` that sorts rows by decreasing values of its argument, to quickly identify suspicious plates (consider removing the call to `desc()` if you have negative values for the coefficients of variation).

```{r}
extr_avg |> 
  dplyr::arrange(desc(blank_coeff_var_percent))
```

Here, we see that the suspicious plates are `NO3_1F2` and `NO3_1F4`. This is easy with a small dataset. For larger data sets, `qc_raw_extr()` performs a quality check of raw absorbance data for extractant wells. It returns a vector containing the "suspicious" `plate_id`s of plates exceeding a user-defined threshold. Note that we work from `raw_meta` because in this pipeline, there has not been any prior outlier removal, but of course, we should work with the "cleanest" data that we have.

A `max_coeff` threshold of 5% is reasonable to segregate "acceptable" coefficients of variation if, like here, you have 8 wells per plate for the extractant.

```{r}
threshold <- 5

suspicious_extr_per_plate <- raw_meta |> 
  qc_raw_extr(suppress_warning = FALSE, max_coeff = threshold)
```

We get a warning message (to suppress it, opt for `suppress_warning = TRUE`). The identifiers of the suspicious plates have been recorded in `suspicious_plate_ids`.

```{r}
suspicious_extr_per_plate
```

Should all plates have a coefficient of variation for the absorbance of the extractant below the threshold, `qc_raw_extr()` sends a happy message, instead of a warning, which can also be suppressed, using `suppress_message = TRUE`. Note that in case there are no suspicious extractant wells, the function does not "return" anything, meaning that if you write `suspicious_extr_per_plate <- raw_meta |> qc_raw_extr(suppress_warning = FALSE, max_coeff = 60)` and try to check out the content of `suspicious_extr_per_plate`, you will only get `NULL`.

```{r}
# run it with a very unreasonably high threshold
raw_meta |> 
  qc_raw_extr(suppress_warning = FALSE, max_coeff = 60)
```

To obtain the full extractant data corresponding to our `suspicious plate_ids`, we can use `suspicious_extr()`. Make sure to use the same threshold.

```{r}
(suspicious_extr <- suspicious_extr(
  raw_meta, 
  suspicious_extr_per_plate = suspicious_extr_per_plate, 
  max_coeff = threshold))
```

Finally, we can use `boxplot_outlier_extr()` to plot extractant plates containing suspicious wells, so that we can decide whether or not to remove some outlier wells, and which ones. Should there be too many plates for a proper visualization, split `suspicious_extr` in subsets.

```{r}
# plot outliers
suspicious_extr |> 
  boxplot_outlier_extr(max_coeff = threshold) 
```

### 3.3.2 - Outlier removal steps

To remove outlier wells, we can then, once more, take advantage of `remove_wells()` (see also above, `?remove_wells()` and the vignette `handling-outliers`).

He have here 3 obvious outliers: wells D8 and C8 in plate 1 (NO3_1F2) and well F8 in plate 2 (NO3_1F4), which we want to remove. The plot given by `boxplot_outlier_extr()` displays all necessary information to do so.

in the next chunk, we create a small tibble that will serve to construct the tibble of wells to remove: first get all unique combinations of dataset and plate_ids from `suspicious_extr`, then add a column "plate_order" that will help checking which plate is which (compared to the plot, useful when several plates are plotted).

```{r}
# Construct tibble to remove
(plate_ids <- suspicious_extr |> 
    dplyr::ungroup() |> 
    dplyr::select(dataset, plate_id) |> 
    unique()) 

# save numbers for plate order as in the plot
(plate_ids <- plate_ids |> 
  dplyr::mutate(plate_order = seq(1, nrow(plate_ids))))
```

Then, we create a vector with wells to remove (reading through boxplots from top to bottom).

:::: callout-tip
## Manually remove outliers

::: {#tip-input-threshold-extr .callout-tip}
In the following chunk, we need to manually decide which wells to remove, based on the boxplots produced above.

- Make sure to deal appropriately with plates that require 2 outliers or no outlier to be removed (see example below)
- Use a number \> nb of plates if there is no plate with 2 or zero outlier
:::
::::

To remove well C8 and D8 from plate 1, and wells F8 from plate 2:

```{r}
#** !!! MANUAL INPUT !!! *

# Which plate needs 2 outliers removed?
plate_with_2_outliers <- c(1) 
# Which plate needs no outlier removed?
plate_without_outliers <- c(9) 

# Which wells are outliers? 
well_ids <- c("C8", "D8", "F8") # only fill in well_ids that need to be removed, in the order of the plates
```

Note that we did not consider the case of having 3 outliers to remove in a single plate. We considered that this situation will be rare enough that it does not justify having a per default step. However, you can easily remove single wells (the 3rd outlier of a plate) by adding an additional row manually to `to_remove`, or just add an additional step of `remove_wells()` with a one-line tibble.

Now, we finish constructing the tibble of wells to be removed.

**TODO: consider making a function out of this**

```{r}
to_remove <- plate_ids |> 
  # double rows for plates with 2 outliers
  dplyr::bind_rows(
    plate_ids |> dplyr::filter(plate_order %in% plate_with_2_outliers)) |> 
  #remove plate without outliers
  dplyr::filter_out(plate_order %in% plate_without_outliers) |> 
  #reorder plates (if some were doubled)
  dplyr::arrange(plate_order) |> 
  # add column with well ids from vector defined above
  dplyr::mutate(well_id = well_ids) |> 
  # remove plate_order(optional)
  dplyr::select(!plate_order)

# check it out
to_remove
```

Always check out that the content of `to_remove` corresponds to the outliers spotted in the plot, as this is a possible source of error, especially for larger datasets.

Now, we remove it from extractant data (which is the same as removing 3 rows)

```{r}
extr_data_clean <- extr_data |> remove_wells(to_remove) 
```

Check out that indeed, 3 rows have been removed:

```{r}
nrow(extr_data) ; nrow(to_remove); nrow(extr_data_clean)
```

Finally, we re-run the average, this time on cleaned, outlier-free, extractant data

```{r}
extr_avg_clean <- extractant_average(
  extractant_data = extr_data_clean, extr_def = "extr") 
```

Check that the highest coefficients of variation are now indeed satisfactory (below our threshold of 5%).

```{r}
extr_avg_clean |> dplyr::arrange(dplyr::desc(blank_coeff_var_percent)) |> head()
```

### 3.3.3 - When there are 2 or more extractants per plate {#sec-multiple-extr}

The steps are the same as with one extractant, but the syntax changes slightly. Let's take the example data set `dbl_extr_plate`. Note that the structure is very similar to what we are used to with `tidy_plates`, but we now have an additional column, called `extr_id`, that contains the correspondence of which sample needs to be blank-corrected from which extractant. Of course, standard curve data have "none" under that column.

```{r}
# check out the example data set
dbl_extr_plate

# compare to tidy_plates
tidy_plates

# Check out mapping for extr_data
dbl_extr_plate |> dplyr::filter(map %in% c("extr_1", "extr_2"))
```

Here is, in brief, how to adapt the same steps as described above:

```{r}
# extracting only extractant data
(extr_data_dbl <- extract_extractant(dbl_extr_plate,extr_def = c("extr_1", "extr_2"))) 

#computing extractant average
(extr_avg_dbl <- extractant_average(dbl_extr_plate, extr_def = c("extr_1", "extr_2"))) 

# Quality checking of coefficient of variation
plot_blank_var_distrib(extr_avg_dbl)

# identifying suspicious plate-extractant combinations
(suspicious_extr_per_plate_dbl <- dbl_extr_plate |> 
  qc_raw_extr(suppress_warning = FALSE, extr_def = c("extr_1", "extr_2"), max_coeff = 5))

# subsetting from extractant data only the plate-extractant combinations that are suspicious
(suspicious_extr_dbl <- suspicious_extr(
  dbl_extr_plate, extr_def = c("extr_1", "extr_2"),
  suspicious_extr_per_plate = suspicious_extr_per_plate_dbl, 
  max_coeff = 5))

# plot outliers
suspicious_extr_dbl |> boxplot_outlier_extr(max_coeff = 5) 
```

Now, adapting the outlier removal steps

::: callout-tip
## Even with several panels, read each plate only once

When determining outliers to remove, still read plates only once. Indeed, even if each plate receives up to 2 boxplots, any well within a boxplot will be unique, as both boxplots come from the data of the same plate. Therefore, inputting outlier removal follows strictly the same steps as above
:::

```{r}
# Construct tibble to remove
(plate_ids_dbl <- suspicious_extr_dbl |> 
    dplyr::ungroup() |> 
    dplyr::select(dataset, plate_id) |> 
    unique()) 

# save numbers for plate order in the plot
(plate_ids_dbl <- plate_ids_dbl |> 
  dplyr::mutate(plate_order = seq(1, nrow(plate_ids_dbl))))

#** !!! MANUAL INPUT !!! *

# Which plate needs 2 outliers removed?
plate_with_2_outliers <- c(2) # if no plate fits this description: give a "too high" number 
# Which plate needs no outlier removed?
plate_without_outliers <- c(1, 4) 

# Which wells are outliers? 
well_ids <- c("C8", "D8", "F8") # only fill in well_ids that need to be removed, in the order of the plates

# complete the table of wells to remove (strictly identical as above)
to_remove_dbl <- plate_ids_dbl |> 
  # double rows for plates with 2 outliers
  dplyr::bind_rows(
    plate_ids_dbl |> dplyr::filter(plate_order %in% plate_with_2_outliers)) |> 
  #remove plate without outliers
  dplyr::filter_out(plate_order %in% plate_without_outliers) |> 
  #reorder plates (if some were doubled)
  dplyr::arrange(plate_order) |> 
  # add column with well ids from vector defined above
  dplyr::mutate(well_id = well_ids) |> 
  # remove plate_order(optional)
  dplyr::select(!plate_order)

# check it out
to_remove_dbl

# remove those wells
extr_data_clean_dbl <- extr_data_dbl |> remove_wells(to_remove_dbl) 

# check out nb of rows
nrow(extr_data_dbl) ; nrow(to_remove_dbl); nrow(extr_data_clean_dbl)

# re-run average
extr_avg_clean_dbl <- extractant_average(
  extractant_data = extr_data_clean_dbl, extr_def = c("extr_1", "extr_2")) 
```

Here we do not re-check that the coefficient of variation is under the threshold because: the data under "extr_2" were artificially changed, but their absorbance readings are really those of samples, therefore the distribution of values was not really presenting the sort of outliers that was expected. But it would be good to run it in real life

**TO DO: the last paragraph is confusing –\> change the raw data so that we can remove this bit and re-run the coefficient of variation as we should**

## 3.4 - Blank-correction of sample absorbance

Now that we are confident in the per-plate average value of raw absorbance of extractant wells, we can finally blank-correct all sample data. `blank_correct_abs()` takes 3 arguments:

- `raw_wells_data` containing the absorbance data that needs blank-correcting

- `per_plate_avg_blank` which contains absorbance data of the extractant, averaged per plate (one row per plate), based on trusted extractant wells

- `map_to_exclude`, a vector of the strings representing non-sample data under the column `map` within `raw_wells_data`, i.e., blanks, standard curves and empty wells. It defaults to `c("empty","Std","extr")`. The output will thus only contain blank-corrected sample data. This means that for (linear) modelling of the standard curve, `std_corrected` will still be needed.

::: callout-note
## An assumption worth noting: spatial uniformity of the blank

`per_plate_avg_blank` is a single value per plate, computed from whichever wells you defined as extractant (or blank) wells, then subtracted uniformly from every sample well on that same plate. This assumes the background absorbance is spatially uniform across the plate — that wells in different positions don't systematically differ in their background reading beyond ordinary well-to-well noise. This is a reasonable assumption for most standard plate readers, but if you have reason to suspect position-dependent optical effects (e.g. known edge effects specific to your instrument), it's worth checking rather than assuming.
:::

```{r}
sample_corrected <- 
  blank_correct_abs(
    raw_wells_data = raw_meta, 
    per_plate_avg_blank = extr_avg_clean,
    map_to_exclude = c("empty","Std","extr")) 

# for case of 2 extractants:
sample_corrected_dbl <- 
  blank_correct_abs(
    raw_wells_data = dbl_extr_plate, 
    per_plate_avg_blank = extr_avg_clean_dbl,
    extr_def = c("extr_1", "extr_2"),
    map_to_exclude = c("empty","Std","extr_1", "extr_2")) 
```

Let's have a look at the output and notice the absence of the value `1` in the column `column` (no data for standard curve[^5]), and of the value `extr` in the column `map` (though it was present in `raw_meta` and still can be found under `extr_id` in the case of 2 extractants per plate, which we kept in case the info is relevant for downstream analysis.

[^5]: In case the blanks are the same for sample wells and standard curve wells, this blank-correction should also be applied on Standard curve data. In this case, remove `"Std"` from the call to `map_to_exclude`.

```{r}
# Check it out
sample_corrected

# "extr" was in raw_meta
raw_meta |> dplyr::select(map) |> dplyr::arrange(dplyr::desc(map))

# "extr" no longer in "sample_corrected
sample_corrected |> dplyr::select(map) |> dplyr::arrange(dplyr::desc(map))

# check out column extr-id
sample_corrected_dbl
```

# 4 - Epilogue

We now have blank-corrected both standard data (`std_corrected`) and sample data (`sample_corrected`), and we can proceed to computing the linear or polynomial model to obtain the regression equation to convert blank-corrected absorbance data to (N-species) concentration data, as is detailed in vignette **`abs-to-conc`** **(under development)**
