---
title: "blank-correction"
vignette: >
  %\VignetteIndexEntry{blank-correction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
```

::: callout-caution
## Work in progress

This vignette is still under development, bugs are to be expected
:::

# TO DO

- Consider making a function of creating the "to_remove" table for extractant outliers
- Split actions within extract_std_blank so that average computes in a separate function, so it can be more easily re-run after outlier removal, then adapt code of the re-run

# Introduction

This vignette shows the basic functions that allow the correction of raw absorbance data (`abs`) to blank-corrected absorbance data, referred to as `abs_corrected` throughout this package.

This pipeline is adapted to the case where the standard curve solutions were prepared with a different blank than the samples. Because of that, there are 2 parallel pipelines to perform the blank-correction (first on standard curves data, then on samples data), then data from standard curves and from samples are merged again.

In theory, the "samples" pipeline should be adaptable to the case where all wells used the same blank (including those containing standard curve solutions), though this has yet to be tested, and bugs may occur. Feel free to contact the authors to suggest improvements.

To avoid confusion between "blank of the standard curve" and "blank of the samples", the sample-blank is referred to as "extractant" or `extr` throughout this vignette, to refer to the solution that was used to extract N-compounds from soil samples and now serves as blank. The standard blank is referred to as `std_blank`.

# 1 - Getting raw absorbance data

The vignette `import-tidy` shows how to import and tidy absorbance data. The vignette `handling-outliers` shows how to run some preliminary quality checks and possibly remove some first outliers. To access those vignettes, run the following commands

`vignette("import-tidy", package = "plate2N")`

`vignette("handling-outliers", package = "plate2N")`

Tidy data as imported according to `import-tidy` should look something like this:

```{r}
tidy_plates
```

For quality checking of the standard curve, we will require some metadata for each 96-well plate containing at least the concentrations of the dilutions of the standard curve.

```{r}
(meta <- metadata |> dplyr::select(dataset, plate_id, std_sp, std_unit, std_conc))
```

We now join (raw) plate data and metadata (meta). Note that there has to be a correspondence in the columns `dataset` and `plate_id` between the raw absorbance data and the metadata.

```{r}
(raw_meta <- tidy_plates |> 
  dplyr::left_join(meta, by = dplyr::join_by(dataset, plate_id)))
```

# 2 - Blank-correction of standard curves

## 2.1 - Extract curve concentrations

To fit in one row per plate, the concentrations for standard curve are, so far, stored in a compact manner, like this:

```{r}
raw_meta |> dplyr::select(plate_id, std_conc) |> head(n = 3)
```

But in truth, only one of those concentration values corresponds to each value of well of the standard curve.

Note that concentration values are separated by a `-` and digits are marked by a `.`, which is important in the function call to `extract_curve()`. Also, concentration values MUST be in ascending order in the metadata file. See also `?metadata` and `?extract_curve()`

```{r}
(curve_concentration <- extract_curve(meta, pipetting_direction = "top_down"))
```

Notice now that each row only contains a single value under the column `std_conc`

::: callout-tip
## Too many rows?

A common mistake is to run `extract_curve()` on `raw_meta` instead of `meta`. But `meta` has one row per plate, whereas `raw_meta` has one row per well, so that calling `extract_curve(raw_meta)` would result on a table that is much too long (\~96 times too long). If you run into issues later, this might be the source.
:::

We now have curve concentrations corresponding to the `dataset`, `plate_id` and `row` of a plate, which we can use for all downstream steps. This goes under the assumption that standard curves are pipetted vertically in a complete column of the 96-well plate. Check `?extract_curve()` for more details.

## 2.2 - Extract Standard data

Here, we get a subset of the table `raw_meta`, keeping only absorbance reads of the wells of the standard curve (defined by the value `Std` in the column `map`), from which we remove the "old" column `std_conc` (with all dilutions of the curve), and replace it by the "new" column `std_conc` (only 1 dilution per row).

```{r}
std_data <- raw_meta |> 
  extract_std_data(std_def = "Std") |> 
  dplyr::select(!std_conc) |> 
  dplyr::left_join(curve_concentration, by = dplyr::join_by(row, dataset, plate_id))

# Check it out (rearranging rows and columns for better readibility)
std_data |> 
  dplyr::arrange(plate_id, column) |> 
  dplyr::relocate(well_id, dataset, plate_id, map, abs, std_conc, std_unit)
```

## 2.3 - Compute per-plate average of std_blank

`extract_std_blanc()` returns a list with several elements concerning standard blanks

```{r}
std_blank <- std_data |> 
  extract_std_blank(
    std_def = "Std",
    pipetting_direction = "top_down")
```

- `std_blank$all` fetches all wells that are expected to contain the standard blank, i.e., all standard data from plate-row A (`pipetting_direction = "top_down"`) or H (`pipetting_direction = "bottom_up"`)

```{r}
  std_blank$all
```

- `std_blank$untrusted` identifies expected blank wells that do not correspond to the lowest absorbance value of their standard curve[^1]. This item and may be empty

[^1]: This quality check was added because, in case of top_down pipetting, A1 often will contain the standard blank, but it is also often the first well to be filled during pipetting. With automated pipettes, a small ejection of liquid has to occur before dispensing the first dose of solution. Failure to intentionally eject will result in the first well receiving the "ejection" dose, which is not the correct volume. This can end up in various responses in terms of absorbance, depending on which reagent has been wrongly pipetted.

```{r}
  std_blank$untrusted
```

- `std_blank$trusted` is the complement to `std_blank$untrusted`.

```{r}
  std_blank$trusted
```

- `std_blank$average` computes the per-plate average of standard blanks (relevant if there are several standard curves pipetted in one plate), as well as the standard deviation and the coefficient of variation.

```{r}
  std_blank$average
```

## 2.4 - Quality check of std_blank & outlier removal

::: callout-note
## Check out untrusted standard blanks

The computation of `std_blank$average` is made on trusted blank wells only. It is therefore important to check curves containing "untrusted" blank wells and decide whether to keep them or not
:::

The function `plot_std()` allows the visualization of (a subset of) curves.

```{r}
# Select subset of std_data to be plotted because curves are in "untrusted"
to_plot <- std_data |> 
  dplyr::filter(
    unique_curve_id %in% std_blank$untrusted$unique_curve_id) 

# look at "suspicious" curves
to_plot |> 
  plot_std(through_origin = FALSE) +
  ggplot2::facet_wrap(~plate_id, scales = "free") +
  ggplot2::theme(legend.position = "none")
```

If the absorbance in well A is very obviously wrong (like here), then remove those wells, either by keeping `std_blank$average` as is[^2], or by using `remove_wells()` and re-computing blank averages (see hereunder). Of course, this only works if there were several standard curves on problematic plates, otherwise you will be removing the only `std_blank` of the plate[^3].

[^2]: As detailed above, `std_blank$average` is already computed from the subset of `std_blank$trusted`, thus, by default, already excludes `std_blank$untrusted`

[^3]: In such cases, you must consider your options. If the inter-plate variability of `std_blank` is sufficiently small, taking an across-dataset or an across-batch mean might do the trick.

Here is an example of how to re-compute average blanks in case we decide to trust some of the wells from `std_blank$untrusted`: let's say the first one, for plate NO3_1F3 (which is obviously wrong in this case).

```{r}
# create artificial version of std_blank$untrusted
untrusted_blanks <- std_blank$untrusted |> dplyr::filter(plate_id == "NO3_1F3")

# remove those untrusted wells from std_blank$all ~ new version of std_blank$trusted
blank_clean <- std_blank$all |> remove_wells(untrusted_blanks)

# re-compute per-plate blank average from that new trusted table
std_blank_avg <- blank_clean |> 
  dplyr::ungroup() |> 
  dplyr::summarise(
    .by = c(dataset, plate_id),
    blank_avg = mean(abs),
    blank_sdev = stats::sd(abs)
  ) |>
  dplyr::mutate(blank_coeff_var_percent = 100 * blank_sdev / blank_avg)
```

In case of a manual re-computation of blank averages, all occurrences of `std_blank$average` below should be replaced by this new `std_blank_avg`.

::: callout-note
## Outlier check of the whole curve comes later

In this pipeline, we search for and remove outliers prior to critical aggregation steps that will not be trustworthy otherwise (see also vignette `handling-outliers`. With this logic in mind, the outlier removal of other wells[^4] in the standard curve will be done later on, before computing the regression between absorbance and concentration. This is shown in the next vignette, `abs-to-conc`.
:::

[^4]: I.e., wells in the rows "B" to "H" (top-down pipetting) or "A" to "G" (bottom-up pipetting)

## 2.5 - Blank-correction of Standard Curve

Once we are confident in our `std_blank`s, we can use the `std_blank$average` (or `std_blank_avg`) to blank-correct the raw absorbance values for the whole standard curves. This is done with the function `blank_correct_abs()`, which will also be used to correct sample absorbance data in the next section.

`blank_correct_abs()` takes 3 main arguments:

- `raw_wells_data` takes the standard curves data. It should be ungrouped and contain only non-blank data (i.e., in the case of "top_down" pipetting, rows B to H)

- `per_plate_avg_blank` takes the version of blank averages that we trust (e.g., `std_blank$average` or `std_blank_avg` if we needed to re-compute untrusted wells

- `map_to_exclude` tells which rows to exclude from `raw_wells_data` based on their value for the column "map". Its default setting fits the blank-correction of sample data (next section), not that of standard data, so we simply replace it by `""`, as we do not wish to exclude any rows here

You can ignore the message `Joining with by = join_by(...)`, which just depends on the additional columns you may have in your data tables

```{r}
# prepare the first argument
raw_wells_data <- std_data |>
      dplyr::ungroup() |>
      dplyr::filter_out(row == "A")

# check it out
head(raw_wells_data)

# blank-correct standard curve data
std_corrected <-
  blank_correct_abs(
    # ungroup std data, remove rows with the blanks (here: row A)
    raw_wells_data = raw_wells_data,
    per_plate_avg_blank = std_blank$average,
    map_to_exclude = ""
  ) |> 
  # only keep relevant columns (remove metadata clutter, optional)
  dplyr::select(row:abs_corrected)

# Check it out
std_corrected
```

Notice that the `abs` column has been removed to avoid mistaking it for corrected absorbance. Instead, it has been replaced by `abs_corrected`.

# 3 - Blank-correction of samples

## 3.1 - Extract extractant data (sample blank)

In a real world, `raw_meta` will have probably undergone some cleaning steps (e.g., outlier removal, see `handling-outliers`). In this example dataset, there are always 8 wells attributed to the sample blank (or extractant), which is found because its mapping (column "map" in `raw_meta`) contains the string "extr", as can be seen here:

```{r}
# look at extractant data
raw_meta |> dplyr::filter(map == "extr")
```

This filtering is also what `extract_extractant()` does in the background, which is the first step of `extractant_average()` (see below).

```{r}
extr_data <- extract_extractant(raw_meta)

# Check it out
extr_data
```

## 3.2 - Compute per-plate average of extractant

This string "extr" is the default of the argument `extr_def` of `extractant_average()` and can be adapted to reflect your mapping. Like `extract_std_blank(..)$average`, `extractant_average()` computes the average, standard deviation and coefficient of variation (%) of the blanks.

```{r}
(extr_avg <- extractant_average(raw_meta, extr_def = "extr")) 
```

::: callout-tip
## More than 1 extractant per plate?

`extractant_average()` should also work when there are several extractants per plate, but the argument `extr_def` must be given a vector with extractant (e.g., `extr_def = c("extr_1", "extr_2")`. See also `?extractant_average()` for examples.
:::

## 3.3 - Quality check of extractant and outlier removal

### 3.3.1 - One extractant per plate

`plot_blank_var_distrib()` plots a distribution of this coefficient of variation throughout the data set (which becomes relevant in big data sets).

```{r}
plot_blank_var_distrib(extr_avg)
```

In big data sets, there is bound to be some plate where one or two wells went wrong in the lab, and seeing that there are some plates with much higher variation can be a sign that you need to investigate to remove outliers (like here, with a maximum coefficient of variation of almost 60%). You can take advantage of `dplyr::arrange(desc())` that sorts rows by decreasing values of its argument, to quickly identify suspicious plates (consider removing the call to `desc()` if you have negative values for the coefficients of variation).

```{r}
extr_avg |> 
  dplyr::arrange(desc(blank_coeff_var_percent))
```

Here, we see that the suspicious plates are `NO3_1F2` and `NO3_1F4`. This is easy with a small dataset. For larger data sets, `qc_raw_extr()` performs a quality check of raw absorbance data for extractant wells. It returns a vector containing the "suspicious" `plate_id`s of plates exceeding a user-defined threshold. Note that we work from `raw_meta` because in this pipeline, there has not been any prior outlier removal, but of course, we should work with the "cleanest" data that we have.

A `max_coeff` threshold of 5% is reasonable to segregate "acceptable" coefficients if, like here, you have 8 wells per plate for the extractant.

```{r}
threshold <- 5

suspicious_extr_per_plate <- raw_meta |> 
  qc_raw_extr(suppress_warning = FALSE, max_coeff = threshold)
```

We get a warning message (to suppress it, opt for `suppress_warning = TRUE`). The identifiers of the suspicious plates have been recorded in `suspicious_plate_ids`.

```{r}
suspicious_extr_per_plate
```

Should all plates have a coefficient of variation for the absorbance of the extractant below the threshold, `qc_raw_extr()` returns a happy message, instead of a warning, which can also be suppressed, using `suppress_message = TRUE`.

```{r}
# run it with a very unreasonably high threshold
raw_meta |> 
  qc_raw_extr(suppress_warning = FALSE, max_coeff = 60)
```

To obtain the full extractant data corresponding to our `suspicious plate_ids`, we can use `suspicious_extr()`.

```{r}
(suspicious_extr <- suspicious_extr(
  raw_meta, 
  suspicious_extr_per_plate = suspicious_extr_per_plate, 
  max_coeff = threshold))
```

Finally, we can use `boxplot_outlier_extr()` to plot extractant plates containing suspicious wells, so that we can decide whether or not to remove some outlier wells. To do so, we can, once more, take advantage of `remove_wells()`, see also above, `?remove_wells()` and the vignette `handling-outliers`. Should there be too many plates for a proper visualization, split `suspicious_extr` in subsets.

```{r}
# plot outliers
suspicious_extr |> boxplot_outlier_extr(max_coeff = threshold) + ggplot2::facet_wrap(~map)
```

### 3.3.2 - Outlier removal steps

He have here 2 very obvious outliers: wells D8 and C8 in plate 1 (NO3_1F2) and well F8 in plate 2 (NO3_1F4), which we want to remove. The plot given by `boxplot_outlier_extr()` gives all necessary information to do so. First, we create a small tibble that will serve to construct the tibble of wells to remove: first get dataset and plate_ids from `suspicious_extr`, then add a column "plate_order" that will help checking which plate is which (compared to the plot, useful when several plates are plotted).

```{r}
# Construct tibble to remove
(plate_ids <- suspicious_extr |> 
    dplyr::ungroup() |> 
    dplyr::select(dataset, plate_id) |> 
    unique()) 

# save numbers for plate order in the plot
(plate_ids <- plate_ids |> 
  dplyr::mutate(plate_order = seq(1, nrow(plate_ids))))
```

Then, we create a vector with wells to remove (reading through boxplots from top to bottom).

:::: callout-tip
## Manually remove outliers

::: {#tip-input-threshold-extr .callout-tip}
In the following chunk, we need to manually decide which wells to remove, based on the boxplots produced above.

- Make sure to deal appropriately with plates that require 2 outliers or no outlier to be removed (see example below)
- Use a number \> nb of plates if there is no plate with 2 or zero outlier
:::
::::

To remove well C8 and D8 from plate 1, and wells F8 from plate 2:

```{r}
#** !!! MANUAL INPUT !!! *

# Which plate needs 2 outliers removed?
plate_with_2_outliers <- c(1) 
# Which plate needs no outlier removed?
plate_without_outliers <- c(9) 

# Which wells are outliers? 
well_ids <- c("C8", "D8", "F8") # only fill in well_ids that need to be removed, in the order of the plates
```

Then we finish constructing the tibble of wells to be removed.

**TODO: consider making a function out of this**

```{r}
to_remove <- plate_ids |> 
  # double rows for plates with 2 outliers
  dplyr::bind_rows(
    plate_ids |> dplyr::filter(plate_order %in% plate_with_2_outliers)) |> 
  #remove plate without outliers
  dplyr::filter_out(plate_order %in% plate_without_outliers) |> 
  #reorder plates (if some were doubled)
  dplyr::arrange(plate_order) |> 
  # add column with well ids from vector defined above
  dplyr::mutate(well_id = well_ids) |> 
  # remove plate_order(optional)
  dplyr::select(!plate_order)

# check it out
to_remove
```

Always check out that the content of `to_remove` corresponds to the outliers spotted in the plot, as this is a possible source of error, especially for larger datasets.

Now, we remove it from extractant data (which is the same as removing 3 rows)

```{r}
extr_data_clean <- extr_data |> remove_wells(to_remove) 
```

Check out that indeed, 3 rows have been removed:

```{r}
nrow(extr_data) ; nrow(to_remove); nrow(extr_data_clean)
```

Finally, we re-run the average on cleaned extractant data

```{r}
extr_avg_clean <- extractant_average(
  extractant_data = extr_data_clean, extr_def = "extr") 
```

Check that the highest coefficients of variation are now indeed satisfactory (below our threshold of 5%).

```{r}
extr_avg_clean |> dplyr::arrange(dplyr::desc(blank_coeff_var_percent)) |> head()
```

### 3.3.3 - 2 or more extractants per plate

The steps are the same as with one extractant, but the syntax changes slightly. Let's take the example data set `dbl_extr_plate`:

```{r}
dbl_extr_plate
```

Here is, in brief, how to adapt the same steps as described above:

```{r}
# extracting only extractant data
(extr_data_dbl <- extract_extractant(dbl_extr_plate,extr_def = c("extr_1", "extr_2"))) 

#computing extractant average
(extr_avg_dbl <- extractant_average(dbl_extr_plate, extr_def = c("extr_1", "extr_2"))) 

# Quality checking of coefficient of variation
plot_blank_var_distrib(extr_avg_dbl)

# identifying suspicious plate-extractant combinations
(suspicious_extr_per_plate_dbl <- dbl_extr_plate |> 
  qc_raw_extr(suppress_warning = FALSE, extr_def = c("extr_1", "extr_2"), max_coeff = 5))

# subsetting from extractant data only the plate-extractant combinations that are suspicious
(suspicious_extr_dbl <- suspicious_extr(
  dbl_extr_plate, extr_def = c("extr_1", "extr_2"),
  suspicious_extr_per_plate = suspicious_extr_per_plate_dbl, 
  max_coeff = 5))

# plot outliers
suspicious_extr_dbl |> boxplot_outlier_extr(max_coeff = 5) 


```

Now, adapting the outlier removal steps

::: callout-tip
## Even with facetting, read each plate once

When determining outliers to remove, still read plates only once. Indeed, even if each plate receives up to 2 boxplots, any well within a boxplot will be unique, as both boxplots come from the data of the same plate. Therefore, inputting outlier removal follows strictly the same steps as above
:::

```{r}
# Construct tibble to remove
(plate_ids_dbl <- suspicious_extr_dbl |> 
    dplyr::ungroup() |> 
    dplyr::select(dataset, plate_id) |> 
    unique()) 

# save numbers for plate order in the plot
(plate_ids_dbl <- plate_ids_dbl |> 
  dplyr::mutate(plate_order = seq(1, nrow(plate_ids_dbl))))

#** !!! MANUAL INPUT !!! *

# Which plate needs 2 outliers removed?
plate_with_2_outliers <- c(2) # if no plate fits this description: give a "too high" number 
# Which plate needs no outlier removed?
plate_without_outliers <- c(1, 4) 

# Which wells are outliers? 
well_ids <- c("C8", "D8", "F8") # only fill in well_ids that need to be removed, in the order of the plates

# complete the table of wells to remove (strictly identical as above)
to_remove_dbl <- plate_ids_dbl |> 
  # double rows for plates with 2 outliers
  dplyr::bind_rows(
    plate_ids_dbl |> dplyr::filter(plate_order %in% plate_with_2_outliers)) |> 
  #remove plate without outliers
  dplyr::filter_out(plate_order %in% plate_without_outliers) |> 
  #reorder plates (if some were doubled)
  dplyr::arrange(plate_order) |> 
  # add column with well ids from vector defined above
  dplyr::mutate(well_id = well_ids) |> 
  # remove plate_order(optional)
  dplyr::select(!plate_order)

# check it out
to_remove_dbl

# remove those wells
extr_data_clean_dbl <- extr_data_dbl |> remove_wells(to_remove_dbl) 

# check out nb of rows
nrow(extr_data_dbl) ; nrow(to_remove_dbl); nrow(extr_data_clean_dbl)

# re-run average
extr_avg_clean_dbl <- extractant_average(
  extractant_data = extr_data_clean_dbl, extr_def = c("extr_1", "extr_2")) 
```

Here we do not re-check that the coefficient of variation is under the threshold because: the data under "extr_2" were artificially changed, but their absorbance readings are actually those of samples, therefore the distribution of values was not really presenting the sort of outliers that was expected. But it would be good to run it in real life

## 3.4 - Blank-correction of sample absorbance

Now that we are confident in the per-plate average value of raw absorbance of extractant wells, we can finally blank-correct all sample data. `blank_correct_abs()` takes 3 arguments:

- `raw_wells_data` containing the absorbance data that needs blank-correcting

- `per_plate_avg_blank` which contains absorbance data of the extractant, averaged per plate (one row per plate), based on trusted extractant wells

- `map_to_exclude`, a vector of the strings representing non-sample data under the column `map` within `raw_wells_data`, i.e., blanks, standard curves and empty wells. It defaults to `c("empty","Std","extr")`. The output will thus only contain blank-corrected sample data. This means that for (linear) modelling of the standard curve, `std_corrected` will still be needed.

```{r}
sample_corrected <- 
  blank_correct_abs(
    raw_wells_data = raw_meta, 
    per_plate_avg_blank = extr_avg_clean,
    map_to_exclude = c("empty","Std","extr")) 
```

Let's have a look at the output and notice the absence of the value `1` in the column `column` (no data for standard curve[^5]), and of the value `extr` in the column `map` (though it is present in `raw_meta`.

[^5]: In case the blanks are the same for sample wells and standard curve wells, this blank-correction should also be applied on Standard curve data. In this case, remove `"Std"` from the call to `map_to_exclude`.

```{r}
# Check it out
sample_corrected

# "extr" was in raw_meta
raw_meta |> dplyr::select(map) |> dplyr::arrange(dplyr::desc(map))

# "extr" no longer in "sample_corrected
sample_corrected |> dplyr::select(map) |> dplyr::arrange(dplyr::desc(map))
```

# 4 - Epilogue

We now have blank-corrected both standard data (`std_corrected`) and sample data (`sample_corrected`), and we can proceed to computing the linear or polynomial model to obtain the regression equation to convert blank-corrected absorbance data to (N-species) concentration data, as is detailed in vignette **`abs-to-conc`** **(under development)**
