---
title: "blank-correction"
vignette: >
  %\VignetteIndexEntry{blank-correction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
```

::: callout-caution
## Work in progress

This vignette is still under development, bugs are to be expected
:::

# TO DO

- Make sure there is a "bad" curve in the example dataset (with blank with higher absorbance than row B)

# Introduction

This vignette shows the basic functions that allow the correction of raw absorbance data (`abs`) to blank-corrected absorbance data, referred to as `abs_corrected` throughout this package.

This pipeline is adapted to the case where the standard curve solutions were prepared with a different blank than the samples. Because of that, there are 2 parallel pipelines to perform the blank-correction, then data from standard curves and from samples are merged again.

In theory, the "samples" pipeline should be adaptable to the case where all wells used the same blank (including those containing standard curve solutions), though this has yet to be tested, and bugs may occur. Feel free to contact the authors to suggest improvements.

To avoid confusion between "blank of the standard curve" and "blank of the samples", the sample-blank is referred to as "extractant" or `extr` throughout this vignette, to refer to the solution that was used to extract N-compounds from soil samples and now serves as blank. The standard blank is referred to as `std_blank`.

# 1 - Getting raw absorbance data

The vignette `import-tidy` shows how to import and tidy absorbance data. The vignette `handling-outliers` shows how to run some preliminary quality checks and possibly remove some first outliers. To access those vignettes, run the following commands

`vignette("import-tidy", package = "plate2N")`

`vignette("handling-outliers", package = "plate2N")`

```{r}
tidy_plates
```

For blank-correction of the standard curve, we will also require some metadata for each 96-well plate containing at least the concentrations of the dilutions of the standard curve

```{r}
(meta <- metadata |> dplyr::select(dataset, plate_id, std_sp, std_unit, std_conc))
```

We now join (raw) plate data and metadata (meta)

```{r}
raw_meta <- tidy_plates |> 
  dplyr::left_join(meta, by = dplyr::join_by(dataset, plate_id))
```

# 2 - Blank-correction of standard curves

## 2.1 - Extract curve concentrations

To fit in one row per plate, the concentrations for standard curve are, so far, stored in a compact manner, like this:

```{r}
raw_meta |> dplyr::select(plate_id, std_conc) |> head(n = 3)
```

Note that concentration values are separated by a `-` and digits are marked by a `.`, which is important in the function call to `extract_curve()`. Also, concentration values MUST be in ascending order in the metadata file. See also `?metadata` and `?extract_curve()`

```{r}
(curve_concentration <- extract_curve(meta, pipetting_direction = "top_down"))
```

::: callout-tip
## Too many rows?

A common mistake is to run `extract_curve()` on `raw_meta` instead of `meta`. But `meta` has one row per plate, whereas `raw_meta` has one row per well, so that calling `extract_curve(raw_meta)` would result on a table that is much too long (\~96 times too long). If you run into issues later, this might be the source.
:::

We now have curve concentrations connected to the dataset, plate_id and row of a plate, which we can use for all downstream steps. This goes under the assumption that standard curves are pipetted vertically in a complete column of the 96-well plate. Check `?extract_curve()` for more details.

## 2.2 - Extract Standard data

```{r}
std_data <- raw_meta |> 
  extract_std_data(std_def = "Std") |> 
  dplyr::select(!std_conc) |> 
  dplyr::left_join(curve_concentration, by = dplyr::join_by(row, dataset, plate_id))
std_data
```

## 2.3 - Compute per-plate average of std_blank

`extract_std_blanc()` returns a list with several elements concerning standard blanks

- `std_blank$all` fetches all wells that are expected to contain the standard blank (all standard data from plate-row A (`pipetting_direction = "top_down"`) or H (`pipetting_direction = "bottom_up"`)

- `std_blank$untrusted` identifies expected blank wells that do not correspond to the lowest absorbance value of their standard curve[^1]. This item and may be empty

- `std_blank$trusted` is the complement to `std_blank$untrusted`.

- `std_blank$average` computes the per-plate average of standard blanks (relevant if there are several standard curves pipetted in one plate), as well as the standard deviation and the coefficient of variation.

[^1]: This quality check was added because, in case of top_down pipetting, A1 often will contain the standard blank, but it is also often the first well to be filled during pipetting. With automated pipettes, a small ejection of liquid has to occur before dispensing the first dose of solution. Failure to intentionally eject will result in the first well receiving the "ejection" dose, which is not the correct volume. This can end up in various responses in terms of absorbance, depending on which reagent has been wrongly pipetted.

```{r}
std_blank <- std_data |> 
  extract_std_blank(
    std_def = "Std",
    pipetting_direction = "top_down")

std_blank$all

std_blank$average

std_blank$untrusted ; std_blank$trusted
```

## 2.4 - Quality check of std_blank & outlier removal

::: callout-note
## Check out untrusted standard blanks

The computation of `std_blank$average` is made on trusted blank wells only. It is therefore important to check curves containing "untrusted" blank wells and decide whether to keep them or not
:::

The function `plot_std()` allows the visualization of a subset of curves. Because there is no untrusted blank for now (**Make sure there is a "bad" curve in the example dataset**), we artificially take a subset of `std_blank$all` just for the purpose of demonstrating the plots.

```{r}
# Subset: create an artificial std_data
artificial_std <- std_data |> 
  dplyr::filter(
    unique_curve_id %in% std_blank$all$unique_curve_id[1:4]) 

# look at (fake) "suspicious" curves
artificial_std |> 
  plot_std(through_origin = FALSE) +
  ggplot2::facet_wrap(~plate_id, scales = "free") +
  ggplot2::theme(legend.position = "none")
```

If the absorbance in well A is very obviously wrong, then remove those wells, either by keeping `std_blank$average` as is, or by using `remove_wells()` and re-computing blank averages. Of course, this only works if there were several standard curves on problematic plates, otherwise you will be removing the only std_blank of the plate. In such cases, you must consider your options. If the inter-plate variability of std_blank is sufficiently small, taking an across-dataset or an across-batch mean might do the trick.

Here is an example of how to re-compute average blanks in case we decide to trust some of the wells from `std_blank$untrusted`, replaced here by `artificial_blank_untrusted` because our untrusted example is empty (for now).

```{r}
# create artificial version of std_blank$untrusted
(artificial_blank_untrusted <- (artificial_std |> extract_std_blank(std_def = "Std"))$all)

# remove those untrusted wells from std_blank$all ~ new version of std_blank$trusted
(artificial_blank_clean <- std_blank$all |> remove_wells(artificial_blank_untrusted))

# re-compute per-plate blank average from that new trusted table
artificial_std_blank_avg <- artificial_blank_clean |>
  dplyr::ungroup() |> 
  dplyr::summarise(
    .by = c(dataset, plate_id),
    blank_avg = mean(abs),
    blank_sdev = stats::sd(abs)
  ) |>
  dplyr::mutate(blank_coeff_var_percent = 100 * blank_sdev / blank_avg)

```

In case of a manual re-computation of blank avergaes, all occurences of std_blank\$average below should be replaced by this new `artificial_std_blank_avg`.

## 2.5 - Blank-correction of Standard Curve

Once we are confident in our `std_blank`s, we can use the `std_blank$average` to blank-correct the raw absorbance values for the whole standard curves. This is done with the function `blank_correct_abs()`, which will also be used to correct sample absorbance data in the next section.

`blank_correct_abs()` takes 3 main arguments.

- `raw_wells_data` takes the standard curves data. It should be ungrouped and contain only non-blank data (in the case of "top_down" pipetting, rows B to H)

- `per_plate_avg_blank` takes the version of blank averages that we trust (e.g., `std_blank$average` or `artificial_std_blank_avg)`

- `map_to_exclude` tells which rows to exclude from `raw_wells_data` based on their value for the column "map". Its default setting fits the blank-correction of sample data, not that of standard data, so we simply replace it by `""`, as we do not wish to exclude any rows here

You can ignore the message `Joining with by = join_by(...)`, which just depends on the additional columns you may have in your data tables

```{r}
std_corrected <-
  blank_correct_abs(
    raw_wells_data = std_data |>
      dplyr::ungroup() |>
      dplyr::filter_out(row == "A"),
    per_plate_avg_blank = std_blank$average,
    map_to_exclude = ""
  ) |> 
  # only keep relevant columns (remove metadata clutter)
  dplyr::select(row:abs_corrected)
std_corrected
```

Notice that the `abs` column has been removed to avoid mistaking it for corrected absorbance. Instead, it has been replaced by `abs_corrected`.

```{r}

```

# 3 - Blank-correction of samples

## plot_blank_var_distrib() (& extractant_average()?)

## qc_raw_extr()

## suspicious_extr()

## plot extractant outliers with boxplot_outlier_extr() or multiplot_outlier_extr()
