---
title: "handling-outliers"
vignette: >
  %\VignetteIndexEntry{handling-outliers}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
```

::: callout-caution
## Work in progress

This vignette is still under development, bugs are to be expected
:::

# TO DO

- Add reference to outlier steps for Std blank, std curve and extr blank

# Introduction

This vignette introduces the handling of several functions to help deal with outliers, at several steps of the data transformation pipeline.

[**--\> Briefly introduce all functions**]{.underline}

# 1 - Manual removal of single wells

It is not unusual to have to manually remove wells from a data set. This is often due to pipetting errors, whether we know that we failed a well in the lab, or we identify aberrant values during data processing (see also **vignette xx - upcoming** for how to identify such wells). `remove_wells()` does just that.

According to the logic of this package, 3 pieces of information uniquely define a well: `dataset`, `plate_id` and `well_id`. The function `remove_wells()` allows manual removal of wells from a table formated as `tidy_table`, taking a small tibble such as `failed_wells_example` to decide which wells to remove.

```{r}
# check out input
tidy_table ; failed_wells_example

# remove wells
(clean_table <- tidy_table |> remove_wells(failed_wells_example))

# check how many rows were removed
tidy_table |> nrow() ; failed_wells_example |> nrow() ; clean_table |> nrow()
```

# 2 - Defining wells to remove

In **vignette xx - upcoming**, we go in detail into how to identify potential outlier wells at various stages of the analytical pipeline. `failed_wells_example` has a very simple structure that can easily be created either in R or imported from a csv file.

## 2.1 - From a csv

It can he handy to have a printed table in the lab, to quickly fill in by hand the information on wells where we know that something went wrong. For this, the function `failed_wells_template()` quickly generates a table that can be exported as a csv, printed, then manually encoded and imported back into R. By default, it will generate a csv with 30 rows.

```{r}
(failed_wells_template <- failed_wells_template(nrow = 30))
```

Exporting as a csv can be done with

`failed_wells_template |> readr::write_csv("path/to/folder/failed_wells_template.csv")`

Importing it back can be done with

`readr::read_csv("path/to/folder/failed_wells_lab.csv")`

## 2.2 - Manually in R

**Option 1:** use `dplyr::select()` and `dplyr::filter()` to extract `failed_wells` from an existing table

```{r}
failed_wells <- tidy_table |> 
  dplyr::select(dataset, plate_id, well_id) |> 
  dplyr::filter(
    dataset == "expe1" &
    ((plate_id == "M16" & well_id == "A1") |
       (plate_id == "M23" & well_id == "H12"))
  )
```

**Option 2:** Encode it manually from scratch, using `tibble::tibble()`.

```{r}
failed_wells <- tibble::tibble(
  dataset = rep("expe1", 2),
  plate_id = c("M16", "M23"),
  well_id = c("A1", "H12")
)
```

# 3 - Quality Check of raw absorbance readings

There is a range of values that are deemed acceptable for absorbance readings. Whereas there is a long-lasting tradition of setting "acceptable" absorbance values between 0.1 and 1, in reality, the acceptable range will depend on experiment, usage and quality of the spectrophotomer. Note that for N dosage, values lower than 0.1 are not rare, especially for blanks and lower values in the standard curve.

`qc_raw_abs()` checks raw absorbance values in a dataset and returns a table allowing the identification of suspicious wells.

By default, `qc_raw_abs()` also draws a plot of the distribution of raw absorbance values in the data set, and sends either a message (all is well) or a warning (some wells are out of range). These outputs can be repressed. See `?qc_raw_abs()` for more details.

Default range of accepted absorbance is \[0.1, 1\].

```{r}
# check out the raw table
tidy_plates

# run the function
(suspicious_wells <- qc_raw_abs(tidy_plates, min_abs = 0.09, max_abs = 1))
```

It can happen that some wells are very striking, e.g., a single well with an absorbance of 3 where the rest of the dataset is around 1 and the experiment has been ran in 4 technical replicates (there should always be 4 wells for each value). In such cases, the output of `qc_raw_abs()` can be used in combination with `remove_wells()`, possibly with additional selection and filtering steps as shown above.

```{r}
tidy_plates |> remove_wells(suspicious_wells)
```
