---
title: "import-tidy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{import-tidy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
**TO DO**

- Create a function to prepare a csv template to be filled (by hand) 
- Create a function to generate a csv from a sample list 

# Introduction
The goal of `plate2N` is to facilitate data wrangling and data analysis for raw data that originates from absorbance readings with a 96-well plate reader (spectrophotometry). 
```{r setup}
library(plate2N)
```

This particular vignette adresses the first steps of the pipeline: getting from raw data in various forms to a tidy data table that can be easily manipulated in downstream steps. Here, we cover two steps:

**1) Data import** from various raw file formats with `txt_to_tibble()`, `csv_to_tibble()` and `skanit_to_tibble()`, where data is brought into a shape that is common to all import file. We refer to the output of this first step in an object called "tibble" where data is still in a "plate" format (see hereunder). In this tibble, the first columns contains the plate id in the first row, followed by the letters caracterising the 8 rows of the plate (from A to H). The rest of the first row contains the plate-column numbers (from 1 to 12). The rest of the table contains the original plate data, which can be numerical (e.g., absorbance data) or characters (e.g., plate mapping)
```{r}
tibble_example
```

**2) Transformation** of the "tibble" into a tidy data table that is easily usable for downstream data transformations and analyses, with `verticalize_plates()`. The tidy table presents data in a verticalized format, i.e., data from each plate corresponds to a single column where the name of the column is the plate id. The first 2 columns of those verticalized plates contain the well identifier, in the columns "row" (of the plate, from A to H), and "column" (from 1 to 12). See example hereunder
```{r}
vertical_plates
```

Note: If the need arises to fit this package to other plate formats than 96-well plates, please reach out to the authors.

# 1 - Import

All 3 import functions hereunder (`csv_to_tibble()`, `skanit_to_tibble()` and `txt_to_tibble()`) can be used to import absorbance **and** mapping data into a single tibble. However, in that case, it is recommended to attribute different plate ids to the 2 "plates" that represent the same plate (e.g., "NO3_plate1_abs" and "NO3_plate1_map", see also description of `verticalize_plates()` in section 2 to see the final output). 

## 1.1 - From a .csv file

The function `csv_to_tibble()` imports plate data from a csv file where data is structured in a specific way (see hereunder). For other csv formats, see sections below about "skanit csv" and "other formats".
The required data structure is strictly the same as in the outputted tibble (see `tibble_example` hereabove). 

In that sense, the function `csv_to_tibble()`constitutes the simplest import option and takes advantage of the `readr` package (part of the `tidyverse`), by using either the function `read_csv()` or `read_csv2()`, for comma-separated or semi-colon separated values, respectively, which can be modulated by the parameter `delim`, which defaults with the value `","` (`read_csv()`), but accepts `";"` (`read_csv2()`). 

Here is an example of how to call the `csv_to_tibble()` function.

```{r}
# see the function in action
example_csv <- system.file("extdata", "csv_example.csv", package = "plate2N")
preview_csv <- csv_to_tibble(example_csv)
print(preview_csv, n = 20)
```

## 1.2 - From a skanit .csv

A "skanit csv" is a csv where data is structured also in a plate format, but where plates are organized differently, according to a format that is generated by a software called *Skanit* and is linked to the plate reader. The raw csv format can be observed in the raw data attached to this package and is displayed hereunder.

**It would make no sense to manually create a csv file in the "skanit csv" format**, and using the function `skanit_to_tibble()` is only useful for users of the Skanit software. See also `?skanit_to_tibble()` for more details.

```{r}
library(readr)
skanit_csv <- system.file("extdata", "skanit.csv", package = "plate2N")
preview_skanit_raw <- readr::read_csv(skanit_csv, col_names = FALSE, show_col_types = FALSE)
print(preview_skanit_raw, n = 40)
```

Here is an example of how to call the `skanit_to_tibble()` function and see its output. Notice the identical structure between `skanit$abs_tibble` and `skanit$map_tibble`.

```{r}
skanit <- skanit_to_tibble(skanit_csv)
skanit_tibble_abs <- skanit$abs_tibble
skanit_tibble_map <- skanit$map_tibble
print(skanit_tibble_abs, n = 12) ; print(skanit_tibble_map, n = 12)
```


## 1.3 - From .TXT files

The function `txt_to_tibble()` has its own specific logic. Indeed, whereas both `csv_to_tibble()` and `skanit_to_tibble()` take a single **file** as input, `txt_to_tibble()` takes a single **folder** as input, in which an indefinite amount of single .TXT files can be found. It is however important that the folder contains no other .TXT files than the input data, though it may contain other file types. More details on file requirements can be found under `?txt_to_tibble()`. Here is an example of folder containing 5 .TXT files.

```{r}
# path to the folder containing .TXT plate files
(txt_folder <- system.file("extdata", "txt_examples", package = "plate2N"))

#examine contents of txt_folder 
(files <- list.files(txt_folder))
file <- paste0(txt_folder, "/", files[1]) # an example of file to look up in your computer
```

Each .TXT file contains the absorbance data of a single plate. This means that, to create the `tibble`, `txt_to_tibble()` iterates through all .TXT files in a given folder (`txt_folder` in the example above), extracts their plate data, adds the plate id in the upper left corner of each plate data, then combines all plates into a single `tibble`.

To examine the aspect of the raw file, go and manually visit the file path displayed hereunder for `txt_folder` in your computer's file directory. Alternatively, you can see an example of raw .TXT file on our github repository: *Add here a link to something*. Indeed, the raw file structure cannot be easily shown here because import functions return error messages unless we add some additional tweaking to extract only the plate data. These error messages come because the column-structure of the .TXT file is confusing to the import functions (see tentative below). Therefore the import with `readr::read_tsv()` already needs to skip some lines and extracts only the plate data, as shown below, and cannot display the raw file structure. 

```{r}
# import the file "properly" (as in `txt_to_tibble()`), and loose the raw structure of the original .TXT
extracted_plate_data <- readr::read_tsv(file, col_names = TRUE, skip = 5, show_col_types = FALSE, name_repair = "unique_quiet", col_types = cols(.default = col_character())) |>
      tidyr::drop_na()
extracted_plate_data
```

Just as for `skanit_to_tibble()`, it would make little sense to use `txt_to_tibble()` to extract plate data from .TXT files that have been generated in another raw structure than the one provided here. Indeed, just as the "skanit csv", the "txt" structure was automatically generated by our plate reader (*Add here reference of the plate reader*) when data is not exported through the skanit software, but indeed is directly exported as txt.

```{r}
tibble_txt <- txt_to_tibble(filepath = txt_folder)
print(tibble_txt, n = 20)
```

**Note:** Should you have raw .TXT files with a different structure, please reach out, and we can propose an adapted importing step into our function definition.

## 1.4 - From other formats

Should you have data in different raw formats, there are 2 options.

**1) With small data sets**, a quick option would be to simply copy-paste your raw plate data into a clean csv sheet (outside of R, for example in Excel), adopting the raw structure of `tibble_example`. You can generate an empty csv template using the function *Add here the name of other funtion*. Or you can lookup the csv structure of the following file:
```{r}
(example_csv <- system.file("extdata", "csv_example.csv", package = "plate2N"))
```

**2) With bigger data sets**, we recommend using an export function from the `readr` package (e.g., `readr::read_csv()`, `readr::read_tsv()` or `readr::read_delim()`), possibly with a selection of columns or rows, then using tools of the `dplyr` or `tidyr` packages to bring the resulting tibble into the required format.

# 2 - Tidy table

Finally, once the `tibble` has been generated with any of the import options hereabove, we can proceed to the data verticalization. See also `?verticalize_plates()` for a detailed overview of function options. 

```{r}
# compute vertical plate
## from txt
(vertical_txt <- verticalize_plates(tibble_txt))

## from csv
(vertical_csv <- verticalize_plates(preview_csv))

## from skanit --> choose abs_data or map_data
(vertical_skanit_abs <- verticalize_plates(skanit_tibble_abs))
(vertical_skanit_map <- verticalize_plates(skanit_tibble_map))

# compare tibble to its verticalized format
print(tibble_txt, n = 12) ; print(vertical_txt, n = 12)
```

# Next steps

Once the data is in this practical format, we can start data transformation of all sorts, see also vignettes *xxx and xxx (under development)* of this package.

# Final note: issue of absorbance and mapping data with identical plate ids.

Usually, a physical plate that is used in the lab will come with 2 or more data tables in plate format. Indeed, the measure of absorbance of a well needs to be accompanied by some mapping information (e.g., to each well corresponds also a sample id or information on treatment). Consequently, a single plate (let's call it "NO3_plate1") will result in the import of at least 2 "plate data tables". Those 2 plate data sets characterising the same plate will need to be handled differently in downstream analysis. We see here 2 typical use cases

**1) The 2 (or more) plate data sets are separated** into separate raw files to be imported (e.g., absorbance data as a raw .TXT and mapping data recorded as a .csv). In that case, 2 distinct `tibble`s and vertical plates will be created, and it is quite easy to manipulate them separately in subsequent steps.

**2) The 2 (or more) plate data sets are imported from a single raw file**. This can be done in several ways. 

  - data is imported from a "skanit csv" where the mapping of the plates had been encoded. By default, absorbance and mapping data will be returned into 2 separate items in the list generated by `skanit_to_csv()` and can be easily manipulated separately
  - data is imported from a single csv file. In this particular case, it is recommended to name the 2 "plate data" with distinct plate ids (e.g., "NO3_plate1_abs" and "NO3_plate1_map"), so that they will simply be considered as distinct plates by the function `csv_to_skanit()`, and can easily be selected separately in subsequent steps, for example, taking advantage of the helper function `tidyselect::ends_with()` (try this: `dplyr::select(my_vertical_plate, tidyselect::ends_with("abs"))`).

