---
title: "import-tidy"
vignette: >
  %\VignetteIndexEntry{import-tidy}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

# Introduction

## Where this vignette fits

Before diving in, here's how `import-tidy` fits into the full `plate2N` pipeline — click any step to jump to its vignette.

```{mermaid}
 flowchart TD
   A(Import &amp; tidy) --> B(Handle outliers)
   B --> C1(Blank correction: std curve)
   B --> C2(Blank correction: samples)
   C1 --> D(Regression &amp; concentration)
   C2 --> D
   B -.-> M(MicroResp variant):::sideThing

   click A "articles/import-tidy.html" "import-tidy"
   click B "articles/handling-outliers.html" "handling-outliers"
   click C1 "articles/blank-correction.html" "blank-correction"
   click C2 "articles/blank-correction.html" "blank-correction"
   click D "articles/abs-to-conc.html" "abs-to-conc"
   click M "articles/microresp.html" "microresp"
   
%% color = font color
   
   classDef sideThing fill:#f8fafc,stroke:#cbd5e1,color:#64748b,stroke-width:1px; 
   linkStyle 5 stroke:#cbd5e1,stroke-width:1px;
```

The goal of `plate2N` is to facilitate data wrangling and data analysis for raw data that originates from absorbance readings with a 96-well plate reader (spectrophotometry).

```{r}
#| label: setup
library(plate2N)
```

This particular vignette adresses the first steps of the pipeline: getting from files of raw data in various forms to a uniform tidy data table that can be easily manipulated in downstream steps. The final `tidy_table` will look something like this:

```{r}
tidy_table
```

## The Whole Game

Before we dig into the specifics of each individual step, here is a glimpse of the whole pipeline. Once you've understood the principle of each indivual step as detailed in the sections below, it only takes 6 lines of code to import, join and tidy 2 datasets that could contain data from an indefinite number of 96-well plates. No worries though if this feels cryptic, we'll explain it all underneath.

```{r}
# identifying paths to raw files
map_file <- system.file("extdata", "csv_map.csv", package = "plate2N")
abs_folder <- system.file("extdata", "txt_examples/", package = "plate2N")

# importing both mapping and absorbance data
map_tibble <- csv_to_tibble(map_file)
abs_tibble <- txt_to_tibble(abs_folder)

# verticalizing and joining both data layers
joined_vertical <- join_abs_map(
  list(abs_tibble, map_tibble),
  dataset = "Nmin-", abs_map = c("abs-", "map-"))

# tidying the data
(tidy_data <- vertical_to_tidy(joined_vertical, column_def = c("abs", "map")))
```

## The Cheat Sheet

| Step | Function | Purpose |
|------------------------|------------------------|------------------------|
| Import | `csv_to_tibble()`, `txt_to_tibble()`, `skanit_to_tibble()`, `tecan_to_tibble()` | Read raw plate-format files (csv/txt/skanit/Tecan) into a common tibble structure |
| Verticalize only | `verticalize_plates()` | Reshape a single plate-format tibble into one column for the plate (use this alone if your data is already vertical/mapped and only needs one layer verticalized) |
| Verticalize + join | `join_abs_map()` | Reshape plate-format tibbles into 96 rows and merge layers (e.g., absorbance & mapping) data, to get one column per plate and per layer |
| Tidy | `vertical_to_tidy()` | Reshape into one row per well and one single column per layout, ready for downstream analysis |

## Overview of import-tidy

In this vignette, we cover three steps:

**1) Data import** from various raw file formats with **`txt_to_tibble()`**, **`csv_to_tibble()`**, **`skanit_to_tibble()`** and **`tecan_to_tibble()`** where data is brought into a shape that is common to all import files, i.e., cleaned from the file format-specific noise, but still unpractical to work with in R. We refer to the output of this first step in an object called "tibble" where data is still in a "plate" format (see `tibble_example` hereunder). In this tibble, the first column, called `row`, contains the `plate_id` in the first row, followed by the letters characterising the 8 rows of the plate (from A to H). Throughout the other 12 columns, the rest of the first row contains the plate-column numbers (from 1 to 12) in columns called X1 to X12. The rest of the table contains the original plate data, which can be numerical (e.g., absorbance data) or characters (e.g., plate mapping).

```{r}
tibble_example
```

**2) Verticalization** of the "tibble" into a verticalized table with **`verticalize_plates()`**. The `vertical_plates` present the same raw data in a verticalized format, i.e., data from each plate is displayed in a single column where the name of the column is the plate id. The first 2 columns contain the well identifier, stored in the columns `row` (of the 96-well plate, from A to H), and `column` (from 1 to 12). This vertical format is a good format to select a subset of plates (e.g., based on experiment, treatment, or anything else that the user stored into column names which are the plate ids). See example hereunder:

```{r}
vertical_plates 

# Selecting plate ids that end with "04"
vertical_plates |> 
  dplyr::select(row:column, tidyselect::ends_with("04"))
```

**3) Tidying** of the verticalized plate into the `tidy_table` displayed at the beginning of this vignette, a format that is easily usable for downstream data transformations and analyses. This step is done using the `tidyr` package

Note: If the need arises to fit this package to other plate formats than 96-well plates, please reach out.

------------------------------------------------------------------------

# 1 - Import

All 4 import functions hereunder (`csv_to_tibble()`, `skanit_to_tibble()`, `txt_to_tibble()` and `tecan_to_tibble()`) can be used to import files containing both absorbance **and** mapping data. However, in that case, it is recommended to attribute different plate ids to the 2 datasets that represent the same 96-well plate (e.g., "NO3_plate1_abs" and "NO3_plate1_map", see also description of `verticalize_plates()` in section 2 to see the final output).

## 1.1 - From a .csv file

The function `csv_to_tibble()` imports plate data from a csv file where data is structured in a specific way (see hereunder). For other csv formats, see sections below about "skanit csv" and "other formats". The required data structure is strictly the same as in the outputted tibble (see `tibble_example` hereabove).

In that sense, the function `csv_to_tibble()` constitutes the simplest import option and takes advantage of the `readr` package (part of the `tidyverse`), by using either the function `read_csv()` or `read_csv2()`, for comma-separated or semi-colon separated values, respectively. Value separators can be modulated in `csv_to_tibble()` by the parameter `delim`, which defaults with the value `","` (`read_csv()`), but accepts `";"` (`read_csv2()`).

| Argument | Purpose | Typical values to change |
|------------------------|------------------------|------------------------|
| `delim` | Value separator used in the raw csv file | `","` (default, comma-separated) or `";"` (semicolon-separated) |

Here is an example of how to call the `csv_to_tibble()` function.

```{r}
# see the function in action
# save the file path to an object called example_csv
example_csv <- system.file("extdata", "csv_example.csv", package = "plate2N")

# import the data
preview_csv <- csv_to_tibble(example_csv, delim = ",")

# previe the first 20 lines
print(preview_csv, n = 20)
```

## 1.2 - From a skanit .csv

A "skanit csv" is a csv where data is also structured in a plate format, but where plates are organized differently within the file, as generated by a software called *Skanit* that is linked to the plate *Multiskan FC* plate reader. The raw csv format can be observed in the raw data attached to this package and is displayed hereunder.

**It would make no sense to manually create a csv file in the "skanit csv" format**, and using the function `skanit_to_tibble()` is only useful for users of the *Skanit* software. See also `?skanit_to_tibble()` for more details.

```{r}

# getting file path
skanit_csv <- system.file("extdata", "skanit.csv", package = "plate2N")

# Store raw structure of a skanit-structured csv
preview_skanit_raw <- readr::read_csv(
  skanit_csv, col_names = FALSE, show_col_types = FALSE)

# Preview the first 40 lines
print(preview_skanit_raw, n = 40)
```

Here is an example of how to call the `skanit_to_tibble()` function and see its output. Notice the identical structure between `skanit$abs_tibble` and `skanit$map_tibble`.

```{r}

# Import skanit csv
skanit <- skanit_to_tibble(skanit_csv)

# store absorbance and mapping in 2 separate objects
skanit_tibble_abs <- skanit$abs_tibble
skanit_tibble_map <- skanit$map_tibble

# Preview both absorbance and mapping data
print(skanit_tibble_abs, n = 12) ; print(skanit_tibble_map, n = 12)
```

## 1.3 - From .TXT files

As for the function `skanit_to_tibble()`, `txt_to_tibble()` was designed fo import .TXT files structured in a specific way, as generated by the plate reader *Multiskan FC* when data is exported as .txt directly from the machine (without the help of the *Skanit* software. **It would make no sense to manually create a txt file in this format**.

The function `txt_to_tibble()` has its own specific logic. Indeed, whereas both `csv_to_tibble()` and `skanit_to_tibble()` take a single **file** as input, `txt_to_tibble()` takes a **folder** as input, in which an indefinite amount of single .TXT files can be found. It is however important that the folder contains no other .TXT files than the desired data, though it may contain other file types. More details on file requirements can be found under `?txt_to_tibble()`. Here is an example of folder containing 5 .TXT files.

```{r}
# path to the folder containing .TXT plate files
(txt_folder <- system.file("extdata", "txt_examples", package = "plate2N"))

#examine contents of txt_folder 
(files <- list.files(txt_folder))
```

You can find on our GitHub repository [examples of those .TXT](https://github.com/mdetoeuf/plate2N/tree/main/inst/extdata/txt_examples) files to examine.

Each .TXT file contains the absorbance data of a single plate. This means that, to create a tibble similar to `tibble_example`, `txt_to_tibble()` iterates through all .TXT files in a given folder (`txt_folder` in the example above), extracts their plate data, adds the plate_id (file name without .TXT extension) in the upper left corner of each plate data, then combines all plates into a single `tibble`.

To examine the aspect of the raw file, go and visit our GitHub repository with [examples of those .TXT](https://github.com/mdetoeuf/plate2N/tree/main/inst/extdata/txt_examples).

Here is how to run `txt_to_tibble()` from a folder path:

```{r}
# import from .TXT format
tibble_txt <- txt_to_tibble(filepath = txt_folder)

# Preview the first 20 lines
print(tibble_txt, n = 20)
```

**Note:** Should the need arise to import .TXT files structured in a different way, please reach out, and we can propose an adapted importing step into our function definition.

## 1.4 - From a Tecan .xlsx

A "Tecan .xlsx" is an Excel file where data is also structured in a plate format, but where plates are organized differently within the file, as generated by the software linked to the "Tecan" plate reader. An [example of Tecan files](https://github.com/mdetoeuf/plate2N/tree/main/inst/extdata/tecan_example) can be downloaded from our GitHub repository. The raw format is displayed hereunder.

**It would make no sense to manually create an .xlsx file in the "Tecan .xlsx" format**, and using the function `tecan_to_tibble()` is only useful for users of the Tecan plate reader. See also `?tecan_to_tibble()` for more details.

The Tecan records absorbance data with a single plate in each file, so that the `tecan_to_tibble()` function takes as first argument a `folderpath` rather than a `filepath`, in a logic similar to that of `txt_to_tibble()`. `folderpath` should give the path to the folder containing all Tecan-generated files. The names of the files (without the ".xlsx" extension) will be taken as plate_id

```{r}
# path to the folder containing .TXT plate files
(tecan_folder <- system.file("extdata", "tecan_example", package = "plate2N"))

#examine contents of tecan_folder 
(files <- list.files(tecan_folder))
```

Each .xlsx file contains the absorbance data of a single plate. This means that, to create a tibble similar to `tibble_example`, `tecan_to_tibble()` iterates through all .xlsx files in a given folder (`tecan_folder` in the example above), extracts their plate data, adds the plate_id (file name without .xlsx extension) in the upper left corner of each plate data, then combines all plates into a single `tibble`.

Here is how to run `tecan_to_tibble()` from a folder path:

```{r}
# import tecan data
tibble_tecan <- tecan_to_tibble(folderpath = tecan_folder)

# check it out
tibble_tecan
```

**Note:** Should the need arise to import Tecan files as .csv, or should your data not display properly (especially with commas instead of dots as digit separator), please reach out, and we can propose an adapted importing step into our function definition. This also applies if you need more flexibility for the import or if you need to recover other data than absorbance data from the Tecan file (hint: check out the function `readxl::read_excel()`).

## 1.5 - From other formats

For data structured in different raw formats, there are 2 options.

**1) With small data sets**, a quick option would be to simply copy-paste raw plate data into a clean csv sheet (outside of R, for example in Excel), adopting the raw structure of `tibble_example`. You can generate an empty csv template using the function *Add here the name of other function*. Or you can lookup the csv structure [here](https://github.com/mdetoeuf/plate2N/blob/main/inst/extdata/csv_example.csv).

**2) With bigger data sets**, it may be more efficient to use an import function from the `readr` package (e.g., `readr::read_csv()`, `readr::read_csv2()`, `readr::read_tsv()` or `readr::read_delim()`), possibly with a subset of columns or rows, then using tools of the `dplyr` or `tidyr` packages to bring the resulting tibble into the required format.

# 2 - Verticalized table

Finally, once the `tibble` has been generated with any of the import options hereabove, we can proceed to the data verticalization.

## 2.1 - Simple verticalization of a single tibble

Here are examples of verticalized plates. Note that the function does the exact same thing each time as the input format is strictly identical. See also `?verticalize_plates()` for a detailed overview of function options. Note that `verticalize_plates()` only verticalizes — it doesn't join. If you need to combine a verticalized layer with another dataset (e.g., mapping data that's already in a vertical shape), you'll need a join from `dplyr` (e.g., `dplyr::left_join()`) afterward.

```{r}
# compute vertical plate
## from txt
(vertical_txt <- verticalize_plates(tibble_txt))

## from csv
(vertical_csv <- verticalize_plates(preview_csv))

## from skanit --> choose abs_data or map_data
(vertical_skanit_abs <- verticalize_plates(skanit_tibble_abs))
(vertical_skanit_map <- verticalize_plates(skanit_tibble_map))

# compare tibble to its verticalized format
print(tibble_txt, n = 12) ; print(vertical_txt, n = 12)
```

## 2.2 - Joined verticalization of 2 or more related tibbles {#sec-join-abs-map}

In practice, we often import absorbance data with one function (typically `skanit_to_tibble()` or `txt_to_tibble()`) and mapping data from another (typically `csv_to_tibble()` or even simply `readr::read_csv()`). This means that, in the end, we need not only to verticalize both those datasets, but we also need to merge them in a single data set, with one column for absorbance date and another for the mapping of the wells (e.g., containing sample ids).

**`join_abs_map()`** does just that. It works directly from 2 imported `tibble`s. In the background, it relies on `verticalize_plates()` and `dplyr::left_join()` with a series of default presets[^1], adding a prefix to column names to incorporate information on dataset and data type (e.g., absorbance vs mapping).

[^1]: Should any need arise for different preset to be given optionally as argument to `join_abs_map()`, please reach out

```{r}
# Example of joined verticalized format with 2 input tibbles (2 "layers" of data)
(joined_vertical <- join_abs_map(
  tibble_list = list(skanit_tibble_abs,
  skanit_tibble_map),
  dataset = "expe1-"))
```

We recognize the same verticalized structures, but now column names have been added a double prefix, containing "expe1" which was attributed to the argument `dataset` and either "abs" or "map", added by default to the plates corresponding respectively to the first and second elements of the list given to the argument `tibble_list`.

- Defining a dataset in the column names makes sense especially if several datasets will be joined down the line. The column dataset is needed for several downstream steps, so we recommend to give it some name, even if you only work with data from a single experiment.
- The "abs/map" prefixes can be modulated by adapting the default value of the argument `abs_map = c("abs-", "map-")`

Additional info can be found under `?join_abs_map()`.

::: callout-tip
## When there are more than 2 layers for each plate

We displayed here steps to import, for each physical plate, 2 layers of data, containing absorbance and mapping data (columns `abs` and `map`, respectively). In some studies, there may be more layers of data for each physical plate. Here are some examples

- Several types of mappings containing categorical data. This is the case when factors are represented in a plate format (ex: one layer contains treatment, another contains dilutions, the nature of a drug added, or drug dosage categories, etc.)

- Several absorbance readings. For example: absorbance before and after incubation, or even at several time steps to get a time series, or readings at several wavelengths to quantify different compounds, etc.)

In such cases, the pipeline still works, but small tweaks need to be made at several steps. See also the [microresp vignette](https://mdetoeuf.github.io/plate2N/articles/microresp.html) for an example of implementation
:::

# 3 - Tidy table

The vertical format is practical for manual/human handling, with an easy-to-handle number of rows (96). For example, now is an easy moment to subset some plates based on dataset, abs/map or even some features that are embedded in plate name (such as N species in our case where plate names contained either "NO3", "NO2" or "NH4"). This can be done using the `dplyr` package and its function `dplyr::select()`, aided by some helper functions from the `tidyselect` package, such as `tidyselect::starts_with()`, `tidyselect::contains()`, etc.

However, this verticalized format is not ideal for bulk computing and transformations, where the might of the `tidyverse` sits in computations done on columns. So, a typical last step of tidying would be to bring this dataset in a format that has a single column per data type, namely one column for the absorbance data, and another for the mapping data, where each unique well would be represented by a row. Such transformations can be done with the `tidyr` and "`dplyr` packages, for example with the successive use of `tidyr::pivot_longer()`and `tidyr::pivot_wider()`.

To simplify this process, this is exactly what our function **`vertical_to_tidy()`** does.

```{r}
# tidying 
(tidy_data <- vertical_to_tidy(joined_vertical, column_def = c("abs", "map")))
```

Here is, once more, the complete pipeline from import to tidy table

```{r}
# identifying raw files
map_file <- system.file("extdata", "csv_map.csv", package = "plate2N")
abs_folder <- system.file("extdata", "txt_examples/", package = "plate2N")

# importing both map and abs data
map_tibble <- csv_to_tibble(map_file)
abs_tibble <- txt_to_tibble(abs_folder)

# joining both data
joined_vertical <- join_abs_map(
  list(abs_tibble, map_tibble),
  dataset = "Nmin-", abs_map = c("abs-", "map-"))

# tidying 
(tidy_data <- vertical_to_tidy(joined_vertical, column_def = c("abs", "map")))
```

# 4 - Variation on the import pipeline: more than 2 data layers per plate

In some studies, there may be more than 2 layers of data for each physical plate (see tip above). This case is covered separately, with a full worked example, in the import section of the [microresp vignette](https://mdetoeuf.github.io/plate2N/articles/microresp.html).

# 5 - Next steps

Once the data is in this practical and tidy format, we can start data transformation of all sorts, see also vignettes [**handling-outliers**](https://mdetoeuf.github.io/plate2N/articles/handling-outliers.html)**, [blank-correction](https://mdetoeuf.github.io/plate2N/articles/blank-correction.html) and [abs-to-conc](https://mdetoeuf.github.io/plate2N/articles/abs-to-conc.html) (in that order)** of this [package](https://mdetoeuf.github.io/plate2N/index.html). This is a good place to save the tidy table into an intermediary file, for example, using `tidy_data |> readr::write_rds("path/to/my/output/file.rds")`.

# 6 - Final note: when absorbance and mapping data with identical plate ids are imported from a single source file.

Usually, a physical plate that is used in the lab will come with 2 or more data tables in plate format. Indeed, the measure of absorbance of a well needs to be accompanied by some mapping information (e.g., to each well corresponds also a sample id or information on treatment). Consequently, a single plate (let's call it "NO3_plate1") will result in the import of at least 2 "plate data tables". Those 2 plate data sets characterizing the same "physical" plate will need to be handled differently in downstream analysis. We see here 2 typical use cases

**1) The 2 (or more) plate data sets are separated** into separate raw files to be imported (e.g., absorbance data as a raw .TXT and mapping data recorded as a .csv). In that case, 2 distinct `tibble`s will be created, and it is quite easy to manipulate them separately in subsequent steps or to use `join_abs_map()` to merge them and take profit of the automatic prefix that the function adds during the verticalization/merging process.

**2) The 2 (or more) plate data sets are imported from a the same raw file**. This can be done in several ways.

- data is imported from a "skanit csv" where the mapping of the plates had been encoded. By default, absorbance and mapping data will be returned into 2 separate items in the list generated by `skanit_to_csv()` and can be easily manipulated separately, e.g., using `join_abs_map()` as mentioned above.
- data is imported from a single csv file. In this particular case, it is recommended to name the 2 "plate data" with distinct plate ids (e.g., "abs-NO3_plate1" and "map-NO3_plate1"), so that they will simply be considered as distinct plates by the function `csv_to_skanit()`, and can easily be selected separately in subsequent steps, for example, taking advantage of the helper function `tidyselect::starts_with()` (try this on the verticalized plate: `dplyr::select(my_vertical_plate, tidyselect::starts_with("abs"))`).
