---
title: "microresp"
vignette: >
  %\VignetteIndexEntry{microresp}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
#| label: setup
library(plate2N)
library(patchwork)
library(ggridges)
```

# TO DO

- Explain the choice of absorbance thresholds used in the QC step (Section 4.1)
- Double-check the µg CO2-C conversion formula (Section 5.3) against the MicroResp manual — currently taken from an Excel document
- Check where the MBC and qCO2 formulas (Section 8) come from, and confirm their units
- Update the example dataset to include more plates (e.g. plates 11-15, for 10 total across 2 runs), so the `run_id` parameter demonstrates meaningfully rather than showing a single run
- Reevaluate which variables `MR_indices` keeps, for readability (Section 9)
- Section 10's standard-soil QC plot only shows one run currently — revisit once more run data is available
- Turn Section 11's per-variable plotting into a loop over each index (basal_resp through shannon)
- Investigate and resolve the one dropped data point warning in `rSIRs` (Section 9) — likely a negative-or-zero value going into `log()`

# 1 - Introduction {#sec-microresp}

This vignette walks through the full analysis pipeline for a MicroResp experiment, from raw plate import to substrate-level CO2 respiration values. It follows the same overall logic as the main N-dosage pipeline (`import-tidy`, `blank-correction`, `abs-to-conc`), but MicroResp experiments differ enough — more data layers per plate, no standard curve, and a different final unit (a respiration rate rather than a concentration) — that they're covered here as their own self-contained walkthrough rather than split across the main vignettes.

We start, as in `import-tidy`, with importing and tidying data recorded with more than 2 layers per plate (see the tip on multiple data layers there) — then move on to the MicroResp-specific steps: joining sample metadata, quality-checking and cleaning raw absorbance, normalizing across plates, converting to %CO2 and then to a respiration rate, and finally checking for and removing per-sample outliers.

Here, we quickly go through the import pipeline with an example of data from a [MicroResp](https://www.hutton.ac.uk/scientific-services/analytical/analytical-services/microresp/) experiment. To understand what this means: here plates are incubated for 5h. An absorbance reading is done at t = 0h (`t0`) and t = 5h (`t5`). There is also mapping data, containing the info on which substrate has been added in which well. Substrates are, in this experiment: glucose (Glu), water (H2O), oxalic acid (OA), lignin (Lgn), N-acetylglucosamine (NAG), gamma-acetylbutyric acid (gABA), alanine (Ala) and urea (Urea).

So, the 3 layers of data are referred to as `abs_t0`, `abs_t5`, and `map`.

## Overview

This vignette runs through 12 sections in a single, mostly linear pipeline. Here's the shape of it, grouped into the main phases:

```{mermaid}
flowchart TD
  A[Import, tidy & join metadata]:::contribution --> C[Clean data, normalize & convert to CO2 rate]:::caseStudy
  C --> D[Average per plate & substrate]:::loopTarget
  D --> O{{QC / outlier removal}}:::contribution
  O --> D
  D --> F1[Correct for basal respiration]:::caseStudy
  F1 --> H[QC standard soils]:::caseStudy
  F1 --> F2[Compute indices: MBC, qCO2, Shannon + Final wrangling & export]:::caseStudy

  classDef contribution fill:#dbeafe,stroke:#2563eb,color:#1e3a8a,stroke-width:1.5px,font-weight:bold;
  classDef caseStudy fill:#f8fafc,stroke:#cbd5e1,color:#64748b,stroke-width:1px;
  classDef loopTarget fill:#f8fafc,stroke:#2563eb,color:#334155,stroke-width:1.5px,stroke-dasharray:4 2,font-weight:bold;
  
```

Bold boxes use `plate2N` functions directly; the remaining boxes apply the tidyverse toolbox [@wickham2019] to MicroResp data.

- **Sections 2-3**: import raw plate files, tidy them, and join in sample metadata.
- **Section 4**: quality-check raw absorbance, remove empty/edge wells.
- **Section 5**: normalize absorbance across plates, then convert to %CO2 and a respiration rate.
- **Section 6**: identify and remove per-plate[^1], per-substrate outliers.
- **Section 7**: average remaining replicates, and correct for basal respiration.
- **Sections 8-9**: compute the summary indices (MBC, qCO2, MSIR, Shannon).
- **Section 10**: a QC check on the standard soils.
- **Sections 11-12**: final study-specific wrangling and export.

[^1]: In the MicroResp experiment, one plate was attributed to each sample (with dedicated columns for comparison with the standard soil), so that "per-plate" really means "per-sample"

# 2 - Import and tidy with 2+ layers

## 2.1 - Importing 2 layers of absorbance data

```{r}
# Get path to absorbance data, t0 and t5
MR_t0_csv <- system.file("extdata", "MR_abs_t0.csv", package = "plate2N")
MR_t5_csv <- system.file("extdata", "MR_abs_t5.csv", package = "plate2N")

# import from csv file type
MR_abs_t0 <- csv_to_tibble(MR_t0_csv)
MR_abs_t5 <- csv_to_tibble(MR_t5_csv)

# Check them out
MR_abs_t0; MR_abs_t5
```

::: callout-note
## More layers of data are possible

In theory, any number of data layers are possible - you just need to extend the logic displayed here, with additional elements to some arguments that are vectors (e.g., `abs_map = c("abs_t0", "abs_t5", "map")`) and others that are lists of all data sets (one per layer).
:::

## 2.2 - Being creative with importing mappings

In a classical MicroResp experiment, a whole 96-well plate is dedicated to a single sample, though this may vary between experimental design. In this particular case, the mapping did not relate to samples, but to substrates added. And because the template of substrate addition was strictly the same for each plate, with a complete column attributed per substrate, there is a much simpler way to import the mapping data than creating a plate layout csv and importing it. This holds particularly true for an experiment with a large number of plates (we had hundreds). So feel free to be creative in generating layers of data, as best fits your needs.

Plate mapping is strictly the same for each plate throughout the experiment, so we propose to take advantage of the function `plate2N::map_plates()` (see also the `prepare-plates` [vignette](https://mdetoeuf.github.io/plate2N/articles/prepare_plates.html)) to create this mapping data in R.

First, create a vector with the names of the substrates (in the order of the plate map)

```{r}
# vector of substrates
MR_columns <- c(
  "Std_Glu", "Std_H2O", 
  "H2O", "OA", "Glu", "Lgn", "NAG", "gABA", "Ala", "Urea")
```

Then, compute the number of plates required (here, 5 plates) by counting how many cells in the first column of `MR_abs_t0` do not correspond to one of the 26 `LETTERS` (= number of plate ids).

```{r}
# nb of plates
(nb_plates <- MR_abs_t0 |> 
  dplyr::filter(!(row %in% LETTERS)) |> 
  nrow())
```

Finally, run the mapping function with

- plate_id that pastes “P” with numbers from 01 to the nb of plates

- “sample list” = repetition of the vector of substrates x the nb of plates

- no std curves or blank (not needed in MicroResp experiments), and 2 empty columns (1 and 12, typically disregarded in those experiments due to large edge effects)

- 8 wells per “sample” (in this case: not samples but substrates)

```{r}
# compute the mapping
MR_map <- map_plates(
  plate_ids = paste0("P", sprintf("%02d", seq_len(nb_plates))), 
  samples = rep(MR_columns, nb_plates),
  n_samples_per_plate = 10,
  column_curves = c(), column_blank = c(), column_empty = c(1,12),
  n_wells_samples = 8)

# Check it out
MR_map
```

Such creative methods are particularly time-saving with large datasets. Here, we have only 5 plates \<=\> 45 rows in our tibble. But in a 100-plate dataset, this would correspond to 900 rows, quite time-consuming to encode by hand.

## 2.3 - Verticalize and join all 3 layers

Notice the use of a 3-element (rather than 2-element) list and a 3-element vector for the arguments `tibble_list` and `abs_map`, respectively. The logic stays the same as with 2 layers though.

```{r}
# verticalize and join the 3 layers
MR_joined <- join_abs_map(
  tibble_list = list( MR_abs_t0, MR_abs_t5, MR_map), 
  abs_map = c("abs_t0-", "abs_t5-", "map-"), 
  coerce_numeric = FALSE, 
  dataset = "MR-" )

# Check it out
MR_joined
```

Notice the names of the columns, that now received 3 options for the layer prefix: `abs_t0-`, `abs_t5-` and `map-`.

```{r}
# Check out column names
names(MR_joined)
```

## 2.4 - Tidy the data, 3 columns for 3 layers

Tidy the table to reach 1 row per unique well (96 x nb of physical plates), and 3 important columns: `abs_t0`, `abs_t5`, `map`. The syntax for this step is strictly the same as with 2 layers, but we now have to specify the argument `column_def`\` as its default value, containing only `c("abs-", "map-")` would no longer work.

```{r}
# From vertical to tidy format with 3 layers
MR_tidy <- vertical_to_tidy(
  MR_joined, 
  column_def = c("abs_t0", "abs_t5", "map"))

# Check it out
MR_tidy

# If you like to reorder columns to have mapping first, then absorbance
MR_tidy |> dplyr::relocate(map, .before = abs_t0)
```

::: callout-warning
## Work in progress from here on

Everything from Section 3 onward is functional, but still being refined — expect rough edges, and expect this section to keep growing (an index-computation step is still to be added at the end).
:::

# 3 - Join metadata to plate data

```{r}
# Get path to metadata
metadata_csv <- system.file("extdata", "MR_metadata.csv", package = "plate2N")

# import it
(MR_metadata <- readr::read_csv(metadata_csv, show_col_types = FALSE))
```

Notice the absence of column called `std_conc`. For the MicroResp experiment, there is no standard curve. However, there is other per-sample data that we will need, namely weight of soil added to each deep-well, and soil water content.

Now we can join it to the plate data

```{r}
(MR <- MR_tidy |> dplyr::left_join(MR_metadata, by = dplyr::join_by(plate_id)))
```

# 4 - Cleaning the data

## 4.1 - QC - suspicious wells

`plate2N::qc_raw_abs()` checks whether absorbance data is within a user-defined range or not. Check the documentation on this function for more details.

Here, we run the quality check to see if there is any really abnormal absorbance reading. The plots are saved in the working environment as objects `abs_t0_distrib` and `abs_t5_distrib`, which can be called on later.

```{r}
# Check for t0
MR |> 
  dplyr::mutate(abs = abs_t0) |> 
  qc_raw_abs(
    min_abs = 0.9, max_abs = 1.5, 
    plot_col_facet = "dataset", 
    show_plot = FALSE, 
    export_plot = "abs_t0_distrib") 

# And for t1
MR |> 
  dplyr::mutate(abs = abs_t5) |> 
  qc_raw_abs(
    min_abs = 0.2, max_abs = 1.5, 
    plot_col_facet = "dataset", 
    show_plot = FALSE, 
    export_plot = "abs_t5_distrib") 
```

Good news, all absorbance reads are within expected range. Let's plot them

```{r}
(abs_t0_distrib + ggplot2::labs(title = "t0")) / 
  (abs_t5_distrib + ggplot2::labs(title = "t5")) + 
  patchwork::plot_annotation(title = "Distribution of raw absorbance")
```

## 4.2 - Remove "empty" and NAs {#sec-remove-empty}

With MicroResp experiments, edge effects are not rare, so we exclude any data from columns 1 and 12 of the 96-well plate. Those have been encoded in the mapping as "empty" (though they were not physically).

We also change data format so that all absorbance data becomes numeric (until now stored as character)

```{r}
MR_clean <- MR |> 
  dplyr::filter_out(map == "empty" | is.na(abs_t0)) |> 
  dplyr::mutate(abs_t0 = as.numeric(abs_t0), abs_t5 = as.numeric(abs_t5))

# notice that there are now less rows (in particulare, it starts with column 1)
MR_clean
```

# 5 - Data transformation

## 5.1 - Normalisation

In preliminary works we noticed that `detection_plate_id` was the biggest driver of variability in our dataset. This is due to some lab issues and should not be a recurring problem, but it is a good illustration of the question of "at which scale should we normalize?". Let's visualize t0 absorbance readings for standard soils and glucose, supposedly the same between plates (see plot)

```{r}
unnormed_plot <- MR_clean |> 
  dplyr::filter(map == "Std_Glu") |> 
  ggplot2::ggplot(ggplot2::aes(x = detection_plate_id, y = as.numeric(abs_t5))) +
  ggplot2::theme_minimal() +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~dataset, scales = "free_x") +
  ggplot2::labs(title = "Raw data at t5",
       subtitle = "no normalisation") +
  ggplot2::ylim(0.38,0.58)

unnormed_plot
```

So, instead of a "per-plate" normalization, we propose a "per-dataset" normalisation. I.e., instead of

- $normalized_{abs} = \frac{abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{plate,t0})$

We get

- $normalized_{abs} = \frac {abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{dataset,t0})$

```{r}
MR_norm <- MR_clean |> 
  # compute dataset-wise mean of t0
  dplyr::mutate(mean_t0_dataset = mean(abs_t0)) |> 
  # compute plate-wise mean of t0
  dplyr::mutate(mean_t0_plate = mean(abs_t0), .by = plate_id) |> 
  # compute both norms: plate-wise and dataset-wise
  dplyr::mutate(
    norm_plate = mean_t0_plate * abs_t5 / abs_t0,
    norm_dataset = mean_t0_dataset * abs_t5 / abs_t0
  ) |> 
  # reorder columns to improve readibility
  dplyr::relocate(unique_well_id, abs_t0, abs_t5, norm_dataset, norm_plate, mean_t0_plate, mean_t0_dataset)

# Check it out
MR_norm
```

### Choice of normalisation - graphical

Redo the same plots to confirm the improvement coming from the normalisation, and look at plots with original vs new normalisation

```{r}
dataset_normed_plot <- MR_norm |> 
  dplyr::filter(map == "Std_Glu") |> 
  ggplot2::ggplot(ggplot2::aes(x = detection_plate_id, y = as.numeric(norm_dataset))) +
  ggplot2::theme_minimal() +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~dataset, scales = "free_x") +
  ggplot2::labs(title = "Normalised data", subtitle = "dataset-level") +
  ggplot2::ylim(0.38,0.58)

plate_normed_plot <- MR_norm |> 
  dplyr::filter(map == "Std_Glu") |> 
  ggplot2::ggplot(ggplot2::aes(x = detection_plate_id, y = as.numeric(norm_plate))) +
  ggplot2::theme_minimal() +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~dataset, scales = "free_x") +
  ggplot2::labs(title = "Normalised data", subtitle = "plate-level") +
  ggplot2::ylim(0.38,0.58)

unnormed_plot + plate_normed_plot + dataset_normed_plot
```

Each step, from raw to plate-level to dataset-level is an improvement on the data. This confirms the proposition, and we keep the dataset-level normalised data for downstream steps

## 5.2 - Convert to %CO2

Those are the coefficients as given in the MicroResp manual. However, if you are interested in this particular experiment, note that the manual recommends that you calibrate the equation on your own lab material.

%Co2 = -0.2265 + (-1.606)/(1+(-6.771)\*norm)

```{r}
MR_percent_co2 <- MR_norm |> 
  dplyr::mutate(
    percent_co2 = -0.2265 + ((-1.606) / (1 + ((-6.771) * norm_dataset))),
    .after = norm_dataset)

# Check it out
MR_percent_co2
```

## 5.3 - Convert to µg CO2-C / g dry soil / h

co2_emitted = (((percent_co2/100)x850x(44/22.4)x*(12/44)x*(273/(273+25)))/(soil_per_well_gx(soil_dw/100)))/5

```{r}
MR_co2_g_h <- MR_percent_co2 |> 
  dplyr::mutate(
    co2_g_h = (
      ((percent_co2 / 100) * 850 * (44/22.4) * (12/44) * (273 / (273+25))) / 
        dplyr::case_when(
          # for std soils
          column %in% c(2,3) ~ 
            (std_soil_well_g * (std_soil_dm_content_percent / 100)),
          column %in% c(4:11) ~ 
            (soil_well_g * (soil_dm_content_percent / 100)))
      ) / 5
  )

MR_co2_g_h
```

Check-out value distribution

```{r}
MR_co2_g_h |> 
  ggplot2::ggplot(ggplot2::aes(x = co2_g_h)) + 
  ggplot2::theme_minimal() +
  ggridges::geom_density_ridges(ggplot2::aes(y = map)) +
  ggplot2::facet_wrap(~dataset)
```

# 6 - Per plate & per substrate outlier removal

## 6.1 - Compute preliminary average (pre-outlier removal)

MicroResp plates commonly show edge effects — sometimes marginal when the kit was well sealed, but often visible even to the naked eye along one or more edges of the plate. This is the same underlying issue already addressed for columns 1 and 12 in [Remove "empty" and NAs](#sec-remove-empty), and it applies to rows as well: a preliminary run of the steps below showed that wells in row A and row H are outliers on many plates, often a majority of them. Following both this expectation and the recommendation of our colleagues at FiBL, we therefore exclude rows A and H from the subset below before computing the average, rather than treating them case-by-case as ordinary outliers. If you'd like to check the unfiltered data, simply comment out the filtering line. We then compute coefficient of variation which will serve as a threshold to identify potential outliers

```{r}
#| output: false
MR_avg_prelim <- MR_co2_g_h |> 
  dplyr::filter_out(row %in% c("A", "H")) |> 
  dplyr::group_by(dataset, plate_id, map) |> 
  dplyr::summarize(
    substrate_avg = mean(co2_g_h),
    std_dev = stats::sd(co2_g_h)
  ) |> 
  dplyr::mutate(coef_var_percent = 100 * std_dev / substrate_avg) |> 
  dplyr::arrange(desc(coef_var_percent))
```

Check out the distribution of the coefficient of variation

```{r}
#| message: false
#| warning: false
MR_avg_prelim |> ggplot2::ggplot(ggplot2::aes(x = coef_var_percent)) +
  ggplot2::theme_minimal() +
  ggplot2::geom_histogram() 
```

We have higher values than with the N-dosage pipeline, which is expected with this experiment: there are many sources of noise.

With a large data set, we set a threshold of 10-15%, i.e., data where the coefficient of variation is under the threshold will not be checked. Here we have a small data set and for the sake of showing the principle of outlier removal, we will set a somewhat stricter threshold at 8%

```{r}
suspicious_substrates <- MR_avg_prelim |> 
  dplyr::filter(abs(coef_var_percent) > 8) |> 
  dplyr::select(dataset, plate_id, map) |> unique()

suspicious_MR <- MR_co2_g_h |> 
  dplyr::filter_out(row %in% c("A", "H")) |> 
  dplyr::right_join(suspicious_substrates, by = dplyr::join_by(dataset, plate_id, map)) 
```

Now, with a threshold of 8% as max coef of variation, we are left with only 6 combinations of plate x substrate to look at (instead of 50).

This kind of automated flagging is a trade-off: it saves you from manually reviewing every single well, but it also means you're trusting the threshold to actually catch what matters. When in doubt, err on the side of a stricter threshold — one that flags more plates than strictly necessary — rather than a looser one. Reviewing a few extra, ultimately fine plates costs little; missing a real outlier because the threshold was too permissive costs much more.

## 6.2 - create QC plots

Create one list of QC plots per dataset.

Notice that we get a list with 1 plot per substrate (but not all bc not all had outliers in this reduced dataset)

```{r}
qc_plots <- plot_list_qc_microresp(
  suspicious_MR, 
  map_col = "map",
  run_id_col = "run_id")

names(qc_plots)
```

For the sake of illustration, we just plot 2 of the plots (one per dataset per substrate): each facets by run, and only displays data that is suspicious (coefficient of variation above the threshold defined above).

- H2O: we would remove B3
- Glu: no wells to remove

Notice that the number of plates differs between the two plots — only the plates flagged as suspicious for that particular substrate are shown.

```{r}
qc_plots$Std_H2O
qc_plots$Glu
```

To get a general overview, we can plot all substrates together in a single figure. Each substrate's plot is itself already a `patchwork` composition (boxplot + ridgeline) with its own title — nesting one `patchwork` inside another normally drops that inner title, so we wrap each one with `patchwork::wrap_elements()` first, which "flattens" it into a single element that keeps its title intact when combined further. In a real dataset, depending on how many plates are flagged per substrate, this combined view can get crowded quickly — treat it as a quick overview, and fall back to inspecting substrates individually (as above, possibly in a loop) when there's a lot to look through.

```{r}
(patchwork::wrap_elements(full = qc_plots$gABA) +
 patchwork::wrap_elements(full = qc_plots$Glu) +
 patchwork::wrap_elements(full = qc_plots$NAG)) /
(patchwork::wrap_elements(full = qc_plots$OA) +
 patchwork::wrap_elements(full = qc_plots$Std_H2O))

```

Save QC plots for review outside R (optional), in a loop. Notice that because plots are named in the list, they can be easily accessed by name (= substrate), and this can then serve as filenaming for export

```{r}
#| eval: false
#| include: true

for (i in seq_along(qc_plots)) {
  
  plot <- qc_plots[[i]]
  plot_name <- names(qc_plots)[i]
  
  filepath <- paste0(
      "output/figures/QC/microresp/",
      plot_name, "_", Sys.Date(), ".pdf")
  
  ggplot2::ggsave(
    filename = filepath, 
    plot = plot, width = 18, height = 10)
}
```

## 6.3 - identifying outliers + removal

Then, by visual assessment of the plots one by one, we create a table of wells to remove, and clean the dataset from obvious outliers

Starting with greenhouse outliers

```{r}
to_remove <- tibble::tribble(
  ~ plate_id, ~well_id, 
  # gABA
  "P01", "F9",   
  # NAG
  "P01", "E8",
  # Std-H2O
  "P01", "B3",      
) |> 
  dplyr::mutate(dataset = "MR")

to_remove
```

In a real dataset with hundreds of plates, this can take a while, and we recommend a second reading of those plots after "sleeping on it", and seeing if we still agree with ourselves, and didn't make any mistake

Remove those outliers

```{r}
MR_wash1 <- MR_co2_g_h |> 
  remove_wells(to_remove) 
```

It is possible to run once more through the quality check before deciding to move on, and possibly create a "wash2", "wash3", etc. version, though we never needed it so far. But it might become relevant, for example, in the case of one single really extreme value, that would "squeeze" the rest of the data for all plates of the same panel in a plot too tight to discern patterns. In such cases, a removal of the extreme outlier first, followed by re-plotting, allows the unfolding of squeezed plots

Store last version in a new variable (easier for edits...)

```{r}
MR_outliers_removed <- MR_wash1
```

# 7 - Average per plate per substrate

## 7.1 - Compute sample\*substrate average

Now that outliers are removed, we can compute the average value per plate per substrate.

Note that we create a new column where we store, for each substrate, whether the well contained "sample soil" or "standard soil" (this will be needed for downstream analysis)

```{r}
#| output: false
MR_avg <- MR_outliers_removed |> 
  dplyr::group_by(dataset, plate_id, map) |> 
  dplyr::summarize(
    co2_g_h_avg = mean(co2_g_h)) |> 
  # attribute sample soil or std soil
  dplyr::mutate(std_sample = dplyr::case_when(
      map %in% c("Std_H2O", "Std_Glu") ~ "std_soil",
      .default = "sample"
    )) 

# Check it out
MR_avg
```

Plot distribution of values

```{r}
#| message: false
distrib_uncorrected <- MR_avg |> 
  ggplot2::ggplot(ggplot2::aes(x = co2_g_h_avg)) +
  ggplot2::theme_minimal() +
  ggplot2::geom_histogram() +
  ggplot2::labs(
    title = "Distribution of per plate per substrate average\nof CO2 emission rate",
    x = "CO2 emission rate [gCO2 / g dry soil / hour]") + 
  ggplot2::facet_wrap(~dataset)

distrib_uncorrected
```

## 7.2 - Correction of averages - remove basal respiration

[@bongiorno2020] write: *Thereafter, we corrected for the average respiration rate of soil to which the deionized water was added.*

Although not fully explicit in the paper, we read this as meaning that the substrate-induced respiration computed so far is the sum of basal respiration plus the "real," substrate-specific part — so correcting for it means subtracting the basal (H2O) respiration from each substrate's value.

So, now we correct SIR's for basal respiration (subtract H2O-resp from substrate-induced resp)

First, extract basal respiration and keep only H2O rows, so that we have, separately, the basal respiration of the samples vs the standard soil

```{r}
# Extract basal resp data only
basal_resp <- MR_avg |> 
  dplyr::filter(map %in% c("H2O", "Std_H2O")) |> 
  dplyr::select(dataset, plate_id, std_sample, basal_resp = co2_g_h_avg)

# Check it out
basal_resp
```

Now we can join this info to the rest of the data and compute H2O-corrected data

```{r}
#| output: false
H2O_corrected <- MR_avg |> 
  dplyr::left_join(
    basal_resp, 
    by = dplyr::join_by(dataset, plate_id, std_sample)) |> 
  dplyr::mutate(H2O_corrected_resp = dplyr::case_when(
    map %in% c("H2O", "Std_H2O") ~ basal_resp,
    .default = co2_g_h_avg - basal_resp
  ))
```

Check it out

```{r}
H2O_corrected
```

In the following sections, we follow computations as in [@bongiorno2020], i.e.:

- Compute MSIR, MBC, metabolic quotient

- Compute relative contributions of SIR's (SIR / MSIR), refered to as rSIR's

- Derive from this relative contribution an intermediary value needed for Shanon (x\*ln(x)), refered to as sSIR's

- Compute the Shannon index

Then, we can still discuss some N-based index

# 8 - MBC, basal respiration, metabolic quotient, MSIR

Now, we compute the MBC, qCO2 and MSIR. We use H2O-corrected data throughout, and exclude both the water (basal respiration) and the standard-soil rows from the MSIR sum, since it should only reflect the real substrates.

```{r}
# Extract Glu's own respiration value per plate (needed for MBC below)
glu_resp <- H2O_corrected |> 
  dplyr::filter(map == "Glu") |> 
  dplyr::select(dataset, plate_id, glu_resp = H2O_corrected_resp)

# Keep only real substrates (exclude basal/H2O and both std soils),
# drop the now-redundant raw co2_g_h_avg, and compute MSIR as a
# grouped sum - no pivot needed
substrate_data <- H2O_corrected |> 
  dplyr::filter_out(map %in% c("H2O", "Std_H2O", "Std_Glu")) |> 
  dplyr::rename(substrate = map, SIR = H2O_corrected_resp) |> 
  dplyr::select(!co2_g_h_avg) |> 
  dplyr::group_by(dataset, plate_id) |> 
  dplyr::mutate(MSIR = sum(SIR)) |> 
  dplyr::ungroup()


```

# 9 - Shannon index

We can now join in each plate's Glu respiration (needed for MBC) and directly compute relative SIRs (rSIR) and the -x\*ln(x) intermediary value (sSIR), all while staying in the same long, per-substrate format.

```{r}
# Join in basal_resp and glu_resp, then compute MBC, qCO2, rSIR, sSIR -
# all still in long format
rSIRs <- substrate_data |> 
  dplyr::left_join(
    glu_resp, by = dplyr::join_by(dataset, plate_id)) |> 
  dplyr::mutate(
    MBC = (glu_resp * 40.04) + 0.37, # not sure where those numbers come from!
    qCO2 = basal_resp * 1000 / MBC, # same here
    rSIR = SIR / MSIR,
    sSIR = -rSIR * log(rSIR)
  )
```

! There is still one warning: we are loosing one data point (possibly still a negative or null value given to the log?)

Still, check out the output

```{r}
rSIRs
```

Now, we sum sSIRs within each plate to get the Shannon index.

```{r}
shannon <- rSIRs |> 
  dplyr::group_by(dataset, plate_id) |> 
  dplyr::summarize(
    shannon = sum(sSIR),
    # basal_resp, MBC, qCO2, MSIR are already constant within each plate
    # (computed once per plate, repeated across its substrate rows), so
    # taking the first row's value is equivalent to taking "the" value
    basal_resp = dplyr::first(basal_resp),
    MBC = dplyr::first(MBC),
    qCO2 = dplyr::first(qCO2),
    MSIR = dplyr::first(MSIR),
    .groups = "drop"
  )

# Check it out
shannon
```

Gather all data (except std soil data)

```{r}
#| output: false
MR_indices <- MR_outliers_removed |> 
  # update this selection to what makes sense in your study
  dplyr::select(dataset, plate_id, run_id:sample_name) |> 
  unique() |> 
  dplyr::left_join(shannon, by = dplyr::join_by(dataset, plate_id)) |> 
  dplyr::relocate(plate_id, sample_id, soil, basal_resp, MBC, qCO2, MSIR, shannon)
```

Check it out

```{r}
MR_indices 
```

This is per-plate aggregated data, meaning that we lost the per-substrate data, including raw `co2_g_h` and its corrected version `SIR`, as well as relative SIR (`rSIR`) and precomputed `sSIR`, which are stored in `rSIRs` (just in case). Standard-soil data is set aside too — it's rebuilt directly from `H2O_corrected` in [QC - check standard soils](#sec-qc-std-soils).

```{r}
rSIRs
```

# 10 - QC - check standard soils {#sec-qc-std-soils}

For this QC check, we use H2O-corrected data for the glucose standard.

```{r}
# Extract standard-soil data
std_soil_data <- H2O_corrected |> 
  dplyr::filter(map %in% c("Std_H2O", "Std_Glu")) |> 
  dplyr::rename(substrate = map, SIR = H2O_corrected_resp) |> 
  dplyr::select(dataset, plate_id, substrate, SIR) |>
  # joining MR_outliers_removed based on dataset and plate_id
  # add columns as needed for your study (here: run_id and detection_plate_id)
  dplyr::left_join(
    MR_outliers_removed |> 
      dplyr::select(dataset, plate_id, run_id, detection_plate_id) |> 
      unique(), 
    by = dplyr::join_by(dataset, plate_id))

# plot it
std_soil_data |> 
  ggplot2::ggplot(ggplot2::aes(x = run_id, y = SIR)) +
  ggplot2::theme_minimal() + 
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~substrate, scales = "free")
```

# 11 - Last data wrangling

For our Field experiment, some modalities are to be removed: Bloc 1 is not a real bloc, and the faba bean did not take in the Ref treatment, so we removed it from MR data. Not relevant to write it out here though (didn't make it into plates 1 to 5), but nice to mention that some more post analysis might be relevant

Plot the indices

```{r}
MR_indices |> 
  ggplot2::ggplot(ggplot2::aes(x = soil, y = basal_resp)) + 
  ggplot2::theme_minimal() +
  ggplot2::geom_boxplot()
```

# 12 - Next steps

Can be exported with `MR_indices |> write_rds("path/to/the/file.rds")` or as a csv with `MR_indices |> write_csv("path/to/the/file.csv")`

Now would come: plotting, modelling, etc. –\> "usual data analysis"
