---
title: "microresp"
vignette: >
  %\VignetteIndexEntry{microresp}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
library(patchwork)
library(ggridges)
```

# TO DO

- Add index computations at the end of the pipeline (not yet written)
- Explain the choice of absorbance thresholds used in the QC step (Section 4.1)
- Double-check the µg CO2-C conversion formula (Section 5.3) against the MicroResp manual — currently taken from an Excel document

# 1 - Introduction {#sec-microresp}

This vignette walks through the full analysis pipeline for a MicroResp experiment, from raw plate import to substrate-level CO2 respiration values. It follows the same overall logic as the main N-dosage pipeline (`import-tidy`, `blank-correction`, `abs-to-conc`), but MicroResp experiments differ enough — more data layers per plate, no standard curve, and a different final unit (a respiration rate rather than a concentration) — that they're covered here as their own self-contained walkthrough rather than split across the main vignettes.

We start, as in `import-tidy`, with importing and tidying data recorded with more than 2 layers per plate (see the tip on multiple data layers there) — then move on to the MicroResp-specific steps: joining sample metadata, quality-checking and cleaning raw absorbance, normalizing across plates, converting to %CO2 and then to a respiration rate, and finally checking for and removing per-sample outliers.

Here, we quickly go through the import pipeline with an example of data from a [MicroResp](https://www.hutton.ac.uk/scientific-services/analytical/analytical-services/microresp/) experiment. To understand what this means: here plates are incubated for 5h. An absorbance reading is done at t = 0h (`t0`) and t = 5h (`t5`). There is also mapping data, containing the info on which substrate has been added in which well. Substrates are, in this experiment: glucose (Glu), water (H2O), oxalic acid (OA), lignin (Lgn), N-acetylglucosamine (NAG), gamma-acetylbutyric acid (gABA), alanine (Ala) and urea (Urea).

So, the 3 layers of data are referred to as `abs_t0`, `abs_t5`, and `map`.

::: callout-note
## More layers of data are possible

In theory, any number of data layers are possible, you just need to extend the logic displayed here, with additional elements to some arguments that are vectors (e.g., `abs_map = c("abs_t0", "abs_t5", "map")` and others that are lists of all data sets (one per layer).
:::

# 2 - Import and tidy with 2+ layers

## 2.1 - Importing 2 layers of absorbance data

```{r}
# Get path to absorbance data, t0 and t5
MR_t0_csv <- system.file("extdata", "MR_abs_t0.csv", package = "plate2N")
MR_t5_csv <- system.file("extdata", "MR_abs_t5.csv", package = "plate2N")

# import from csv file type
MR_abs_t0 <- csv_to_tibble(MR_t0_csv)
MR_abs_t5 <- csv_to_tibble(MR_t5_csv)

# Check them out
MR_abs_t0; MR_abs_t5
```

## 2.2 - Being creative with importing mappings

In a classical MicroResp experiment, a whole 96-well plate is dedicated to a single sample, though this may vary between experimental design. In this particular case, the mapping did not relate to samples, but to substrates added. And because the template of substrate addition was strictly the same for each plate, with a complete column attributed per substrate, there is a much simpler way to import the mapping data than creating a plate layout csv and importing it. This holds particularly true for an experiment with a large number of plates (we had hundreds). So feel free to be creative in generating layers of data, as best fits your needs.

Plate mapping is strictly the same for each plate throughout the experiment, so we propose to take advantage of the function `plate2N::map_plates()` (see also the `prepare-plates` [vignette](https://mdetoeuf.github.io/plate2N/articles/prepare_plates.html)) to create this mapping data in R.

First, create a vector with the names of the substrates (in the order of the plate map)

```{r}
# vector of substrates
MR_columns <- c(
  "Std_Glu", "Std_H2O", 
  "H2O", "OA", "Glu", "Lgn", "NAG", "gABA", "Ala", "Urea")
```

Then, compute the number of plates required (here, 5 plates) by counting how many cells in the first column of `MR_abs_t0` do not correspond to one of the 26 `LETTERS` (= number of plate ids).

```{r}
# nb of plates
(nb_plates <- MR_abs_t0 |> 
  dplyr::filter(!(row %in% LETTERS)) |> 
  nrow())
```

Finally, run the mapping function with

- plate_id that pastes “P” with numbers from 01 to the nb of plates

- “sample list” = repetition of the vector of substrates x the nb of plates

- no std curves or blank (not needed in MicroResp experiments), and 2 empty columns (1 and 12, typically disregarded in those experiments due to large edge effects)

- 8 wells per “sample” (in this case: not samples but substrates)

```{r}
# compute the mapping
MR_map <- map_plates(
  plate_ids = paste0("P", sprintf("%02d", seq(01:nb_plates))), 
  samples = rep(MR_columns, nb_plates),
  n_samples_per_plate = 10,
  column_curves = c(), column_blank = c(), column_empty = c(1,12),
  n_wells_samples = 8)

# Check it out
MR_map
```

Such creative methods are particularly time-saving with large datasets. Here, we have only 5 plates \<=\> 45 rows in our tibble. But in a 100-plate dataset, this would correspond to 900 rows, quite time-consuming to encode by hand.

## 2.3 - Verticalize and join all 3 layers

Notice the use of a 3-element (rather than 2-element) list and a 3-element vector for the arguments `tibble_list` and `abs_map`, respectively. The logic stays the same as with 2 layers though.

```{r}
# verticalize and join the 3 layers
MR_joined <- join_abs_map(
  tibble_list = list( MR_abs_t0, MR_abs_t5, MR_map), 
  abs_map = c("abs_t0-", "abs_t5-", "map-"), 
  coerce_numeric = FALSE, 
  dataset = "MR-" )

# Check it out
MR_joined
```

Notice the names of the columns, that now received 3 options for the layer prefix: `abs_t0-`, `abs_t5-` and `map-`.

```{r}
# Check out column names
names(MR_joined)
```

## 2.4 - Tidy the data, 3 columns for 3 layers

Tidy the table to reach 1 row per unique well (96 x nb of physical plates), and 3 important columns: `abs_t0`, `abs_t5`, `map`. The syntax for this step is strictly the same as with 2 layers, but we now have to specify the argument `column_def`\` as its default value, containing only `c("abs-", "map-")` would no longer work.

```{r}
# From vertical to tidy format with 3 layers
MR_tidy <- vertical_to_tidy(
  MR_joined, 
  column_def = c("abs_t0", "abs_t5", "map"))

# Check it out
MR_tidy

# If you like to reorder columns to have mapping first, then absorbance
MR_tidy |> dplyr::relocate(map, .before = abs_t0)
```

::: callout-warning
## Work in progress from here on

Everything from Section 3 onward is functional, but still being refined — expect rough edges, and expect this section to keep growing (an index-computation step is still to be added at the end).
:::

# 3 - Join metadata to plate data

```{r}
# Get path to metadata
metadata_csv <- system.file("extdata", "MR_metadata.csv", package = "plate2N")

# import it
(MR_metadata <- readr::read_csv(metadata_csv, show_col_types = FALSE))
```

Notice the absence of column called `std_conc`. For the MicroResp experiment, there is no standard curve. However, there is other per-sample data that we will need, namely weight of soil added to each deep-well, and soil water content.

Now we can join it to the plate data

```{r}
(MR <- MR_tidy |> dplyr::left_join(MR_metadata, by = dplyr::join_by(plate_id)))
```

# 4 - Cleaning the data

## 4.1 - QC - suspicious wells

`plate2N::qc_raw_abs()` checks whether absorbance data is within a user-defined range or not. Check the documentation on this function for more details.

Here, we run the quality check to see if there is any really abnormal absorbance reading. The plots are saved in the working environment as objects `abs_t0_distrib` and `abs_t5_distrib`, which can be called on later.

```{r}
# Check for t0
MR |> 
  dplyr::mutate(abs = abs_t0) |> 
  qc_raw_abs(
    min_abs = 0.9, max_abs = 1.5, 
    plot_col_facet = "dataset", 
    show_plot = FALSE, 
    export_plot = "abs_t0_distrib") 

# And for t1
MR |> 
  dplyr::mutate(abs = abs_t5) |> 
  qc_raw_abs(
    min_abs = 0.2, max_abs = 1.5, 
    plot_col_facet = "dataset", 
    show_plot = FALSE, 
    export_plot = "abs_t5_distrib") 
```

Good news, all absorbance reads are within expected range. Let's plot them

```{r}
(abs_t0_distrib + ggplot2::labs(title = "t0")) / 
  (abs_t5_distrib + ggplot2::labs(title = "t5")) + 
  patchwork::plot_annotation(title = "Distribution of raw absorbance")
```

## 4.2 - Remove "empty" and NAs {#sec-remove-empty}

With MicroResp experiments, edge effects are not rare, so we exclude any data from columns 1 and 12 of the 96-well plate. Those have been encoded in the mapping as "empty" (though they were not physically).

We also change data format so that all absorbance data becomes numeric (until now stored as character)

```{r}
MR_clean <- MR |> 
  dplyr::filter_out(map == "empty" | is.na(abs_t0)) |> 
  dplyr::mutate(abs_t0 = as.numeric(abs_t0), abs_t5 = as.numeric(abs_t5))

# notice that there are now less rows (in particulare, it starts with column 1)
MR_clean
```

# 5 - Data transformation

## 5.1 - Normalisation

In preliminary works we noticed that `detection_plate_id` was the biggest driver of variability in our dataset. This is due to some lab issues and should not be a recurring problem, but it is a good illustration of the question of "at which scale should we normalize?". Let's visualize t0 absorbance readings for standard soils and glucose, supposedly the same between plates (see plot)

```{r}
unnormed_plot <- MR_clean |> 
  dplyr::filter(map == "Std_Glu") |> 
  ggplot2::ggplot(ggplot2::aes(x = detection_plate_id, y = as.numeric(abs_t5))) +
  ggplot2::theme_minimal() +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~dataset, scales = "free_x") +
  ggplot2::labs(title = "Raw data at t5",
       subtitle = "no normalisation") +
  ggplot2::ylim(0.38,0.58)

unnormed_plot
```

So, instead of a "per-plate" normalization, we propose a "per-dataset" normalisation. I.e., instead of

- $normalized_{abs} = \frac{abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{plate,t0})$

We get

- $normalized_{abs} = \frac {abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{dataset,t0})$

```{r}
MR_norm <- MR_clean |> 
  # compute dataset-wise mean of t0
  dplyr::mutate(mean_t0_dataset = mean(abs_t0)) |> 
  # compute plate-wise mean of t0
  dplyr::mutate(mean_t0_plate = mean(abs_t0), .by = plate_id) |> 
  # compute both norms: plate-wise and dataset-wise
  dplyr::mutate(
    norm_plate = mean_t0_plate * abs_t5 / abs_t0,
    norm_dataset = mean_t0_dataset * abs_t5 / abs_t0
  ) |> 
  # reorder columns to improve readibility
  dplyr::relocate(unique_well_id, abs_t0, abs_t5, norm_dataset, norm_plate, mean_t0_plate, mean_t0_dataset)

# Check it out
MR_norm
```

### Choice of normalisation - graphical

Redo the same plots to confirm the improvement coming from the normalisation, and look at plots with original vs new normalisation

```{r}
dataset_normed_plot <- MR_norm |> 
  dplyr::filter(map == "Std_Glu") |> 
  ggplot2::ggplot(ggplot2::aes(x = detection_plate_id, y = as.numeric(norm_dataset))) +
  ggplot2::theme_minimal() +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~dataset, scales = "free_x") +
  ggplot2::labs(title = "Normalised data", subtitle = "dataset-level") +
  ggplot2::ylim(0.38,0.58)

plate_normed_plot <- MR_norm |> 
  dplyr::filter(map == "Std_Glu") |> 
  ggplot2::ggplot(ggplot2::aes(x = detection_plate_id, y = as.numeric(norm_plate))) +
  ggplot2::theme_minimal() +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~dataset, scales = "free_x") +
  ggplot2::labs(title = "Normalised data", subtitle = "plate-level") +
  ggplot2::ylim(0.38,0.58)

unnormed_plot + plate_normed_plot + dataset_normed_plot
```

Each step, from raw to plate-level to dataset-level is an improvement on the data. This confirms the proposition, and we keep the dataset-level normalised data for downstream steps

## 5.2 - Convert to %CO2

Those are the coefficients as given in the MicroResp manual. However, if you are interested in this particular experiment, note that the manual recommends that you calibrate the equation on your own lab material.

%Co2 = -0.2265 + (-1.606)/(1+(-6.771)\*norm)

```{r}
MR_percent_co2 <- MR_norm |> 
  dplyr::mutate(
    percent_co2 = -0.2265 + ((-1.606) / (1 + ((-6.771) * norm_dataset))),
    .after = norm_dataset)

# Check it out
MR_percent_co2
```

## 5.3 - Convert to µg CO2-C / g dry soil / h

co2_emitted = (((percent_co2/100)x850x(44/22.4)x*(12/44)x*(273/(273+25)))/(soil_per_well_gx(soil_dw/100)))/5

```{r}
MR_co2_g_h <- MR_percent_co2 |> 
  dplyr::mutate(
    co2_g_h = (
      ((percent_co2 / 100) * 850 * (44/22.4) * (12/44) * (273 / (273+25))) / 
        dplyr::case_when(
          # for std soils
          column %in% c(2,3) ~ 
            (std_soil_well_g * (std_soil_dm_content_percent / 100)),
          column %in% c(4:11) ~ 
            (soil_well_g * (soil_dm_content_percent / 100)))
      ) / 5
  )

MR_co2_g_h
```

Check-out value distribution

```{r}
MR_co2_g_h |> 
  ggplot2::ggplot(ggplot2::aes(x = co2_g_h)) + 
  ggplot2::theme_minimal() +
  ggridges::geom_density_ridges(ggplot2::aes(y = map)) +
  ggplot2::facet_wrap(~dataset)
```

# 6 - Per plate & per substrate outlier removal

## 6.1 - Compute preliminary average (pre-outlier removal)

MicroResp plates commonly show edge effects — sometimes marginal when the kit was well sealed, but often visible even to the naked eye along one or more edges of the plate. This is the same underlying issue already addressed for columns 1 and 12 in [Remove "empty" and NAs](#sec-remove-empty), and it applies to rows as well: a preliminary run of the steps below showed that wells in row A and row H are outliers on many plates, often a majority of them. Following both this expectation and the recommendation of our colleagues at FiBL, we therefore exclude rows A and H from the subset below before computing the average, rather than treating them case-by-case as ordinary outliers. If you'd like to check the unfiltered data, simply comment out the filtering line.
We then compute coefficient of variation which will serve as a threshold to identify potential outliers

```{r}
#| output: false
MR_avg_prelim <- MR_co2_g_h |> 
  dplyr::filter_out(row %in% c("A", "H")) |> 
  dplyr::group_by(dataset, plate_id, map) |> 
  dplyr::summarize(
    substrate_avg = mean(co2_g_h),
    std_dev = stats::sd(co2_g_h)
  ) |> 
  dplyr::mutate(coef_var_percent = 100 * std_dev / substrate_avg) |> 
  dplyr::arrange(desc(coef_var_percent))
```

Check out the distribution of the coefficient of variation

```{r}
#| message: false
#| warning: false
MR_avg_prelim |> ggplot2::ggplot(ggplot2::aes(x = coef_var_percent)) +
  ggplot2::theme_minimal() +
  ggplot2::geom_histogram() 
```

We have higher values than with the N-dosage pipeline, which is expected with this experiment: there are many sources of noise.

With a large data set, we set a threshold of 10-15%, i.e., data where the coefficient of variation is under the threshold will not be checked. Here we have a small data set and for the sake of showing the principle of outlier removal, we will set a somewhat stricter threshold at 8%

```{r}
suspicious_substrates <- MR_avg_prelim |> 
  dplyr::filter(abs(coef_var_percent) > 8) |> 
  dplyr::select(dataset, plate_id, map) |> unique()

suspicious_MR <- MR_co2_g_h |> 
  dplyr::filter_out(row %in% c("A", "H")) |> 
  dplyr::right_join(suspicious_substrates, by = dplyr::join_by(dataset, plate_id, map)) 
```

Now, with a threshold of 8% as max coef of variation, we are left with only 6 combinations of plate x substrate to look at (instead of 50).

## 
