---
title: "prepare_plates"
vignette: >
  %\VignetteIndexEntry{prepare_plates}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(plate2N)
```

::: callout-caution
## Work in progress

This vignette is still under development, bugs are to be expected
:::

# 1 - Introduction

Functions introduced in this vignette are mostly helpful in a preparation step, before going to the lab and running the absorbance readings in the 96-well plate reader.

# 2 - Mapping samples to plate(s)

`map_1_plate()` takes a single plate_id and a vector containing sample names and returns a mapped plate format where the plate id is in the upper left corner.

```{r}
(samples <- sample(LETTERS, size = 15))
map_1_plate(plate_id = "test_plate", samples = samples)
```

Usually, more than one plate needs to be prepared. Let's say we need to prepare 5 plates, and we have a list of 5 x 14 samples (we want exactly 14 samples in each plate). `map_plates()` builds on `map_1_plate()` to map several plates.

```{r}
# take random fruit names as sample names
samples <- stringr::fruit[1:70]

# Have a look at the first 5
samples[1:5]

tibble <- map_plates(
  samples, 
  n_samples_per_plate = 14, 
  column_blank = 6) 
tibble |> print(n = 15)
```

The interest of this tibble is that it can easily

- be exported as a csv, then made "pretty" for printing and usage in the lab, for example with `readr::write_csv("path/to/my/mapping_data.csv)`

- be inputted back into the import pipeline, either with `csv_to_tibble()` or as this tibble in post-import steps.

Either way, but especially if the tibble is not exported as a csv, we recommend saving this tibble on the disk, for example using `readr::wrtite_rds("path/to/my/output/data.rds)`
